### 2019-02-18 README.Rmd
initial commit to euctrl-pru repo

---
output: github_document
---

This is the repo for the Reproducible ANS Performance paper.

You find here all the instructions to
fully reproduce the paper: the raw and pre-processed data, the code for the analysis and the
text of the paper. The latter two come in the form of an
[RMarkdown](https://rmarkdown.rstudio.com/) file `paper.Rmd`
that embeds prose and code chunks in the form of a 
[literate programming](https://en.wikipedia.org/wiki/Literate_programming) document.

Visit the [companion web site](https://euctrl-pru.github.io/reproducible-ans-performance-paper/)
for further details.

### 2019-02-18 analysis/data.Rmd
initial commit to euctrl-pru repo

---
title: "Data"
output:
  workflowr::wflow_html:
    toc: false
---

Here we describe the
* raw data, i.e. the Reference Trajectory data set
* pre-processed data, i.e. the extracted arrivals and the transition points.

### 2019-02-18 analysis/index.Rmd
initial commit to euctrl-pru repo

---
title: "Home"
output:
  workflowr::wflow_html:
    toc: false
---

This webpage describes the GitHub repository hosting the text, code and data for
the paper on Reproducible ANS Performance.

The aim of this repository is to make it possible to fully reproduce our analysis
and results.

### 2019-02-18 analysis/setup.Rmd
initial commit to euctrl-pru repo

---
title: "Setup"
output:
  workflowr::wflow_html:
    toc: false
---

This RMarkdown describes the setup of the R/RStudio environment needed to reproduce the paper.

## Reproduce the paper

The analysis and production of this paper make use of a number of R packages, which need to be
installed and functioning:

* `tinytex`
  ```{r, eval=FALSE}
  devtools::install_github('yihui/tinytex')
  tinytex::install_tinytex()
  ```

* `rticles`: from [`rstudio`](https://github.com/rstudio/rticles) if
  [PR #207](https://github.com/rstudio/rticles/pull/207) has been merged,
  otherwise from [`espinielli`'s branch](https://github.com/espinielli/rticles/tree/add-support-keywords)

* `tidyverse` and `sf`
  ```{r, eval=FALSE}
  install.packages(c("tidyverse", "sf"))
  ```

* install IEEE Transaction LaTeX style:
    ```{r, eval=FALSE}
    tinytex::tlmgr_install("IEEEtran")
    ```

* in order to overcome the missing small caps fonts (this needs to be done once):
  1. uncomment `\usepackage{times}` in `header.tex`
  2. knit `paper.Rmd`
  3. comment again `\usepackage{times}` in `header.tex`

You should now be in the position to knit `paper.Rmd` and obtain `paper.pdf`.

## Issues

  * References on second or more line not correctly indented. (maybe re-try BibLaTeX)

### 2019-02-18 biographies.tex
initial commit to euctrl-pru repo

\setlength{\parskip}{6pt}

\vspace*{1\baselineskip}
\hypertarget{biographies}{%
\section*{Authors' Biographies}\label{biographies}}
\addcontentsline{toc}{section}{Authors' Biographies}

\vspace*{-3\baselineskip}

\begin{IEEEbiographynophoto}{\normalsize Enrico Spinielli}\normalsize
is the Data Manager of the Performance Review Unit at EUROCONTROL, Brussels.
He holds a Master Degree in Electronic Engineering from the University of Pisa, Italy.
His intellectual interests include data analytics, data visualization and
map projections.
\end{IEEEbiographynophoto}

\vspace*{-2.5\baselineskip}

\begin{IEEEbiographynophoto}{\normalsize Rainer Koelle}\normalsize
heads the Safety and Quality of Service section
of the Performance Review Unit at EUROCONTROL, Brussels.
He holds a PhD from the University of Lancaster, 2013, in Communication Systems.
His research interests include data analytics for operational performance analysis.
\end{IEEEbiographynophoto}

\vfill

### 2019-02-18 exploration/check-coverage.Rmd
initial commit to euctrl-pru repo

---
title: "Checking Completeness"
author: "RQ & ES"
date: "4 February 2019"
output: html_document
---

```{r, message=FALSE}
library(readr)
library(dplyr)
library(lubridate)
library(purrr)
library(ggplot2)
```

## Overview

This RMarkdown document is a quick hack to check the coverage of the ref trajectory data vs APDF.

## Read in APDF Data

For this paper we prepared a data set containing the arrivals at London Heathrow and Dublin (and Zurich - tbd)

```{r}
egll2017 <- read_csv("../data/egll_arrs_fact_2017.csv")
eidw2017 <- read_csv("../data/eidw_arrs_fact_2017.csv")
```

```{r}
start_date = lubridate::dmy_hms("01-08-2017 00:00:00", tz = "UTC")
end_date   = lubridate::dmy_hms("08-08-2017 23:59:59", tz = "UTC")

trim_time_horizon <- function(arrs, .start_date = start_date, .end_date = end_date){
  arrs <- arrs %>% mutate(ARR_TIME = ifelse(is.na(AIBT), STA, AIBT)) %>%
    filter(ARR_TIME >= start_date, ARR_TIME <= end_date) %>%
    select(-ARR_TIME)
  return(arrs)
}
```

## Read In Trajectory Flight Data

Trajectory derived flight data is stored in daily samples.

```{r}
read_flight_trj <- function(apt, .path = "../data/"){
  data_path <- .path
  data_files<- dir(data_path, pattern = paste(apt, "_flt_rt", sep = ""))

  flts <- data_files %>%
    map(~ read_csv(file.path(data_path, .),
      col_types = cols(
         period_start = col_datetime(format = "")
        ,period_finish = col_datetime(format = "")
        ))
      ) %>%
    bind_rows()
  return(flts)
}
```

```{r, warning=FALSE}
egll2017_flt_trj <- read_flight_trj("egll")
```

## Plot Coverage

```{r}
extract_arrs_fm_trj <- function(ds) {
  flt_trj <- ds %>% 
    select(REG = aircraft_reg, TYP = aircraft_type, ADEP = adep, ADES = ades
           , period_start, period_finish) %>%
    mutate(ARR_TIME = if_else(!is.na(period_finish), period_finish, period_start)
          # ,ARR_TIME = as.POSIXct(ARR_TIME, origin = "1970-01-01")
           ) %>%
    select(-contains("period"))
}
```

```{r, warning=FALSE}
egll_arrs_trj <- read_flight_trj("egll") %>% 
  extract_arrs_fm_trj() %>%
  mutate(DOF = lubridate::date(ARR_TIME), SOURCE = "TRJ") %>%
  filter(ARR_TIME <= end_date)

egll_arrs_apdf <- egll2017 %>% trim_time_horizon() %>% 
  mutate(ARR_TIME = if_else(!is.na(AIBT), AIBT, STA) 
         , DOF = lubridate::date(ARR_TIME),SOURCE = "APDF")%>% 
  select(REG, ADEP, ADES, ARR_TIME, DOF, SOURCE)
```

```{r}
tmp <- bind_rows( egll_arrs_trj, egll_arrs_apdf)

ggplot(tmp, mapping = aes(x = factor(DOF), group = SOURCE, fill = SOURCE)) + geom_bar(position = "dodge")
```

```{r}
egll_arrs_trj %>% group_by(REG, DOF) %>% summarise(N = n()) %>% arrange(desc(N))
```

```{r}
tmp <- bind_rows( egll_arrs_trj %>% na.omit, egll_arrs_apdf)

ggplot(tmp, mapping = aes(x = factor(DOF), group = SOURCE, fill = SOURCE)) + geom_bar(position = "dodge")
```

# Dublin EIDW

```{r, warning=FALSE}
eidw_arrs_trj <- read_flight_trj("eidw") %>% 
  extract_arrs_fm_trj() %>%
  mutate(DOF = lubridate::date(ARR_TIME), SOURCE = "TRJ") %>%
  filter(ARR_TIME <= end_date)

eidw_arrs_apdf <-  read_csv("../data/eidw_arrs_fact_2017.csv") %>% trim_time_horizon() %>% 
  mutate(ARR_TIME = if_else(!is.na(AIBT), AIBT, STA) 
         , DOF = lubridate::date(ARR_TIME),SOURCE = "APDF")%>% 
  select(REG, ADEP, ADES, ARR_TIME, DOF, SOURCE)
```

```{r}
tmp <- bind_rows( eidw_arrs_trj %>% na.omit, eidw_arrs_apdf)

ggplot(tmp, mapping = aes(x = factor(DOF), group = SOURCE, fill = SOURCE)) + 
  geom_bar(position = "dodge")
```

## Export

The aforementioned data processing is now wrapped into utility functions and ported to asma.R.
```{r}
rm(list = ls())  # reset to empty working environment
source("../R/asma.R")
```

```{r, warning=FALSE}
# load data sets
egll_arrs_apdf <- read_arrs_apdf("egll", .path = "../data/")
egll_arrs_trj  <- read_flts_trj( "egll", .path = "../data/")

# prepare coverage data set
egll_cov_apdf <- egll_arrs_apdf %>% prep_cov_apdf()
egll_cov_trj  <- egll_arrs_trj %>% prep_cov_trj()
```

```{r}
tmp <- bind_rows(egll_cov_apdf, egll_cov_trj) %>%
         trim_time_horizon %>% na.omit

ggplot(data = tmp
       ,mapping = aes(x = factor(DOF), fill = SOURCE)
       ) +
  geom_bar(position = "dodge") + 
  theme_bw()
```
```{r}
cov_tbl <- tmp %>% group_by(SOURCE, DOF) %>%
  summarise(N_ARR = n()) %>%
  tidyr::spread(SOURCE, N_ARR) %>%
  mutate(COV = (RTRJ - APDF)/APDF)
```
```{r}
knitr::kable(cov_tbl)
```

### 2019-02-18 exploration/intersections-with-sf.Rmd
initial commit to euctrl-pru repo

---
title: "Extract Trajectories around Aerodromes"
author: "Enrico Spinielli"
date: "12/21/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(sf)
library(dplyr)
library(ggplot2)

```

## Test Case:

Data prepapration:
```{r}
t1 <- tribble(
  ~lon, ~lat, ~alt,
     0,    0,    0,
     1,    0,    1,
     2,    0,    2,
     3,    0,    3,
     4,    0,    4,
     5,    0,    5,
     6,    0,    6
)

t2 <- tribble(
  ~lon, ~lat, ~alt,
     0,    0,    0,
     1,    0,    1,
     2,    0,    2,
     3,    0,    3,
     3,    1,    4,
     3,    2,    5,
     4,    3,    6,
     5,    4,    6,
     5,    5,    6,
     4,    5,    6,
     3,    5,    6,
     2,    5,    6,
     1,    5,    6,
     0,    4,    6,
    -1,    3,    6,
    -2,    3,    6,
    -3,    3,    7,
    -4,    3,    8,
    -5,    3,    8,
    -6,    3,    8
)

trj1 <- t1 %>%
  st_as_sf(coords = c("lon", "lat", "alt"), crs = 4326, dim = "XYZ") %>%
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("LINESTRING")

trj2 <- t2 %>%
  st_as_sf(coords = c("lon", "lat", "alt"), crs = 4326, dim = "XYZ") %>%
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("LINESTRING")

apt <- st_point( x = c(0, 0, 0), dim = "XYZ") %>%
  st_sfc(crs = 4326) %>%
  st_buffer(dist = 4.5)
```

A circle around an aerodrome and a simple trajectory:
```{r simple-intersection}
ggplot() + geom_sf(data = trj1) + geom_sf(data = apt, fill = NA) + geom_sf(data = trj1 %>% st_cast("POINT"))
```

A circle around an aerodrome and a subtle trajectory:
```{r subtle-intersection}
ggplot() + geom_sf(data = trj2) + geom_sf(data = apt, fill = NA) + geom_sf(data = trj2 %>% st_cast("POINT"))

```

With `sf` we obtain two portions
```{r sf-intersections}
trj2 %>%
  st_zm() %>%
  st_intersection(apt) %>% 
  st_cast("LINESTRING")
```

### 2019-02-18 exploration/prepare_ASMA_ref.Rmd
initial commit to euctrl-pru repo

---
title: "ASMA Reference Table"
author: "RQ"
date: "11 February 2019"
output: html_document
---

This RMarkdown script documents the data preparatory action to trim the ASMA Reference Table for this study.

It broadly

* reads in the ASMA REF TABLE
* subsets for the study airports, i.e. EGLL, EIDW, and LSZH
* corrects for the sector spanning the northerly bearing of 360
* writes out the reference table for this study

## Extracting the ASMA Reference Table

The ASMA Reference Table is extracted from PRISME with the following SQL code.
Note: This is done by hand, i.e. files saved locally, as the ROracle interface could not be used.
Conceptually, the reference table will be made available to the public.

## Reading in and Subsetting the ASMA Reference Table

```{r, message=FALSE}
library(readr)
library(tidyr)
library(dplyr)
library(purrr)
```
```{r, message=FALSE}
study_apts <- c("EGLL", "EIDW", "LSZH")

asma_refs <- readr::read_csv2("../data/ASMA_REF_20190201.csv")

asma_refs <- asma_refs %>%
  filter(AIRPORT %in% study_apts & ASMA_RADIUS == 40) %>% 
  group_by(AIRPORT) %>%
  filter(VERSION == max(VERSION))

asma_refs <- asma_refs %>% group_by(AIRPORT, VERSION, ASMA_RADIUS) %>%
  tidyr::nest()
```

```{r}
# utility function checks if Bearing spans over North, i.e. 360.
# if condition holds, the range will be broken down into 2 sets
# a.) bearing to 360, and b.) 0 to end of span.
# all other values can stay the same, e.g. ASMA Sector, Unimpeded Time
# the changes rows will be merged to the data set
check_north_span <- function(ds){
  new_rows <- ds %>% group_by(AC_CLASS, RUNWAY) %>% 
    filter(FROM_BEARING == max(FROM_BEARING), !is.na(UNIMPEDED_TIME)) %>% 
    mutate(NORTH = TO_BEARING < FROM_BEARING ) %>% filter( NORTH)

  if(nrow(new_rows) >= 0){ message("correct for spanning north!")
    tmp1 <- new_rows %>%
      mutate(TO_BEARING = 360)
    tmp2 <- new_rows %>%
      mutate(FROM_BEARING = 0)

    ds <- anti_join(ds, new_rows) %>% bind_rows(tmp1) %>% bind_rows(tmp2)
  }else{
    message("ASMA sectors not spanning North!")
  }
  ds <- ds %>% select(-NORTH)
  return(ds)
}

# test for one case: tmp <- asma_refs$data[[1]] %>% check_north_span()
# apply to multiple cases
asma_refs <- asma_refs %>% mutate(REF = purrr::map(data, ~ check_north_span(.x)))

refs <- asma_refs %>% select(AIRPORT, VERSION, ASMA_RADIUS, REF) %>%
  tidyr::unnest() %>%
  select(AIRPORT, VERSION, ASMA_RADIUS, AC_CLASS, RUNWAY, ASMA_SECTOR, FROM_BEARING, TO_BEARING, UNIMPEDED_TIME) %>%
  arrange(AIRPORT, VERSION, ASMA_RADIUS, AC_CLASS, RUNWAY, ASMA_SECTOR)
```

Write out reference table for ATM Seminar paper.

```{r}
if(!file.exists("../data/ASMA_REF.csv")){
  message("Writing out trimmed refernce table for paper!")
  readr::write_csv(refs, "../data/ASMA_REF.csv")
}else{
  warning("Reference Table for paper already exists - stopping to write out!")
}
```

### 2019-02-18 exploration/prepare_ASMA_rwy.Rmd
initial commit to euctrl-pru repo

---
title: "Untitled"
author: "RQ"
date: "11 February 2019"
output: html_document
---

Quick hack to augment study with a heuristic approach to identifying landing runway.

```{r}
library(tidyverse)
source("../R/asma.R")
#source("../R/egll-data.R")  # throws: Error in library(trrrj) : there is no package called ‘trrrj’
```
```{r}
# paste and copy EGLL data
egll_apt <- tibble::tribble(
     ~latitude,  ~longitude, ~elevation,  ~icao,          ~id, ~type,        ~name,
  "N51°28.65'", "W0°27.68'",       83.0, "EGLL",       "EGLL", "ARP",   "EGLL ARP",
  "N51°28.65'", "W0°29.10'",         79, "EGLL", "RWY09L/27R", "RWY", "Runway 09L",
  "N51°28.66'", "W0°26.00'",         78, "EGLL", "RWY09L/27R", "RWY", "Runway 27R",
  "N51°27.89'", "W0°28.94'",         75, "EGLL", "RWY09R/27L", "RWY", "Runway 09R",
  "N51°27.90'", "W0°26.04'",         77, "EGLL", "RWY09R/27L", "RWY", "Runway 27L"
)
```

## damn ... in degree, minute decimal notation ==> do it by hand!

```{r, message=FALSE}
egll_pts <- read_pts_trj("egll", .path = "../data/") %>%
  filter(type %in% c("P40", "PLAND"))
```

```{r}
ggplot(data = egll_pts, mapping = aes(x = latitude, y = longitude, colour = type)) +
  geom_point()
```

```{r}
tmp <- egll_pts %>% filter(type == "PLAND", distance_arp <= 5) %>%
  mutate(RWY_BRG = cut(bearing, breaks = c(0,58, 180, 280, 360)))

# note: bearings are from point to ARP!
# 0,58 --> 09R
#(58,180] -->09L
#(180,280] --> 27R
#(280,360] --> 27L

egll_arp <- data.frame(longitude = -0.461389, latitude = 51.4775, type = "ARP", RWY_BRG = NA)

ggplot(data = tmp , mapping = aes(x = longitude, y = latitude, colour = RWY_BRG)) +
  geom_point() +
  geom_point(data = egll_arp, mapping = aes(size = 6))
```

What is wrong with the bearings for ASMA sectors?

```{r, message=FALSE}
egll_pts <- read_pts_trj("egll", .path = "../data/") %>%
  filter(type %in% c("P40", "PLAND"))
```

```{r}
my_ids <- c(egll_pts$flight_id[1], egll_pts$flight_id[17], egll_pts$flight_id[1723])

ggplot(data = egll_pts %>% filter(flight_id %in% my_ids)
       ,mapping = aes(x = longitude, y = latitude, colour = type)) +
  geom_point() + 
  geom_text(aes(label=bearing),hjust=0, vjust=0) +
  geom_text(aes(label=ifelse(bearing < 180, bearing + 180, bearing - 180)),hjust = -0.1 , vjust = 0.9, colour = "blue")
  ## ARP
  geom_point(data = egll_arp)

# bearing TO STATION == QDM !!!
```

Apply heuristic to EIDW

```{r, message=FALSE}
eidw_pts <- read_pts_trj("eidw", .path = "../data/") %>%
  filter(type %in% c("P40", "PLAND"))
```

```{r}
ggplot(data = eidw_pts, mapping = aes(x = latitude, y = longitude, colour = type)) +
  geom_point()
```

```{r}
# EIDW ARP
arp <- data.frame(longitude = -6.27, latitude = 53.421389, type = "ARP", RWY_BRG = NA)

# note: bearings are from point to ARP!
# 0,160 --> 10
#(160,210] --> 16
#(210,260] --> 34
#(260,360] --> 28

tmp <- eidw_pts %>% filter(type == "PLAND", distance_arp <= 5) %>%
  mutate(RWY_BRG = cut(bearing, breaks = c(0,160,210,260, 360)))

ggplot(data = tmp , mapping = aes(x = longitude, y = latitude
                                  , colour = RWY_BRG
                                  )   ) +
  geom_point() +
  geom_point(data = arp, mapping = aes(size = 6), colour = "white")
```

### 2019-02-18 exploration/replicate-ASMA.Rmd
initial commit to euctrl-pru repo

---
title: "Replicate ASMA"
author: "RQ"
date: "3 January 2019"
output: html_document
---

```{r setup, include=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
```

## Introduction

This R Markdown document is an initial stab at replicating the ASMA algorithm on the basis of the data extracted for this study/paper (i.e. ATM Seminar 2019).
The paper follows the concept of "teaching the computer to do the work". This way a track record of the data manipulation is achieved. In particular the first chapter is based on establishing the analytical data for ASMA. For the paper a subset of arrival flights are extracted from the FACT data tables (i.e. July 2017 for EGLL, EIDW, and LSZH).
The following sections work with this analytic data and serves as a validation of the trajectory derived data.

## Airport Operator Data Flow & 40NM data - reference data samples

For the determination of the ASMA indicator, the airport operator data flow (providing relevant flight-by-flight movement data) is augmented with a 40NM crossing 4-D position derived from the NM trajectory (i.e. profile 3).
The associated data is stored in the FACT table in PRISME.
The following extracts the relevant validation information from the FACT table.

With a view to establish a fully reproducible data sample, the respective FACT table data extraction is provided as "raw" data.

Note: The locally stored files have been retrieved in early 2018 as part of the international benchmarking. This is outside the scope of this paper.

```{r}
# utility functions to extract data for ASMA validation data set from FACT tables

# define required data items from FACT table
# Note: datetimes are not consistently stored in PRISME (some / or -). Thus, we 
# avoid throwing an error by defining time format with the any-separator, i.e. '%.'.
data_cols <- cols_only(   # begin pre-defined colums for study
       AP_C_FLTID = col_character()
      ,AP_C_REG   = col_character()
      ,AP_C_ARCTYP= col_character()
      ,AC_CLASS   = col_character()
      ,ADEP_ICAO  = col_character()
      ,ADES_ICAO  = col_character()
      ,SRC_PHASE  = col_character()
      ,SCHED_TIME_UTC = col_datetime(format = "%d%.%m%.%Y %H:%M:%S")
      ,BLOCK_TIME_UTC = col_datetime(format = "%d%.%m%.%Y %H:%M:%S")
      ,MVT_TIME_UTC  = col_datetime(format = "%d%.%m%.%Y %H:%M:%S")
      ,AP_C_STND  = col_character()
      ,AP_C_RWY   = col_character()
      ,C40_CROSS_TIME = col_datetime(format = "%d%.%m%.%Y %H:%M:%S")
      ,C40_CROSS_LAT  = col_double()
      ,C40_CROSS_LON  = col_double()
      ,C40_CROSS_FL  = col_integer()
      ,C100_CROSS_TIME = col_datetime(format = "%d%.%m%.%Y %H:%M:%S")
      ,C100_CROSS_LAT  = col_double()
      ,C100_CROSS_LON  = col_double()
      ,C100_CROSS_FL  = col_integer()
) # end - pre-defined columns

#' read locally stored FACT table
#'
#' FACT tables have been extracted from PRISME and stored locally in a
#' subfolder for all R projects, i.e. __DATA/. In this folder, there is
#' a subfolder per airport (e.g. EGLL/) with the associated FACT table
#' for the year.
#'
#' @param .apt 4-digit ICAO location indicator (chr string), e.g. "EGLL"
#' @param .yr  year (int), e.g. 2017
#' @param .path_to_fact (chr string) for folder & hierarchy
#'
#' @return filename (incldg. path)
#'
fact_file_name <- function(.apt, .yr, .path_to_fact){
  fn <- paste(.path_to_fact,"/", .apt,"/", .apt, "_", .yr, "_FACT.txt", sep = "")
  return(fn)
}

#' load extracted FACT table with predefined columns (data items)
#' 
#' Read in the specified set of data items from extracted FACT tables.
#'
#' @param .file_name (path and) filename of file to read
#' @param .data_cols cols_only() specification for readr::read_csv
#'
#' @return tibble of FACT table
#'
load_fact_data <- function(.file_name, .data_cols){
  ds <- readr::read_csv2(.file_name, col_types = .data_cols)
  return(ds)
}

#' load study data
#'
#' wrapper for defining file name and load of fact table
#'
load_data <- function(apt, yr, data_cols, path_to_fact = "../../__DATA/"){
  fn <- fact_file_name(.apt = apt, .yr = yr, .path_to_fact = path_to_fact)
  ds <- load_fact_data(.file_name = fn, .data_cols = data_cols)
  return(ds)
}

#' prepare loaded data for study
#'
#' The loaded data is trimmed for arrivals and the variables get
#' nicer names
trim_arrs <- function(ds){
  ds <- ds %>% dpyr::filter(SRC_PHASE == "ARR") %>%
    dplyr::select( CALLSIGN = AP_C_FLTID
           ,REG      = AP_C_REG
           ,ADEP     = ADEP_ICAO
           ,ADES     = ADES_ICAO
           ,ALDT     = MVT_TIME_UTC
           ,AIBT     = BLOCK_TIME_UTC
           ,STA      = SCHED_TIME_UTC
           ,TYP      = AP_C_ARCTYP
           ,WTC      = AC_CLASS
           ,C40_TIME = C40_CROSS_TIME
           ,C40_LAT  = C40_CROSS_LAT
           ,C40_LON  = C40_CROSS_LON
           ,C40_ALT  = C40_CROSS_FL
           )
  return(ds)
}

limit_time_horizon <- function(arrs, start = "01-07-2017", end= "31-08-2017"){
  start <- paste(start, "00:00:00", sep = "")
  start <- lubridate::dmy_hms(start, tz = "UTC")
  end   <- paste(end, "23:59:59", sep = "")
  end   <- lubridate::dmy_hms(end, tz = "UTC")

  arrs  <- arrs %>% 
    mutate(HORIZON = ifelse(is.na(AIBT), STA, AIBT)) %>%
    filter(HORIZON >= start, HORIZON <= end) %>%
    select(-HORIZON)

  return(arrs)
}

```

The following checks for the existance of the validation data sets.
If the files do not exist, the FACT table will be read and subsetted.
If the file exists, it will be loaded.

```{r, message=FALSE}
if(!file.exists("../data/egll_arrs_fact_2017.csv")){
  egll2017 <- load_data("EGLL", 2017, data_cols) %>% 
    trim_arrs() %>% 
    limit_time_horizon()
  write_csv(egll2017, "../data/egll_arrs_fact_2017.csv")
}else{
  egll2017 <- readr::read_csv("../data/egll_arrs_fact_2017.csv")
}

if(!file.exists("../data/eidw_arrs_fact_2017.csv")){
  eidw2017 <- load_data("EIDW", 2017, data_cols) %>% 
    trim_arrs() %>% 
    limit_time_horizon()
  write_csv(eidw2017, "../data/eidw_arrs_fact_2017.csv")
}else{
  eidw2017 <- readr::read_csv("../data/eidw_arrs_fact_2017.csv")
}

if(!file.exists("../data/lszh_arrs_fact_2017.csv")){
  lszh2017 <- load_data("LSZH", 2017, data_cols) %>% 
    trim_arrs() %>% 
    limit_time_horizon()
  write_csv(lszh2017, "../data/lszh_arrs_fact_2017.csv")
}else{
  lszh2017 <- readr::read_csv("../data/lszh_arrs_fact_2017.csv")
}
```

## Read In Trajectory Derived Data

Based on the trajectory derived data, we construct a "quick-and-dirty" ASMA data set.

```
To-do: identification of landing runway
--> we know 'landing end' of trajectory and should be able to determine distance and offset.
--> (alternatively, we can "steal" the runway from the FACT table)

To-do: enable read-out of xxxx-data.R without package
```

```{r, message=FALSE}
path_data <- "../data/" 
flt_fn    <- "egll_flt_rt_2017-08-01.csv"
pos_fn    <- "egll_pos_rt_40NM_2017-08-01.csv"
## need to load library trrrj
## source("../R/egll-data.R")
## EGLL ARP := "N51°28.65'", "W0°27.68'"
## converted with: https://www.directionsmag.com/site/latlong-converter/
egll_arp = data.frame(APT = "EGLL", PTS = "ARP", LAT = 51.4775, LON = 0.4613333)

flts <- readr::read_csv(paste(path_data, flt_fn, sep = ""))
poss <- readr::read_csv(paste(path_data, pos_fn, sep = ""))

flts <- left_join(flts, poss, by = "flight_id" ) %>% 
  select(flight_id, CALLSIGN = callsign, REG = aircraft_reg
         , TYP = aircraft_type, ADEP = adep, ADES = ades
         , TIME = timestamp, LAT = latitude, LON = longitude
         , DIST = distance  # is this the flown distance?
         , DIST_ARP = distance_arp 
         ) %>% 
  mutate(DIST_ARP_NM = DIST_ARP / 1852) %>% # convert to NM
  filter(DIST_ARP_NM >= 6)   # approx. short final = Gate ~ ALDT

asma_flts <- flts %>% group_by(flight_id, CALLSIGN) %>%
  summarise( C40_DIST = max(DIST_ARP_NM), C40_TIME = min(TIME)
            ,GATE_DIST= min(DIST_ARP_NM), GATE_TIME= max(TIME)
            ) %>% ungroup() %>%
  left_join(flts, by=c("flight_id", "CALLSIGN", "C40_TIME" = "TIME")
            ) %>% 
  mutate(BRG = geosphere::bearing(c(egll_arp$LON, egll_arp$LAT), cbind(LON, LAT))
         ,C40_BRG = (BRG + 360) %% 360 ) # modulo operation for 360-bearings

```

In order to determine the arrival sector for ASMA, we plot, the C40_BRG:

```{r}
ggplot(data = asma_flts, mapping = aes(x = C40_BRG)) + 
  geom_histogram(binwidth = 2)
```

Next: 
1. determine entry sectors
2. calculate travel time per entry sector and type
3. determine refernce time based on aforemntioned groups (still we need runway)

### 2019-02-18 paper.Rmd
initial commit to euctrl-pru repo

---
title: "Busting the Myth! - Reproducibility in ANS Performance"
author:
- Enrico Spinielli
- Rainer Koelle
date: "`r format(Sys.time(), '%a %b %d %Y @  %X')`"
output:
  rticles::ieee_article:
    base_format: bookdown::pdf_book
    fig_caption: no
    includes:
      in_header: header.tex
      keep_tex: yes
      after_body: biographies.tex
    number_sections: yes
    toc: no
    with_dblfloatfix: no
    with_subfig: no
  html_document: default
  word_document:
    toc: no
bibliography: references.bib
# biblatex: true
color-links: yes
csl: ieee-transactions-on-aerospace-and-electronic-systems.csl
affiliation:
  # author-columnar: yes
  institution-columnar: true
  institution:
  - author:
    - email: enrico.spinielli@eurocontrol.int
      name: Enrico Spinielli
    - email: rainer.koelle@eurocontrol.int
      name: Rainer Koelle
    department: Performance Review Unit
    location: Brussels, Belgium
    name: EUROCONTROL
keywords:
- reproducibility
- ANS performance
- flight trajectory
- open data
- performance indicator
link-citations: yes
author-meta: Enrico Spinielli, Rainer Koelle
papersize: a4paper
abstract: |
  Air transportation is undergoing a fundamental transformation. On the political, strategic, and operational level work is directed to meet the future growth of air traffic.
  To meet this goal, operational concepts and technological enablers are deployed with the promises of distinct performance improvements. These improvements are typically widely marketed, however, the underlying evidence is not made available. From that respect Air Navigation System Performance suffers from the same credibility crisis seen in other industries. 
  This paper implements the reproducibility paradigm by providing data, code, and results openly to enable another researcher or analyst to reproduce - and potentially - improve or build on our work. This work is part of a wider thread to establish Open Data and Open Source based data analytical capabilities for Air Navigation System Performance in Europe.
  The implementation of the paradigm is demonstrated on the basis of the performance reference trajectory by replicating a classical performance measure, and then moving into investigating the application of trajectory based information for enhancing the current state of performance monitoring for the arrival phase. This paper applies the concept and approach as an use case analysis for 2 European airports with different strategies for sequencing arrival traffic (i.e. stack holding and point merge).
  The work reported demonstrates the feasibility of the approach and reflects a stepping stone in the roll-out of the performance reference trajectory within the European context. The work also builds the basis for novel developments on the basis of the implemented reproducibility workflow as this paper is accompanied by a webpage that supports the impartial evaluation of the work by providing the data, code, and how this reports had been produced.

title-meta: Busting the Myth! - Reproducibility in ANS Performance
---

```{r global_options, include=FALSE}
library(png)
library(jpeg)
library(magick)

# library(png) library alread loaded
library(grid)
library(gridExtra)
library(dplyr)
library(readr)
library(tibble)
library(stringr)
library(ggplot2)

source(here::here("R/utils.R"))
source(here::here("R/asma.R"))   # check if functions need to go in other R file

knitr::opts_chunk$set(
  fig.env = 'figure',
  fig.pos = 'H',
  fig.align='center',
  out.width="100%",
  # eval = FALSE, # skip all code chunks unless explicitly "eval = TRUE"
  echo = FALSE, warning = FALSE, message = FALSE
)
```

<!-- Produced on `r format(Sys.time(), '%a %b %d %Y @  %X')`. -->

# Introduction

```{r}
#- teaser 
#- motivation
#- big picture
#- nature of the problem
```

Air transportation is undergoing a fundamental transformation. To meet the future growth of air traffic, political decision makers, strategic planers, practitioners and researchers aim to enhance the performance of air navigation services (ANS).
At the time being there is no consistent and regionally harmonised deployment programme. Typically, the deployment programmes are revolving around public-private partnerships and regional funding programmes resulting in a wide mix of consortia and stakeholder subsets.
The actual deployment of novel concepts or technological enablers is then driven by the programmes and to a lesser extent by operational network wide priorities following a piece meal approach.

Moreover, the success of these deployments is typically supported by promising business cases and operational evaluations - often expressed as operational performance improvements. 
While these improvements are marketed, the underlying data and associated analyses in support of the results are seldomly made available. 
Explanations range from data sensitivity, commercial interests and property rights, through to complexity of the data analyses or simulations, and the associated volume of data.

Accordingly, ANS performance improvements are subject to a "credibility crisis" similar to reported problems in other domains, e.g. medicine [@kolata_2011], psychology [@ritchie_2012], ecology [@fraser_etal_2018]. Fidler and Wilcox (2018) provides a recent review of the reproducibility discussion [@fidler_wilcox_2018]. 
Part of the crisis sensation was driven by the diffusion of reported results as other researchers/analysts were able to discover substantial flaws in the data analyses. However, a crisis also offers room for self-correction.
Throughout the recent years, this lead to revisitng the concept of reproducible research.
This concept is based on the premises that an independent researcher/analyst is enabled to reproduce the reported results.
Reproducibility covers therefore a spectrum and mix of method, data, and analytical code/scripts per se openly available for scrutiny by others.

```{r}
#- High-level objective
#- Background of previous work.
```

Independent (operational) ANS performance builds on transparent and robust performance monitoring and reporting.
Within the European context EUROCONTROL's Performance Review Unit (PRU) is committed to provide impartial analyses of the performance of the European Air Navigation System based on verified data.
With a view to enhance transparency, PRU is - inter alia - making performance-related data publicly available via its Portal, [http://ansperformance.eu](http://ansperformance.eu), since early 2017 combined with the methodological documentation for the calculation of the performance indicators used under the EUROCONTROL Performance Revew System and the European Union Performance Scheme.
There has been recognition of using common and openly available tools and data in ATM research and operational validation.
With the advent of crowd sourced data and aviation databases there is an opportunity to advance the state of the art by combining open data and government organisation collected data to establish a curated platform supporting the quest for reproducibility in ANS Performance.
One of the associated building blocks is the performance reference trajectory [@spinielli_2018].
The performance reference trajectory will be deployed as Open Data throughout 2019. The trajectory project builds on crowd sourced surveillance data fused with correlated position reports collected from European air navigation service providers.
Based on the Open Data model it will be possible to augment the data processing (e.g. data coverage, time synchronisation), establishment of derived data bases (e,g fleet data base), associated performance models (e.g. aircraft performance model), but also the continual further development and validation of operational performance measures/indicators.

```{r}
# Method by which the problem is approached.
```

This paper builds on the previous work and implements the reproducibility paradigm.
As a use-case example, a fully reproducible paper is devised assessing the performance in the arrival phase for 2 European airports. 
The underlying data, supporting data transformation and analysis, and the paper "source text" are made openly available at the paper companion web repository [https://github.com/euctrl-pru/reproducible-ans-performance-paper](https://github.com/euctrl-pru/reproducible-ans-performance-paper).
This enables the independent review and validation of the work reported.
Conceptually, an interested researcher/analyst can reproduce this paper by using the same toolset or validate the results by using the source data or interim data products if the complete reproduction of the study would be too time/resource consuming.

```{r}
#- Purpose and significance of the paper 
#- contribution of paper
```

The contributions of this paper are as follows:

* first fully reproducible ANS Performance related paper: The paper is written in RMarkdown allowing for the analysis of how text, code chunks, and visualisations are produced (including associated parameter settings) on the basis of the "source text".
* transparent data analyses and result generation through supporting web-page: All data used in this paper, intermediate data artifacts, and results including the associated transformation are publicly available for inspection, validation, or reproduction. 
* use-case analysis of ANS Performance during the arrival phase for 2 European airports: As an example the classical 'additional ASMA time' performance indicator is implemented on the basis of the provided data with showcasing trajectory derived capabilities to further study the efficiency of the arrival phase (e.g. impact of sequencing strategies [stack holdings, point merge]).

<!-- note: make sure that Section I does not significantly exceed first page -->

# Background - Reproducibility

Conceptually, reproducibility is not a new topic. 
It is as old as the scientific method and at the heart of the verification/validation of results.
The consistent and robust interpretation of results by independent researchers/analysts forms the evidence and the inherent body of knowledge in support of the applied methods or conclusions. 
This principle is also relevant for the further development on the basis of such results.

The terms replication and reproducibility are often used interchangeably and there is a debate about the necessary and sufficient conditions for these terms. 
Results are considered replicable, "if there is sufficient information available for independent researchers to make the same [consistent] findings using the same procedures with new data" ([@gandrud_2015], p. 3f). 
Replicability, thus, does not aim to produce the "same" (aka identical) results. Here the goal is to establish confirmation of the approach and findings by deriving same conclusions. ^[For example, the same data collection procedures may not necessarily result in "identical" results as the time horizon might be different, (new or different) data for another study case show different result values, but the conclusions are consistent with the replicated approach.] 
<!-- this should be a footnote - caret ^ followed by [] --> 
Accordingly, results are reproducible, if an independent researcher can recreate the findings based on the given information. In our domain (i.e. operational ANS Performance and associated data analysis) this will require the access to data and derived summaries, how associated visualisations have been generated, the accompanying explanatory text, and conclusions drawn for the analysis.

Reproducibility has also to be seen in light of today's "publish or perish" context. It appears that higher emphasis is given to the frequency of publication than the validity of the produced results. Within the medical world cases like the Duke ovarian cancer scandal [@kolata_2011] are not single incidents, but also demonstrate how errors during the data analysis - in this case during the data preparation phase using spread sheets - can discredit the work.
Despite the increasing awareness, the adaption of reproducibility or moving to other open scholarship/investigation techniques is not yet achieved. 
This can be contrasted by emerging practices like data journalism. One interesting example is buzzFeed's article on the use of surveillance planes operated by law enforcement and military for a variety of purposes. The authors were able to demonstrate based on Open Data that these purposes range from tracking drug trafficking to testing new surveillance technology [@peteraldhous_2016]. The credibility of the article comes from the fact that all artifacts are made available to reproduce the findings or study the chosen approach and data transformations made leading to the conclusions of the article.

Arguably there are limits to the access to underlying source data. However, these limitations also offer an opportunity to address reproducibility [@koelle_open_2017]. For example, the data volume size might prohibit the further dissemination of data, its collection is too resource intensive, or there exists restrictive license agreements. Experience has shown that even in such cases, a limited sample of the source data can be made available for others to verify the initial data processing steps. 
For that purpose the creation of an analytical data set as the output of the data collection, gathering, and cleaning process stage can be understood as sufficient in the sense of reproducible research. 
One key transparency aspect is the documentation of the data preparation steps. 
The article Spies in the Skies by Peter Aldhous and Charles Seife [@peteraldhous_2016] is a good example for a reproducible analysis of the US government's airborne surveillance program. Though the license restrictions did not allow for the publication of the underlying aircraft position data, the authors made available the analytical data set and documented their analysis from the data preparation through to the publication stage.

The reproducibility discussion focuses strongly on the recreation of results by other researchers/analysts.
A related problem is the re-generating of already completed (and potentially published) results with the - previously - available same input data. 
This is a fundamental issue that can be summarised as "can a future me re-do what I had done in the past"?
This problem space ranges across the complete data analysis life-cyle, including archiving the orignially used input data or the steps to generate the study data, documenting the data cleaning and transformation steps including the parameters chosen, etc.
By definition such a capability is a precondition for reproduction or replication by others. 
This implies that the analytical steps including the steps to produce summaries or estimates, visualisations, and associated conclusions are documented and archived. 
In order to enable others to recreate the own findings, such steps and documentation would need to be shareable.
This consideration has a strong bearing on the toolset chosen for this paper.

<!-- check if this can be merged with above -->
Some of the reasons why researchers/analysts do not publish data and computational procedures that allow reproduction will be hard to change [@daniel_nust_2018_2361304]. The predominant technical reasons have been listed above. However, there are also personal reasons such as the fear for floating ideas or loss of competitive advantage. Other reasons deal with more technical problems, e.g. lack of capabilities, procedures.

Another incentive for the adoption of an open and reproducible research practice can be seen in the wider recognition through increased citations, media attention, discussion with the community and collaborators, and ultimately funding opportunities.
The recent years have seen a strong use of research blogs and community platforms supporting the publication of smaller steps or intermediate results. The combination of web presentation and discussions offers a good basis for an initial dialogue.
Platforms like google scholar or researchgate.net offer additional features for outreach and communication.
This equally applies for paper achives such as arxiv.org for pre-prints.
Based on the credibility crisis discussion a number of journals are now promoting - or even requiring - the provision of analytical data in support of the paper publication. 
There is also an increasing number of requirements for national or European funded research with respect to data management and Open Data access to the research.

Reproducibility is discussed in various domains. The following examples further promote the use of open source software and/or data:

* archaeology: Marwick et al. (2017) focusses on computational reproducibility in archaeological research by providing basic principles. The case study of their implementation uses open source software (R/Rmarkdown/git approach)  [@marwick_2017]. The paper advocates the publishing (analytical) data in support of the paper. This should not be an issue, even if there are constraints on the overall study data.
* engineering / signal processing: Vandewalle (2009) discuss the motivation and approach to reproducibility in signal processing (i.e. what, why, and how). Also here the case is made to open access to the code and data for validation purposes [@vandewalle_2009].
* operational ANS Performance monitoring: Spinielli et al. (2018) highlight the development and use of Open Flight Trajectories for reproducible ANS [@spinielli_2018].

(ref:reproducible-research) Reproducible research, Fig. 2 in [@marwick_2017].

```{r reproducible-research,fig.cap='(ref:reproducible-research)',fig.pos='hbt',fig.height=4}
knitr::include_graphics("figures/reproducible-research-spectrum.png")
```

Across the literature, researchers have identified a "reproducibility spectrum" ranging from the extreme, i.e. publication only ( advertising and marketing of the done work), to almost or even fully replicable research.
The work of Marwick et al. (2017) [@marwick_2017] and Peng (2011) [@Peng1226] uses the analogy of a spectrum that can be mapped against the scoring of Vandewalle (2009) describing the level of reproducibility on a 0 to 5 scale [@marwick_2017].
In light of [@leek_2015], this paper conceptualises reproducibility as the "ability to recompute data analytic results given an observed dataset and knowledge of the data analysis pipeline" and accordingly "replicability of a study is the chance that an independent experiment targeting the same scientific question will produce a consistent result".

With Fig. \@ref(fig:reproducible-research) and the discussion in this section, this paper implements a fully reproducible workflow.
This is achieved by establishing this paper with open source software which allows to combine text, code, and visualisations.
The associated source text can also easily be read with a simple text editor/viewer. From that perspective, the text is fully inspectable.
This includes the majority of visualisations which are generated based on the data. Accordingly, the parameters and settings chosen can be validated by an independent researcher/analyst. Last but not least, the data for this paper is made available in conjunction with this text as Open Data. This also includes intermediate data artifacts which allows for the reproduction of the complete work or portions of it.
The conceptual approach and its building blocks will be presented in more detail in the next section.

# Method / Approach

## Conceptual Approach

(ref:concept-approach) Conceptual approach to reproducibility by PRU.

```{r concept-approach,fig.cap='(ref:concept-approach)',fig.pos='hbt',fig.height=4}
knitr::include_graphics("figures/approach-stages.png")
```

This paper builds on the conceptual approach presented in Fig. \@ref(fig:concept-approach) to implement replicable data analyses for performance monitoring and associated research.
On the data level this approach abstracts three levels of data:

* source data
* analytical data
* results (and performance related data)

The source data for this paper comprises the PRU Performance Reference Trajectory and operator reported data under the airport operator data flow. This study also uses some aeronautical information for the use case airports accessible through AIP or community maintained open data archives (e.g. http://openaip.net, http://ourairports.com).
These data flows are processed and stored in the EUROCONTROL data Warehouse.
The approach recognises that the full source data may or cannot be provided for a variety of reasons. 
These include volume of the source data sets, technical constraints allowing access to the source data storage, and data sharing policies or use limitations.
For the latter cases, PRU promotes the sharing of a so-called representative sample.
The idea is that with the provision of the associated processing code, other researchers/practitioners are able to reconstruct a part of the analytic data - and subsequently part of the reported results - on the basis of the provided source data set.

Analytical data refers to the data sets established based on the aforementioned processing code.
An essential defining characteristic of analytic data is that it is the data basis on which the reported results can be produced.
The abstraction between source data and analytic data allows for the sharing of such data as limitations are typically only appliable to the source data. We follow the notion of [@marwick_2017] by referring to the provision of a representative sample of 8 days accompanying this paper as source data.

As depicted in Fig. \@ref(fig:reproducible-research) the provision of the data, code, and text allows to move to the right part of the reproducibility spectrum, away from marketing and final results to credible, transparent, and challengeable analysis.
Accordingly, the results (or performance related data) level presents the final abstraction.

## Toolset

This work builds on the RStudio tools for the R language [@rcoreteam_2018] including Git (and the web-based repository managers GitHub and GitLab) as underlying version control system.
The R language  was originally developed within the statistical community supporting the task of statistical reporting by providing routines for the statistical computing and visualisation.
Being open source, the R community is actively engaging and sharing related software packages to enhance the core functionality.
Without limiting the impact of other packages, the development of knitr [@R-knitr] and RMarkdown [@xie2018], ggplot [@wickham_2016] for visualisation, and the so-called tidyverse packages and RStudio IDE [@rstudioteam_2015] provide the basis for an open source data analysis ecosystem.

A key feature for the implementation of the reproducibility paradigm is the fact that RMarkdown supports the combination of text, coding, and visualisations in a single document. 
The latter can be rendered into various output formats, e.g. classical paper report-styles (pdf, MS Word), html web-pages, and slides (both classical Powerpoint style or html-based presentations).
The fact that outputs can be rendered to html pages also supports the embedding of interactive graphics. This also facilitates the communication and outreach amongst researchers/analysts as standard web technology becomes a standard outlet format.
It follows that changes to the code (and associated visualisations) or text can be readily updated by re-rendering the adapted report. 
This allows to develop standard templates for recurring reports based on varying input data, but it also serves the iterative development of larger reports (and studies).

A minimal example of RMarkdown document, a plain-text file with the conventional
extension `.Rmd` is shown below:

\footnotesize

```{r include=FALSE}
b = '`r b[1]`'  # an ugly hack; don't look at me
```
```yaml
---
title: "R Markdown example"
author: "John Doe"
date: "2019-02-17"
output: html_document
---
```
````markdown
Some math 
$\frac{1}{N} \sum_{i=1}^N (x_i -\mu)^2$,
followed by numbered list:

1. *italics* and **bold** text
1. then bullets
    - simple, eh?

and a code chunk:

`r ''````{r}
library(ggplot2)
fit = lm(mpg ~ wt, data = mtcars)
b   = coef(fit)
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() + 
  geom_smooth(method = lm, se = TRUE)
```

The slope of the regression is `r b[1]`.
```
````

\normalsize

(ref:rmd-example) The output of a minimalist example of RMarkdown.

```{r rmd-example,echo=FALSE,fig.cap='(ref:rmd-example)',fig.pos='htb',fig.height=4}
knitr::include_graphics('figures/rmd-example.png', dpi = NA)
```

There are three components of an R Markdown document: the metadata, the text and the code.
The metadata, written between the pair of three dashes `---`, helps in the production of
output document (title, type of document [web page, PDF, MS Word], ...)
The text, which follows the metadata, is witten using Markdown, a simplified formatting syntax.
There are two types of (computer) code:

- A code chunk starts with three backticks and `r` indicates the
  language name,^[R Markdown supports 50 other languages.] and ends with
  three backticks.

- An inline R code expression starts with backtick r and ends with a backtick.

Code portions are executed when the document is rendered.

The result of rendering the R Markdown above is shown in Fig. \@ref(fig:rmd-example).

A particular strong point is that the R/Rstudio ecosystem is open source software.
This includes also its extensibility with other packages. There exists a wide variety of packages for data analytical tasks and techniques. This mechanism can also be useful for the communication of results. Researchers/analysts are able to provide their code and supporting documentation as a package for other researchers to build on.
For example, throughout the last year, PRU has developed certain utility functions for reoccurring tasks of the data analytical pipeline.
It is now planned to wrap these functions into a package saving other researchers/analysts of recoding certain steps.
For this paper, the functions and data transformations are reported as part of the RMarkdown and supporting R files.
All these files are available via the project webpage.

## Capabilities: Performance Reference Trajectory

Throughout the recent year PRU developed the data analytical capabilities to fuse correlated position reports and ADSB surveillance data.
This work has been reported in [@koelle_open_2017] and [@spinielli_2018]. The idea is to establish a pan-European air situation picture openly available to research and practitioners.
Conceptually, the fused performance reference trajectory will be a key milestone in combatting marketing myths.
For example, operational improvements can be directly assessed by interested analysts with a view to validate the observable performance improvements. This allows for the application of commonly accepted methods and models to express performance benefits and  - ultimately - challenge related publications on deployments and achieved improvements.

End of 2018 the development project was closed and PRU is now investigating options for deploying and maintaining the processing modules on a cloud platform.
The goal is to establish and operate the underlying data processing capabilities throughout 2019 and launch the regular provision of the performance reference trajectory as Open Data.
It is also planned to establish a network and complement the air situation picture by deploying additional sensors or determining non-ADSB derived positions as a community effort.
As a mid-term goal associated data bases and models for performance shall be established and jointly maintained.

<!-- not word by word - but the info is contained elsewhere 
## Application Use Cases: ASMA, Point Merge, holdings

This paper revolves around advancing the state of the art by demonstrating the implementation of the reproducibility paradigm.
Accordingly, this paper builds on the concepts behind Fig. \@ref(fig:reproducible-research) and Fig. \@ref(fig:concept-approach).
This includes providing the data, code, and results.
All of this is version controlled and archived on <<LINK TO ARCHIVE OR WEB PAGE>>.

The approach and processing pipeline described in this paper builds on several data sets.
...

As a practical application use case, this paper replicates the additional ASMA time algorithm.
... describe AMSA here ... which data sets used ... how compared ... point to next section for details of this.

In a second step, the ASMA principle is moved to a distance based measure.
... describe what will be done ... how does this link to what we want to achieve ... point to next section for more details.

As examples for further developments based on the aforementioned capabilities, this paper presents a closer look at arrival flow management strategies.

-->

# Results

## Data Preparation

(ref:data-preparation-pipeline) Data preparation pipeline.

```{r data-preparation-pipeline,fig.cap='(ref:data-preparation-pipeline)',fig.pos='htb',fig.height=4}
knitr::include_graphics("figures/data-preparation-pipeline.png")
```

The raw data sources for the use cases of this paper have been processed by a pipeline of R scripts.
For example, the additional ASMA time indicator and the anticipated further breakdown of the arrival phase can be described by a set of tranistion points:

* segment I: entry into the ASMA area to entry into the holding/vectoring segment
* segment II: holding/vectoring segment; for the use case reported holding stacks or point merge procedure
* segment III: final approach alignment and approach

Figure \@ref(fig:data-preparation-pipeline) shows how the transition points 
(40 NM, entry/exit from holding/point-merge area, landing) for the use case airports
have been extracted.
Due to the size of the daily reference trajectory files, the pipeline operates
on one date at the time.
Assuming the relevant RT flight and position files are downloaded
from GCP as described in [@performancereviewunit_2019],
the steps and the relevant scripts (available in the companion repository for this paper) used are:

1. the script `R/extract-arrivals.R` extracts arrivals to the `APT` airport and generates:

    * a file with flight details
    * a file with all the arrivals' positions
    * a file with all arrivals' positions within 40 NM the airport's reference position (ARP)

3. the script `R/extract-transition-points.R` extracts transition points in the terminal area
   and generates a file containing per each flight:

    * the first point within 40 NM, `P40`
    * the first and last point within the holding/point-merge area, `PHOLD`
    * an estimated landing point, `PLAND`

(ref:eidw-arrival-chart-east) Excerpt of Irish AIP showing point merge arc layout East of Dublin Airport.

```{r eidw-arrival-chart-east,fig.cap='(ref:eidw-arrival-chart-east)',fig.pos='hbt',fig.height=4}
knitr::include_graphics("figures/eidw-arrival-chart-east.png")
```

  This stage makes use of data extracted/extrapolated from the relevant AIP, see
  for example Fig. \@ref(fig:eidw-arrival-chart-east), such
  as ARP, runway name and threshold position/elevation, polygons for holding
  stacks or point-merge areas. Specifically the polygons have been heuristically
  defined and enlarged by 2NM in order to cater for the spread of real flown trajectories,
  see the scripts `R/egll-data.R` and `R/eidw-data.R`.

Each position mentioned above consist of at least

* flight ID
* 4D position (timestamp $T$, longitude, latitude, altitude)
* flown distance $D$
* sequence number
* distance from ARP

These allow to calculate for each flight

* the time spent before the holding/point-merge area, $T_{PHOLD|first} - T_{P40}$
* the distance flown before the holding/point-merge area,
  $D_{PHOLD | first} - D_{P40}$
* the time spent within the holding/point-merge area, $T_{PHOLD | last} - T_{PHOLD|first}$
* the distance flown within the holding/point-merge area,
  $D_{PHOLD | last} - D_{PHOLD | first}$
* the time spent out of the holding/point-merge area till landing,
  $T_{PLAND} - T_{PHOLD | last}$
* the distance flown out of the holding/point-merge area till landing,
  $D_{PLAND} - D_{PHOLD | last}$

## Data Coverage

The previous sections described the data preparatory action.
For this paper, a reproducible data set for the analysis of arrival management techniques at London Heathrow and Dublin airport have been extracted. 
The data set comprises flights for the period of 1.-8. August 2017.
It can be inferred from the traffic mix at London Heathrow (major international hub) that only a limited subset of flights is not equipped with ADSB, while a slightly higher level of non-equipped flights is expected at Dublin (higher share of intra-European flights).

```{r, warning=FALSE}
# load data sets
egll_arrs_apdf <- read_arrs_apdf("egll", .path = "./data/")
egll_arrs_trj  <- read_flts_trj( "egll", .path = "./data/")

eidw_arrs_apdf <- read_arrs_apdf("eidw", .path = "./data/")
eidw_arrs_trj  <- read_flts_trj( "eidw", .path = "./data/")

# prepare coverage data set
egll_cov_apdf <- egll_arrs_apdf %>% prep_cov_apdf()
egll_cov_trj  <- egll_arrs_trj %>% prep_cov_trj()

eidw_cov_apdf <- eidw_arrs_apdf %>% prep_cov_apdf()
eidw_cov_trj  <- eidw_arrs_trj %>% prep_cov_trj()
```

(ref:coverage-egll) Number of arrivals in data sets, i.e. 1.-8. July 2017.

```{r coverage-egll, fig.cap='(ref:coverage-egll)',fig.pos='hbt',fig.height=4}
tmp <- bind_rows(eidw_cov_apdf, eidw_cov_trj, egll_cov_apdf, egll_cov_trj) %>%
         trim_time_horizon %>% na.omit

ggplot(data = tmp
       ,mapping = aes(x = factor(DOF), fill = SOURCE)
       ) +
  geom_bar(position = "dodge") + 
  theme_bw() +
  facet_wrap(. ~ ADES) +
  theme(
    #legend.position = "bottom"
    legend.position = c(1, 0.65), legend.justification = c(1, 0)
    , axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5)
    ) +
  labs(x = "date of flight"
       ,fill = "data source") 
```

Fig. \@ref(fig:coverage-egll) shows the overlap of the data between the arrivals extracted from the airport operator data flow (APDF) and the reference trajectory (RTRJ) for London Heathrow (EGLL) and Dublin (EIDW).

```{r}
cov_tbl <- tmp %>% group_by(ADES, SOURCE, DOF) %>%
  summarise(N_ARR = n()) %>%
  tidyr::spread(SOURCE, N_ARR) %>%
  mutate(COV = 1 - (APDF - RTRJ)/APDF)

cov_perc <- cov_tbl %>% summarise(avg = mean(COV))
```

For the time horizon of this paper, the coverage of the reference trajectory derived data is for Heathrow (EGLL) `r round(cov_perc$avg[cov_perc$ADES == "EGLL"], 2) * 100`% and for Dublin `r round(cov_perc$avg[cov_perc$ADES == "EIDW"], 2) * 100`%.

## "replicate" ASMA (time-based)

The additional time in the arrival sequencing and metering area (ASMA) is a globally recognised measure to assess flight efficiency during the arrival phase. 
The ASMA metric calculates the additional time based on a reference time which is determined for a family of flights sharing similar characteristics. These are the entry sector into the arrival airspace, weight turbulence category and engine type, and the actual landing runway.
For each flight $i$ the additional ASMA time is

$${add.~ASMA~time}_i = {actual~travel~time}_i - {reference~time}_i$$

The reference time $i$ is determined for flights with similar arrival characteristics (c.f. [http://ansperformance.eu/references/definition/additional_asma_time.html](http://ansperformance.eu/references/definition/additional_asma_time.html)). These include 1.) the arrival entry sector at 40NM from the airport, 2.) the aircraft weigth turbulence category and engine type, and 3.) the actual landing runway.
The reference times are built for a longer time horizon and extracted from the PRU data base as additional source data.

(ref:ASMA-travel-egll) Distribution of actual ASMA travel times for EGLL (1.-8. July 2017).

```{r ASMA-travel-egll,fig.cap='(ref:ASMA-travel-egll)',fig.pos='hbt',fig.height=4, message=FALSE}
ac_typ_wtc_eng <- read_csv2(
  "data/ACTYPE_WTC_JetTurbo_V3.csv"
  , col_types = cols_only(
    "Aircraft type" = col_character()
    ,"WTC"          = col_character()
    ,"Turboprop"    = col_character()
    ,"ENGINE_TYPE"  = col_character()
  )
  ) %>% mutate(
    WTC2 = paste(WTC, ENGINE_TYPE, sep = "")
    ,WTC2 = ifelse(WTC2 == "HJ", "H", WTC2)
    ) %>% rename("TYP" = "Aircraft type")

# heuristic runway allocation
egll_rwy <- data.frame(RWY_BRG = c("(0,58]","(58,180]","(180,280]","(280,360]")
                       ,RWY    = c( "09R"  ,  "09L"   ,  "27R"    ,  "27L")
                     )  # c.f. analysis heuristic "colouring"

egll_asma_ref <- read_csv("data/ASMA_REF.csv") %>%
  filter(AIRPORT == "EGLL", ASMA_RADIUS == 40) %>%
  mutate(SECTOR = paste("(", FROM_BEARING,",", TO_BEARING, "]", sep = "")) %>%
  filter(!is.na(UNIMPEDED_TIME))

sects <- egll_asma_ref %>% pull(TO_BEARING) %>% unique %>% union(c(0,360)) %>% sort

asma_pts <- function(apt) {
  ds <- read_pts_trj(apt) %>%
    filter(type %in% c("P40", "PLAND")) %>% group_by(flight_id) %>% 
    ## might be better to split in P40 and PLAND and merge both sets & rename vars
    arrange(flight_id) %>% 
    mutate(TYP = unique(aircraft_type)
           , C40_TIME = min(timestamp) , C40_BRG  = ifelse(type == "P40",bearing,NA)
        #   , C40_DIST_FLOWN = ifelse(type == "P40", distance_flown, NA)
           , GATE_TIME = max(timestamp), GATE_BRG = ifelse(type == "PLAND",bearing,NA)
        #   , GATE_DIST_FLOWN = ifelse(type == "PLAND", distance_flown, NA)
           ) %>% 
    mutate(GATE_BRG = lead(GATE_BRG)) %>%
    ## bearings are not in QDR, but QDM (or true equivalent :) )
    mutate(C40_BRG = ifelse(bearing < 180, C40_BRG + 180, C40_BRG - 180)) %>%
    select(flight_id, TYP, C40_TIME, C40_BRG   #, C40_DIST_FLOWN
           ,GATE_TIME, GATE_BRG                #, GATE_DIST_FLOWN 
           ) %>%
    na.omit %>% ungroup() %>%
    mutate(TRAVEL = difftime(GATE_TIME, C40_TIME, units = "min" ))
  return(ds)
}

# add runway info
egll_pts <- asma_pts("egll") %>% mutate(RWY_BRG = cut(GATE_BRG, breaks = c(0,58, 180, 280, 360)))
egll_pts <- egll_pts %>% left_join(egll_rwy)

# add C40 info
egll_pts <- egll_pts %>% mutate(C40_SECT = cut(C40_BRG, breaks = sects))

# add aircraft wtc & engine
egll_pts <- egll_pts %>% left_join(ac_typ_wtc_eng %>% select(TYP, WTC2)) 
egll_asma_ref <- egll_asma_ref %>% select(WTC2 = AC_CLASS, RWY = RUNWAY , C40_SECT = SECTOR, REF_TIME = UNIMPEDED_TIME)

egll_pts <- egll_pts %>% left_join(egll_asma_ref)

# remove outlier
tmp <- egll_pts %>% filter(TRAVEL > 8, TRAVEL <=100)

ggplot(data = tmp, mapping = aes(x = C40_SECT, y = TRAVEL)) +
  geom_boxplot() +
  theme_bw() +
  ylim(0,100) +
  labs(x = "ASMA entry radials", y = "travel time [min]")
```
As an example, the distribution of the actual arrival transit times for arrivals into Heathrow (EGLL) are shown in Fig. \@ref(fig:ASMA-travel-egll). Heathrow operates holding stacks to tactically optimise the arrival throughput.
Dependent on the traffic situation, aircraft may experience longer holding times (i.e. long tails of boxplots in Fig. \@ref(fig:ASMA-travel-egll)).

```{r}
# heuristic runway allocation
eidw_rwy <- data.frame(RWY_BRG = c("(0,160]","(160,210]","(210,260]","(260,360]")
                       ,RWY    = c( "10"  ,  "16"   ,  "34"    ,  "28")
                     )  # c.f. analysis heuristic "colouring"

eidw_asma_ref <- read_csv("data/ASMA_REF.csv") %>%
  filter(AIRPORT == "EIDW", ASMA_RADIUS == 40) %>%
  mutate(SECTOR = paste("(", FROM_BEARING,",", TO_BEARING, "]", sep = "")) %>%
  filter(!is.na(UNIMPEDED_TIME))

sects <- eidw_asma_ref %>% pull(TO_BEARING) %>% unique %>% union(c(0,360)) %>% sort

# add runway info
eidw_pts <- asma_pts("eidw") %>% 
  mutate(RWY_BRG = cut(GATE_BRG, breaks = c(0, 160, 210, 260, 360))) %>%
  left_join(eidw_rwy)

# add C40 info
eidw_pts <- eidw_pts %>% mutate(C40_SECT = cut(C40_BRG, breaks = sects))

# add aircraft wtc & engine
eidw_pts <- eidw_pts %>% left_join(ac_typ_wtc_eng %>% select(TYP, WTC2)) 
eidw_asma_ref <- eidw_asma_ref %>% select(WTC2 = AC_CLASS, RWY = RUNWAY , C40_SECT = SECTOR, REF_TIME = UNIMPEDED_TIME)

eidw_pts <- eidw_pts %>% left_join(eidw_asma_ref) 
```

```{r}
calc_avg_asma <- function(ds){
  ds <- ds %>% mutate(ADD_TIME = as.double(TRAVEL) - REF_TIME) %>%
    summarise(  N_ARRS   = sum(!is.na(REF_TIME))
              , ADD_TIME = sum(ADD_TIME, na.rm = TRUE)
              , AVG_ADD_TIME = ADD_TIME / N_ARRS
              )
}

calc_avg_asma_20 <- function(ds){
  refs <- ds %>% group_by(TYP, RWY, WTC2) %>%
    summarise(REF_TIME_20 = quantile(TRAVEL, probs = 0.2, na.rm = TRUE))

  ds <- ds %>% left_join(refs, by = c("TYP", "RWY", "WTC2")) %>%
    mutate(ADD_TIME = TRAVEL - REF_TIME_20) %>%
    summarise( N_ARRS      = sum(!is.na(REF_TIME_20))
              ,ADD_TIME    = sum(ADD_TIME, na.rm = TRUE)
              ,AVG_ADD_TIME = ADD_TIME / N_ARRS
              )
}

egll_asma <- egll_pts %>% calc_avg_asma() %>% mutate(AIRPORT = "EGLL", MEASURE = "TRJ-REF")
eidw_asma <- eidw_pts %>% calc_avg_asma() %>% mutate(AIRPORT = "EIDW", MEASURE = "TRJ-REF")

egll_asma2<- egll_pts %>% calc_avg_asma_20() %>% mutate(AIRPORT = "EGLL", MEASURE = "TRJ-20PCT")
eidw_asma2<- eidw_pts %>% calc_avg_asma_20() %>% mutate(AIRPORT = "EIDW", MEASURE = "TRJ-20PCT")

```

Fig. \@ref(fig:ASMA-compare) compares the additional ASMA time extracted from the PRU performance dashboard, the trajectory based calculation with the dashboard derived reference time, and a 20th percentile approach purely based on the trajectory data.

<!-- TODO: download and extract ASMA from ansperformance.eu, possibly trajectory 20th percentile -->
<!--
```
if(!file.exists("data/ASMA_PERF.xlsm")){
  message("Download ASMA from ansperformance.eu")
  download.file(url = "http://ansperformance.eu/data/set/asma/ASMA_Additional_Time.xlsm"
                ,destfile = "data/ASMA_PERF.xlsm")
}
    asma_perfeu <- readxl::read_excel("data/ASMA_PERF.xlsm", sheet = "DATA") %>%
      filter(YEAR == 2017, APT_ICAO %in% c("EGLL", "EIDW", "LSZH"), MONTH_NUM == 7)

    asma_perfeu <- select(N_ARRS = FLT_ASMA_UNIMP_2
                          ,ADD_TIME = TIME_ASMA_ADD_2
                          ,AVG_ADD_TIME = ADD_TIME / N_ARRS
                          ,AIRPORT = APT_ICAO
                          ,MEASURE = "APDFNM")
```
WORKED ONCE SINCE THEN ERROR: Error: Evaluation error: error -103 with zipfile in unzGetCurrentFileInfo
==> downloaded xlsm and saved out data tab to ASMA.PERF.csv
-->

(ref:ASMA-compare) Comparison of determined ASMA values, i.e. 1.-8. July 2017.

```{r ASMA-compare, fig.cap='(ref:ASMA-compare)',fig.pos='hbt',fig.height=4}
asma_perfeu <- readr::read_csv2("data/ASMA_PERF.csv") %>%
  dplyr::filter(YEAR == 2017, APT_ICAO %in% c("EGLL", "EIDW"), MONTH_NUM == 7) %>%
  dplyr::rename(N_ARRS = FLT_ASMA_UNIMP_2, ADD_TIME = TIME_ASMA_ADD_2, AIRPORT = APT_ICAO) %>%
  dplyr::mutate(AVG_ADD_TIME = ADD_TIME / N_ARRS ,MEASURE = "APDFNM")

asma_perfeu <- asma_perfeu %>% 
  dplyr::select(N_ARRS, ADD_TIME, AVG_ADD_TIME, AIRPORT, MEASURE)

tmp <- bind_rows( egll_asma, egll_asma2
                 ,eidw_asma, eidw_asma2
                 ,asma_perfeu)

ggplot(data = tmp, mapping = aes(x = AIRPORT, y = AVG_ADD_TIME
                                 , fill = MEASURE)
                                 )  +
  geom_col(width = 0.7, position = position_dodge(width = 0.8)) +
  theme_bw() +
  labs(x = "airport", y = "average additional ASMA time [min./arrival]") +
  scale_fill_discrete(name = "variant", breaks = c("APDFNM", "TRJ-REF", "TRJ-20PCT"))
```

Fig. \@ref(fig:ASMA-compare) shows the variation of the timebased additional ASMA times determined for our sample.
APDFNM refers to the variant used for the EUROCONTROL Performance Review System. The reference time is build on the basis of a one-year sammple. Applying this reference time to the trajectory based travel times (i.e. TRJ-REF), there is a strong increase of the additional ASMA time for Heathrow (EGLL), while in the case of Dublin (EIDW) reasonable decrease takes place. Calculating the additional ASMA time exclusively with the trajectory sample and applying a 20th percentile approach (TRJ-20PCT), shows a significant drop in EGLL and a moderate increase at Dublin. The high variation for Heathrow shows that the additional ASMA time is strongly dependent on the time horizon.
The pure trajectory bases measure - by definition - includes the effect of the operational concept (stack holdings) and distorts the measure. In the case of Dublin it can be argued that the TRJ-20PCT measure reflects the time based ASMA measure suffiently.

## Distance-based ASMA

conceptually, measuring flight inefficiency in terms of flight time accounts for the airspace user perspective.
The previous section has also shown the high dependency of the time based measure on the sample size or reference time attribution.
Flight time is linked to engine time and accordingly, assuming maneuvering for operational separation and sequencing reasons, fuel burn.
From a service provision perspective, efficiency relates to the capability to provide the least distance flown within the ASMA area.
In that respect, the data analytical capability to determine an associated additional distance flown metric addresses an open gap.

(ref:ASMA-time-dist) Relationship travel time and distance flown, i.e. 1.-8. July 2017.

```{r ASMA-time-dist, fig.cap='(ref:ASMA-time-dist)', message = FALSE}
prep_asma_dist <- function(apt) {
   ds <- read_pts_trj(apt) %>%
    group_by(flight_id) %>% arrange(flight_id, sequence_number) %>% 
    summarise(N = n(), DIST_FLOWN = (max(distance_flown) - min(distance_flown))
              , TRAVEL = difftime(max(timestamp), min(timestamp), units = "min")
              , TYP = unique(aircraft_type) 
              ) 
   return(ds)
}

egll_dist <- prep_asma_dist("egll")
eidw_dist <- prep_asma_dist("eidw")

ggplot(data = egll_dist %>% filter(TRAVEL > 6), mapping = aes(x = DIST_FLOWN, y = TRAVEL)) +
  geom_point() + 
  theme_bw() +
  ylim(0,75) +
  labs(x = "distance flown from entry 40NM to touchdown [NM]", y = "time flown from 40NM entry to touchdown")
```

Fig. \@ref(fig:ASMA-time-dist) shows the relationship between the actual distance flown and the actual time flown.
The variation of travel times is influenced by varying airspeeds or weather phenomena (e.g. wind).
This paper applies a simple 15th percentile approach to the distance flown per arrival sector, aircraft type/engine, and landing runway.

```{r tbl-asma-dist}
asma_dist <- function(ds_pts, ds_dist){
  ds <- ds_pts %>% select(flight_id, RWY, C40_SECT, WTC2)
  ds <- ds %>% left_join(ds_dist)

  refs <- ds %>% group_by(RWY, C40_SECT, WTC2) %>% 
    summarise(REF_DIST = quantile(DIST_FLOWN, probs = 0.15)
              ,N_SAMPLE = n())

  ds <- ds %>% left_join(refs, by = c("RWY","C40_SECT","WTC2")) 
  ds <- ds %>% mutate(ADD_DIST = DIST_FLOWN - REF_DIST) %>%
    summarise(TOT_ADD_DIST = sum(ADD_DIST, na.rm = TRUE)
              ,N_ARRS  = sum(!is.na(ADD_DIST))
              ,AVG_ADD_DIST = TOT_ADD_DIST / N_ARRS)
  return(ds)
}

egll_asma_dist <- asma_dist(egll_pts, egll_dist) %>% mutate(AIRPORT = "EGLL")
eidw_asma_dist <- asma_dist(eidw_pts, eidw_dist) %>% mutate(AIRPORT = "EIDW")

dist_asma <- bind_rows(egll_asma_dist, eidw_asma_dist) %>%
  select(AIRPORT, TOT_ADD_DIST, N_ARRS, AVG_ADD_DIST)

knitr::kable(dist_asma, digits = 2
             ,caption = "Average additional distance flown (15th pct)")
```

Tab. \@ref(tab:tbl-asma-dist) lists the determined average additional distance at the study airports.

## Arrival Management Practices

For an initial drill down, this paper uses arrival practices at London Heathrow and Dublin.
The approach to managing the arrival flow differs significantly between both airports.
Conceptually, London Heathrow employs the "pressure on the runway" concept with a focus on achieving and balancing a continuously high level of runway throughput utilisation. 
This is achieved by controlling the demand through 4 arrival stacks that allow for the tactical sequencing on the last portion of the approach.
Dublin is one of the airports that pioneered the implementation of "point merge".
Point merge aims at replacing circular holdings by longitudal holdings to establish the sequence based on predefined legs. 
<!-- ADD A FEW MORE WORDS ON THE CONCEPTS IF NEEDED -->

From a performance perspective it is therefore interesting to identify and measure the impact of the operational concept on the overall flight efficiency during the arrival phase of flight.

```{r, message=FALSE}
asma_dist2 <- function(ds_pts, ds_dist){
  ds <- ds_pts %>% select(flight_id, RWY, C40_SECT, WTC2)
  ds <- ds %>% left_join(ds_dist)

  refs <- ds %>% group_by(RWY, C40_SECT, WTC2) %>% 
    summarise(REF_DIST = quantile(DIST_FLOWN, probs = 0.15)
              ,N_SAMPLE = n())

  ds <- ds %>% left_join(refs, by = c("RWY","C40_SECT","WTC2")) 
  ds <- ds %>% mutate(ADD_DIST = DIST_FLOWN - REF_DIST)
  return(ds)
}

egll_long_dist <- asma_dist2(egll_pts, egll_dist) %>% 
  filter(TRAVEL > 5) %>%    # remove outlier
  arrange(desc(ADD_DIST))
egll_day1 <- readr::read_csv("data/egll_pos_rt_40NM_2017-08-01.csv")

egll_long_dist <- egll_long_dist %>% mutate(IS_DAY1 = ifelse(flight_id %in% egll_day1$flight_id, 1, 0)) %>%
  filter(IS_DAY1 == 1) %>% head(3)
egll_day1 <- egll_day1 %>% filter(flight_id %in% egll_long_dist$flight_id)

source(here::here("R/plot-egll-tma.R"))
```

(ref:egll-trajectory-example) EGLL - long trajectory flown on 1. July 2017.

```{r egll-trajectory-example, fig.cap='(ref:egll-trajectory-example)'}
plot_egll_tma() + geom_point(
  data = egll_day1 %>% select(flight_id, latitude, longitude)
  , mapping = aes(x = longitude, y = latitude, colour = flight_id)
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

(ref:eidw-trajectory-example) EIDW - long trajectory flown on 1. July 2017.

```{r eidw-trajectory-example, fig.cap='(ref:eidw-trajectory-example)'}
eidw_long_dist <- asma_dist2(eidw_pts, eidw_dist) %>% 
  filter(TRAVEL > 5) %>%    # remove outlier
  arrange(desc(ADD_DIST))
eidw_day1 <- readr::read_csv("data/eidw_pos_rt_40NM_2017-08-01.csv")

eidw_long_dist <- eidw_long_dist %>% mutate(IS_DAY1 = ifelse(flight_id %in% eidw_day1$flight_id, 1, 0)) %>%
  filter(IS_DAY1 == 1) %>% filter(ADD_DIST < 50) %>% 
  filter(flight_id %in% c(
    "80788782-6fa7-42f5-b636-b5eee6822aab"
    ,"bbef0049-6748-49dc-ada3-8d058a7a03a6"
  #  ,"8df967dd-6880-4fda-86c3-a573a32abb19"
    ,"2db1b8a1-28ce-4dbf-9115-7404940bc832"
  ,"4d3496e0-06ef-4601-9339-1bc2367fe18f"
  ))
eidw_day1 <- eidw_day1 %>% filter(flight_id %in% eidw_long_dist$flight_id)

source(here::here("R/plot-eidw-tma.R"))

plot_eidw_tma() + geom_point(
  data = eidw_day1 %>% select(flight_id, latitude, longitude)
  , mapping = aes(x = longitude, y = latitude, colour = flight_id)
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

Fig. \@ref(fig:egll-trajectory-example) and Fig. \@ref(fig:eidw-trajectory-example) show examples for long distances flown on 1. July 2017.
The trajectory data allows for the identification of the holdings and/or sequencing area.
This allows to separate the effect of the operational concepts.

(ref:distance-flown-segment) Actual distances flown for each segment of the arrival phase (i.e. EGLL, 1.-8. July 2017).

<!-- TO DO: clean figure - work out bearings -- do it for Dublin as well -->

```{r distance-flown-segment, fig.cap="(ref:distance-flown-segment)"}
segment_approach <- function(ds) {
  ds <- ds %>% group_by(flight_id) %>% arrange(flight_id, timestamp)

  c40s <- ds %>% filter(type == "P40") %>% 
    select(flight_id, C40_TIME = timestamp, C40_BRG = bearing, C40_DIST_FLOWN = distance_flown)
  hins <- ds %>% filter(type == "PHOLD") %>% filter(timestamp == min(timestamp)) %>% 
    select(flight_id, HIN_TIME = timestamp, HIN_BRG = bearing, HIN_DIST_FLOWN = distance_flown)
  houts<- ds %>% filter(type == "PHOLD") %>% filter(timestamp == max(timestamp)) %>%
    select(flight_id, HOUT_TIME = timestamp, HOUT_BRG = bearing, HOUT_DIST_FLOWN = distance_flown)
  gates<- ds %>% filter(type == "PLAND") %>% 
    select(flight_id, GATE_TIME = timestamp, GATE_BRG = bearing, GATE_DIST_FLOWN = distance_flown)

  ds2 <- left_join(c40s, hins, by = "flight_id") %>% 
    left_join(houts, by = "flight_id") %>%
    left_join(gates, by = "flight_id")

  ds2 <- ds2 %>% mutate(
     TIME_1 = difftime(HIN_TIME, C40_TIME, units = "min"),  DIST_1 = HIN_DIST_FLOWN - C40_DIST_FLOWN
    ,TIME_2 = difftime(HOUT_TIME, HIN_TIME, units = "min"), DIST_2 = HOUT_DIST_FLOWN - HIN_DIST_FLOWN
    ,TIME_3 = difftime(GATE_TIME, HOUT_TIME, units = "min"),DIST_3 = GATE_DIST_FLOWN - HOUT_DIST_FLOWN
    ) %>% ungroup()
  return(ds2)
}

tmp <- read_pts_trj("egll") %>% segment_approach() %>% select(flight_id, C40_BRG, DIST_1, DIST_2, DIST_3)
tmp <- tmp %>% tidyr::gather(key = SEGMENT, value = DIST_FLOWN, -flight_id, -C40_BRG )

ggplot(data = tmp, mapping = aes(x = cut(C40_BRG, breaks = c(0,58, 180, 280, 360)), y = DIST_FLOWN, fill = SEGMENT)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(. ~ SEGMENT) +
  ylim(0, 100) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5)
        ,legend.position = c(1, 0.65), legend.justification = c(1, 0)) +
  labs(x = "arrival entry sectors [radial from ARP]"
       ,y = "distance flown in arrival segment"
       ,fill = "arrival segment")
```

(ref:distance-flown-segment-eidw) Actual distances flown for each segment of the arrival phase (i.e. EIDW, 1.-8. July 2017).

```{r distance-flown-segment-eidw, fig.cap="(ref:distance-flown-segment-eidw)"}
tmp <- read_pts_trj("eidw") %>% segment_approach() %>% select(flight_id, C40_BRG, DIST_1, DIST_2, DIST_3)
tmp <- tmp %>% tidyr::gather(key = SEGMENT, value = DIST_FLOWN, -flight_id, -C40_BRG )

ggplot(data = tmp, mapping = aes(x = cut(C40_BRG, breaks = c(0, 160, 210, 260, 360)), y = DIST_FLOWN, fill = SEGMENT)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(. ~ SEGMENT) +
  ylim(0, 100) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5)
        ,legend.position = c(1, 0.65), legend.justification = c(1, 0)) +
  labs(x = "arrival entry sectors [radial from ARP]"
       ,y = "distance flown in arrival segment"
       ,fill = "arrival segment")
```

Fig. \@ref(fig:distance-flown-segment) depicts the flown distances per segment of the arrival phase.
In the case of London Heathrow the classical stack based holding can be seen in the significant variations of the actual distance flown per arrival direction.
Only limited sequencing takes place during the first segment (arrival within 40 NM and the stacks).
Fig. \@ref(fig:distance-flown-segment-eidw) shows different footprint. It follows from the distribution of the initial approach segment that this portion is already used for sequencing and synchronisation by Dublin approach.
The middle segment shows a wider variation and less uniform distribution than the approach at London Heathrow.

```{r}
ds <- read_pts_trj("eidw")

# pick RW28 point merge
pick <- ds %>% select(flight_id, type, holding_id) %>% group_by(flight_id) %>% 
  filter(type == "PHOLD", holding_id =="RWY28W") %>% unique() %>% ungroup() %>%
  pull(flight_id)

ds2 <- ds %>% filter(flight_id %in% pick) %>% segment_approach()

THRU_HIN <- ds2 %>% select(flight_id, HIN_TIME, HOUT_DIST_FLOWN, HIN_DIST_FLOWN)  %>% 
  mutate(BIN = lubridate::floor_date(HIN_TIME, unit = "20 min")) %>%
  group_by(BIN) %>% summarise(THRU_N = n(), MED_DIST = median(HOUT_DIST_FLOWN - HIN_DIST_FLOWN))
THRU_GATE<- ds2 %>% select(flight_id, GATE_TIME) %>% 
  mutate(BIN = lubridate::floor_date(GATE_TIME, unit = "20 min")) %>%
  group_by(BIN) %>% summarise(THRU_NG = n())

rq <- full_join(THRU_GATE, THRU_HIN, by = "BIN")
rq <- rq %>% mutate(THRU_N2 = lead(THRU_N,1), THRU_N3 = lead(THRU_N,2))
```

(ref:eidw-saturation-example) EIDW - arrival throughput saturation (1.-8. July 2017).

```{r eidw-saturation-example, fig.cap='(ref:eidw-saturation-example)'}
g <- ggplot(data = rq %>% na.omit, mapping = aes(x = as.factor(THRU_NG), y = THRU_N2 + THRU_N3)) + 
  geom_boxplot(position="identity") + 
  stat_summary(fun.y=function(x){result = quantile(x,p=0.5)}, geom="step" , direction = "vh"
               #, shape=20
               , size=2
               , color="red", fill="red", group = 1) +
   ylim(0,13) 
g + 
  theme_bw() +
  labs(x = "arrival throughput (runway [flights per 20 min]"
       , y = "sequencing segment throughout [flights per 20 min]")
```
Fig. \@ref(fig:eidw-saturation-example) depicts the variation of the arrival segment throughput at Dublin (EIDW).
The median for every set of flights in the ASMA area versus flights landing is highlighted.
The behavour is a classical saturation curve, i.e. up to an arrival bunching of 10-11 flights per 20 min (30-33 flights per hour), the point merge procedure seems to be able to support the tactical absorption and sequencing.
It is noteworthy that the saturation curve falls for higher arrival throughputs. As no mearurements for 12 through 14 landings had been made in the sample, this behaviour could be caused by the chosen time horizon.

# Conclusions

```{r}
# What is shown by this work and its significance.
```

The work reported establishes a fully reproducible paper for ANS performance in Europe.
It is the one of the first attempts to address the reproducibility crisis within ANS performance by providing the artifacts that allow an independent researcher or professional to reproduce the results and verify upstream processing steps.

```{r}
# Limitations and advantages.
```
This paper follows up on previous work [@koelle_open_2017; @spinielli_2017; @spinielli_2018] and forms part of the roll-out of the performance reference trajectory by the PRU.
As such the paper demonstrates the feasibility to the approach, the data analytical capabilities and preprocessing, and the reporting via the R/RStudio ecosystem.
This represents an essential milestone moving towards the envisaged target environment.
However, further work is required to establish the reproducible environment for ANS Performance.

To overcome these limitations, PRU will be looking into deploying the trajectory processing on a cloud-based platform including the regular provision of the identified analytical data levels.

The move to open data for ANS Performance is a *giant leap* in terms of transparency.
This will foster a new level of cooperation and discussion amongst researchers and professionals interested in this topic.
It removes the burden to acquire and preprocess large-scale surveillance data and associated aircraft data or aeronautical information.
PRU envisages a strong collaboration through a network of research and practitioners.
The departure point for this development was to address the lack and hurdle of verifying results.
A further advantage is that novel applications, methods and algorithms (e.g. machine learning), can be applied to a harmonised dataset.
It is hoped that development efforts can accelerate while offering a platform for improvements to existing processing methods and algorithms.

```{r}
# Applications of the results.
```
The major application of the paper results will inform the performance reference trajectory roll-out project.
This goes in hand with the iterative move of the operational performance measures to trajectory derived data.
This can take on the form of quality assurance measures (e.g. verifying the order of magnitude, complementing missing reported data), but also making the reference trajectory the primary means of data collection.
The latter will support the further development of the operational performance measures responding to the ICAO GANP.

<!--provide one or two examples - animate statement -->

```{r}
# Recommendations for further work.
```
This paper documents a key milestone in the project. 
As such the future work will address:

* roll-out and establishment of the performance reference trajectory: This includes the organisational and technical work to deploy the processing modules and arrange for a regular processing of the source data and its provision as Open Data.
* establishing the current performance reporting based on the trajectory: Since more than 2 years PRU addresses the need for Open Data by providing monthly updates for the key performance indicators used under the EUROCONTROL Performance Review System and the SES Performance Scheme. Underlying analytic data is made available via [http://ansperformance.eu](http://ansperformance.eu). As part of the roll-out, it is foreseen to move associated indicators to a trajectory derived or at least verified process and expand on the analytical data level.
* establishment of a network: The underlying idea is that various levels of future work can be orchestrated in close collaboration with interested researchers and professionals. This can span work addressing the improvement of the actual data processing (e.g. less resource intensive computing) to the further development and validation of novel techniques for measuring operational performance (e.g. application of machine learning, time series analysis).

# References {-}
\vspace{6pt}
\setlength{\parskip}{0pt}
\footnotesize
\setlength{\parindent}{0cm}

### 2019-02-18 paper.tex
initial commit to euctrl-pru repo

%% template.tex
%% from
%% bare_conf.tex
%% V1.4b
%% 2015/08/26
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
%% conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/pkg/ieeetran
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE!
%% User assumes all risk.
%% In no event shall the IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

\documentclass[conference,final,a4paper,]{IEEEtran}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/pkg/ifpdf
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of the IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 5.0 (2009-03-20) and later if using hyperref.sty.
% The latest version can be obtained at:
% http://www.ctan.org/pkg/cite
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at:
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics.
%
% Note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/amsmath

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/pkg/algorithms
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/pkg/algorithmicx

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/array

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
% for subfigure.sty, the latter of which is no longer maintained and is
% incompatible with some LaTeX packages including fixltx2e. However,
% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
% which will override IEEEtran.cls' handling of captions and this will result
% in non-IEEE style figure/table captions. To prevent this problem, be sure
% and invoke subfig.sty's "caption=false" package option (available since
% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
% handling of captions.
% Note that the Computer Society format requires a larger sans serif font
% than the serif footnote size font used in traditional IEEE formatting
% and thus the need to invoke different subfig.sty package options depending
% on whether compsoc mode has been enabled.
%
% The latest version and documentation of subfig.sty can be obtained at:
% http://www.ctan.org/pkg/subfig

% *** FLOAT PACKAGES ***
%

%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure.
% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
% corrections already built into the system in which case a warning will
% be issued if an attempt is made to load fixltx2e.sty as it is no longer
% needed.
% The latest version and documentation can be found at:
% http://www.ctan.org/pkg/fixltx2e

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/stfloats
% Do not use the stfloats baselinefloat ability as the IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that the IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
% \usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/pkg/dblfloatfix

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/url
% Basically, \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

%% BEGIN MY ADDITIONS %%

\usepackage{longtable,booktabs}

\usepackage[unicode=true]{hyperref}

\hypersetup{
            pdftitle={Busting the Myth! - Reproducibility in ANS Performance},
            pdfauthor={Enrico Spinielli, Rainer Koelle},
            pdfkeywords={reproducibility, ANS performance, flight trajectory, open data, performance indicator},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{5}

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}

% Pandoc header
%\def\tightlist{}

% \def\tightlist{%
%   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% \renewcommand*{\bibfont}{9pt}

% \usepackage[utf8x]{inputenc}
% \usepackage{mathptmx}

% \usepackage[T1]{fontenc}

% section title small caps issue
% Uncomment and knit once in order to force use of Times font
% as per IEEE Trans style. It can then stay commented out.
% \usepackage{times}

\usepackage{breakurl}
\usepackage{graphicx}

%% DO NOT use ULEM: it converts italics to underlined
% \usepackage{ulem} % ulem is needed to support strikethroughs (\sout) [with pandoc extension]

% \usepackage{biblatex}
\usepackage{doi}

\usepackage{float}

% as requested by conference Editor (plus hack after References section in Rmd)
\usepackage[skip=6pt,indent]{parskip}

% enable intentionally default disabled IEEE commands: needed for biographies
\IEEEoverridecommandlockouts

\def\IEEEkeywordsname{Keywords}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%% END MY ADDITIONS %%

\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Busting the Myth! - Reproducibility in ANS Performance}

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations

\author{

%% ---- classic IEEETrans wide authors' list ----------------
 % -- end affiliation.wide
%% ----------------------------------------------------------

%% ---- classic IEEETrans one column per institution --------
 %% -- beg if/affiliation.institution-columnar
\IEEEauthorblockN{
  %% -- beg for/affiliation.institution.author
Enrico Spinielli,
  %% -- beg for/affiliation.institution.author
Rainer Koelle %% -- end for/affiliation.institution.author
}
\IEEEauthorblockA{Performance Review Unit\\
EUROCONTROL\\
Brussels, Belgium
  %% -- beg for/affiliation.institution.author
\\enrico.spinielli@eurocontrol.int
  %% -- beg for/affiliation.institution.author
\\rainer.koelle@eurocontrol.int
 %% -- end for/affiliation.institution.author
}
 %% -- end for/affiliation.institution
 %% -- end if/affiliation.institution-columnar
%% ----------------------------------------------------------

%% ---- one column per author, classic/default IEEETrans ----
 %% -- end if/affiliation.institution-columnar
%% ----------------------------------------------------------

}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
%
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3},
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Air transportation is undergoing a fundamental transformation. On the
political, strategic, and operational level work is directed to meet the
future growth of air traffic. To meet this goal, operational concepts
and technological enablers are deployed with the promises of distinct
performance improvements. These improvements are typically widely
marketed, however, the underlying evidence is not made available. From
that respect Air Navigation System Performance suffers from the same
credibility crisis seen in other industries. This paper implements the
reproducibility paradigm by providing data, code, and results openly to
enable another researcher or analyst to reproduce - and potentially -
improve or build on our work. This work is part of a wider thread to
establish Open Data and Open Source based data analytical capabilities
for Air Navigation System Performance in Europe. The implementation of
the paradigm is demonstrated on the basis of the performance reference
trajectory by replicating a classical performance measure, and then
moving into investigating the application of trajectory based
information for enhancing the current state of performance monitoring
for the arrival phase. This paper applies the concept and approach as an
use case analysis for 2 European airports with different strategies for
sequencing arrival traffic (i.e.~stack holding and point merge). The
work reported demonstrates the feasibility of the approach and reflects
a stepping stone in the roll-out of the performance reference trajectory
within the European context. The work also builds the basis for novel
developments on the basis of the implemented reproducibility workflow as
this paper is accompanied by a webpage that supports the impartial
evaluation of the work by providing the data, code, and how this reports
had been produced.
\end{abstract}

% keywords
\begin{IEEEkeywords}
\textit{%
reproducibility; ANS performance; flight trajectory; open data; performance indicator}
\end{IEEEkeywords}

% use for special paper notices

% make the title area
\maketitle

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}\label{introduction}

Air transportation is undergoing a fundamental transformation. To meet
the future growth of air traffic, political decision makers, strategic
planers, practitioners and researchers aim to enhance the performance of
air navigation services (ANS). At the time being there is no consistent
and regionally harmonised deployment programme. Typically, the
deployment programmes are revolving around public-private partnerships
and regional funding programmes resulting in a wide mix of consortia and
stakeholder subsets. The actual deployment of novel concepts or
technological enablers is then driven by the programmes and to a lesser
extent by operational network wide priorities following a piece meal
approach.

Moreover, the success of these deployments is typically supported by
promising business cases and operational evaluations - often expressed
as operational performance improvements. While these improvements are
marketed, the underlying data and associated analyses in support of the
results are seldomly made available. Explanations range from data
sensitivity, commercial interests and property rights, through to
complexity of the data analyses or simulations, and the associated
volume of data.

Accordingly, ANS performance improvements are subject to a ``credibility
crisis'' similar to reported problems in other domains, e.g.~medicine
{[}\protect\hyperlink{ref-kolata_2011}{1}{]}, psychology
{[}\protect\hyperlink{ref-ritchie_2012}{2}{]}, ecology
{[}\protect\hyperlink{ref-fraser_etal_2018}{3}{]}. Fidler and Wilcox
(2018) provides a recent review of the reproducibility discussion
{[}\protect\hyperlink{ref-fidler_wilcox_2018}{4}{]}. Part of the crisis
sensation was driven by the diffusion of reported results as other
researchers/analysts were able to discover substantial flaws in the data
analyses. However, a crisis also offers room for self-correction.
Throughout the recent years, this lead to revisitng the concept of
reproducible research. This concept is based on the premises that an
independent researcher/analyst is enabled to reproduce the reported
results. Reproducibility covers therefore a spectrum and mix of method,
data, and analytical code/scripts per se openly available for scrutiny
by others.

Independent (operational) ANS performance builds on transparent and
robust performance monitoring and reporting. Within the European context
EUROCONTROL's Performance Review Unit (PRU) is committed to provide
impartial analyses of the performance of the European Air Navigation
System based on verified data. With a view to enhance transparency, PRU
is - inter alia - making performance-related data publicly available via
its Portal, \url{http://ansperformance.eu}, since early 2017 combined
with the methodological documentation for the calculation of the
performance indicators used under the EUROCONTROL Performance Revew
System and the European Union Performance Scheme. There has been
recognition of using common and openly available tools and data in ATM
research and operational validation. With the advent of crowd sourced
data and aviation databases there is an opportunity to advance the state
of the art by combining open data and government organisation collected
data to establish a curated platform supporting the quest for
reproducibility in ANS Performance. One of the associated building
blocks is the performance reference trajectory
{[}\protect\hyperlink{ref-spinielli_2018}{5}{]}. The performance
reference trajectory will be deployed as Open Data throughout 2019. The
trajectory project builds on crowd sourced surveillance data fused with
correlated position reports collected from European air navigation
service providers. Based on the Open Data model it will be possible to
augment the data processing (e.g.~data coverage, time synchronisation),
establishment of derived data bases (e,g fleet data base), associated
performance models (e.g.~aircraft performance model), but also the
continual further development and validation of operational performance
measures/indicators.

This paper builds on the previous work and implements the
reproducibility paradigm. As a use-case example, a fully reproducible
paper is devised assessing the performance in the arrival phase for 2
European airports. The underlying data, supporting data transformation
and analysis, and the paper ``source text'' are made openly available at
the paper companion web repository
\url{https://github.com/euctrl-pru/reproducible-ans-performance-paper}.
This enables the independent review and validation of the work reported.
Conceptually, an interested researcher/analyst can reproduce this paper
by using the same toolset or validate the results by using the source
data or interim data products if the complete reproduction of the study
would be too time/resource consuming.

The contributions of this paper are as follows:

\begin{itemize}
\tightlist
\item
  first fully reproducible ANS Performance related paper: The paper is
  written in RMarkdown allowing for the analysis of how text, code
  chunks, and visualisations are produced (including associated
  parameter settings) on the basis of the ``source text''.
\item
  transparent data analyses and result generation through supporting
  web-page: All data used in this paper, intermediate data artifacts,
  and results including the associated transformation are publicly
  available for inspection, validation, or reproduction.
\item
  use-case analysis of ANS Performance during the arrival phase for 2
  European airports: As an example the classical `additional ASMA time'
  performance indicator is implemented on the basis of the provided data
  with showcasing trajectory derived capabilities to further study the
  efficiency of the arrival phase (e.g.~impact of sequencing strategies
  {[}stack holdings, point merge{]}).
\end{itemize}

\section{Background -
Reproducibility}\label{background---reproducibility}

Conceptually, reproducibility is not a new topic. It is as old as the
scientific method and at the heart of the verification/validation of
results. The consistent and robust interpretation of results by
independent researchers/analysts forms the evidence and the inherent
body of knowledge in support of the applied methods or conclusions. This
principle is also relevant for the further development on the basis of
such results.

The terms replication and reproducibility are often used interchangeably
and there is a debate about the necessary and sufficient conditions for
these terms. Results are considered replicable, ``if there is sufficient
information available for independent researchers to make the same
{[}consistent{]} findings using the same procedures with new data''
({[}\protect\hyperlink{ref-gandrud_2015}{6}{]}, p.~3f). Replicability,
thus, does not aim to produce the ``same'' (aka identical) results. Here
the goal is to establish confirmation of the approach and findings by
deriving same conclusions.\footnote{For example, the same data
  collection procedures may not necessarily result in ``identical''
  results as the time horizon might be different, (new or different)
  data for another study case show different result values, but the
  conclusions are consistent with the replicated approach.} Accordingly,
results are reproducible, if an independent researcher can recreate the
findings based on the given information. In our domain (i.e.~operational
ANS Performance and associated data analysis) this will require the
access to data and derived summaries, how associated visualisations have
been generated, the accompanying explanatory text, and conclusions drawn
for the analysis.

Reproducibility has also to be seen in light of today's ``publish or
perish'' context. It appears that higher emphasis is given to the
frequency of publication than the validity of the produced results.
Within the medical world cases like the Duke ovarian cancer scandal
{[}\protect\hyperlink{ref-kolata_2011}{1}{]} are not single incidents,
but also demonstrate how errors during the data analysis - in this case
during the data preparation phase using spread sheets - can discredit
the work. Despite the increasing awareness, the adaption of
reproducibility or moving to other open scholarship/investigation
techniques is not yet achieved. This can be contrasted by emerging
practices like data journalism. One interesting example is buzzFeed's
article on the use of surveillance planes operated by law enforcement
and military for a variety of purposes. The authors were able to
demonstrate based on Open Data that these purposes range from tracking
drug trafficking to testing new surveillance technology
{[}\protect\hyperlink{ref-peteraldhous_2016}{7}{]}. The credibility of
the article comes from the fact that all artifacts are made available to
reproduce the findings or study the chosen approach and data
transformations made leading to the conclusions of the article.

Arguably there are limits to the access to underlying source data.
However, these limitations also offer an opportunity to address
reproducibility {[}\protect\hyperlink{ref-koelle_open_2017}{8}{]}. For
example, the data volume size might prohibit the further dissemination
of data, its collection is too resource intensive, or there exists
restrictive license agreements. Experience has shown that even in such
cases, a limited sample of the source data can be made available for
others to verify the initial data processing steps. For that purpose the
creation of an analytical data set as the output of the data collection,
gathering, and cleaning process stage can be understood as sufficient in
the sense of reproducible research. One key transparency aspect is the
documentation of the data preparation steps. The article Spies in the
Skies by Peter Aldhous and Charles Seife
{[}\protect\hyperlink{ref-peteraldhous_2016}{7}{]} is a good example for
a reproducible analysis of the US government's airborne surveillance
program. Though the license restrictions did not allow for the
publication of the underlying aircraft position data, the authors made
available the analytical data set and documented their analysis from the
data preparation through to the publication stage.

The reproducibility discussion focuses strongly on the recreation of
results by other researchers/analysts. A related problem is the
re-generating of already completed (and potentially published) results
with the - previously - available same input data. This is a fundamental
issue that can be summarised as ``can a future me re-do what I had done
in the past''? This problem space ranges across the complete data
analysis life-cyle, including archiving the orignially used input data
or the steps to generate the study data, documenting the data cleaning
and transformation steps including the parameters chosen, etc. By
definition such a capability is a precondition for reproduction or
replication by others. This implies that the analytical steps including
the steps to produce summaries or estimates, visualisations, and
associated conclusions are documented and archived. In order to enable
others to recreate the own findings, such steps and documentation would
need to be shareable. This consideration has a strong bearing on the
toolset chosen for this paper.

Some of the reasons why researchers/analysts do not publish data and
computational procedures that allow reproduction will be hard to change
{[}\protect\hyperlink{ref-daniel_nust_2018_2361304}{9}{]}. The
predominant technical reasons have been listed above. However, there are
also personal reasons such as the fear for floating ideas or loss of
competitive advantage. Other reasons deal with more technical problems,
e.g.~lack of capabilities, procedures.

Another incentive for the adoption of an open and reproducible research
practice can be seen in the wider recognition through increased
citations, media attention, discussion with the community and
collaborators, and ultimately funding opportunities. The recent years
have seen a strong use of research blogs and community platforms
supporting the publication of smaller steps or intermediate results. The
combination of web presentation and discussions offers a good basis for
an initial dialogue. Platforms like google scholar or researchgate.net
offer additional features for outreach and communication. This equally
applies for paper achives such as arxiv.org for pre-prints. Based on the
credibility crisis discussion a number of journals are now promoting -
or even requiring - the provision of analytical data in support of the
paper publication. There is also an increasing number of requirements
for national or European funded research with respect to data management
and Open Data access to the research.

Reproducibility is discussed in various domains. The following examples
further promote the use of open source software and/or data:

\begin{itemize}
\tightlist
\item
  archaeology: Marwick et al. (2017) focusses on computational
  reproducibility in archaeological research by providing basic
  principles. The case study of their implementation uses open source
  software (R/Rmarkdown/git approach)
  {[}\protect\hyperlink{ref-marwick_2017}{10}{]}. The paper advocates
  the publishing (analytical) data in support of the paper. This should
  not be an issue, even if there are constraints on the overall study
  data.
\item
  engineering / signal processing: Vandewalle (2009) discuss the
  motivation and approach to reproducibility in signal processing
  (i.e.~what, why, and how). Also here the case is made to open access
  to the code and data for validation purposes
  {[}\protect\hyperlink{ref-vandewalle_2009}{11}{]}.
\item
  operational ANS Performance monitoring: Spinielli et al. (2018)
  highlight the development and use of Open Flight Trajectories for
  reproducible ANS {[}\protect\hyperlink{ref-spinielli_2018}{5}{]}.
\end{itemize}

\begin{figure}[hbt]

{\centering \includegraphics[width=1\linewidth]{figures/reproducible-research-spectrum}

}

\caption{Reproducible research, Fig. 2 in
{[}\protect\hyperlink{ref-marwick_2017}{10}{]}.}\label{fig:reproducible-research}
\end{figure}

Across the literature, researchers have identified a ``reproducibility
spectrum'' ranging from the extreme, i.e.~publication only ( advertising
and marketing of the done work), to almost or even fully replicable
research. The work of Marwick et al. (2017)
{[}\protect\hyperlink{ref-marwick_2017}{10}{]} and Peng (2011)
{[}\protect\hyperlink{ref-Peng1226}{12}{]} uses the analogy of a
spectrum that can be mapped against the scoring of Vandewalle (2009)
describing the level of reproducibility on a 0 to 5 scale
{[}\protect\hyperlink{ref-marwick_2017}{10}{]}. In light of
{[}\protect\hyperlink{ref-leek_2015}{13}{]}, this paper conceptualises
reproducibility as the ``ability to recompute data analytic results
given an observed dataset and knowledge of the data analysis pipeline''
and accordingly ``replicability of a study is the chance that an
independent experiment targeting the same scientific question will
produce a consistent result''.

With Fig. \ref{fig:reproducible-research} and the discussion in this
section, this paper implements a fully reproducible workflow. This is
achieved by establishing this paper with open source software which
allows to combine text, code, and visualisations. The associated source
text can also easily be read with a simple text editor/viewer. From that
perspective, the text is fully inspectable. This includes the majority
of visualisations which are generated based on the data. Accordingly,
the parameters and settings chosen can be validated by an independent
researcher/analyst. Last but not least, the data for this paper is made
available in conjunction with this text as Open Data. This also includes
intermediate data artifacts which allows for the reproduction of the
complete work or portions of it. The conceptual approach and its
building blocks will be presented in more detail in the next section.

\section{Method / Approach}\label{method-approach}

\subsection{Conceptual Approach}\label{conceptual-approach}

\begin{figure}[hbt]

{\centering \includegraphics[width=1\linewidth]{figures/approach-stages}

}

\caption{Conceptual approach to reproducibility by PRU.}\label{fig:concept-approach}
\end{figure}

This paper builds on the conceptual approach presented in Fig.
\ref{fig:concept-approach} to implement replicable data analyses for
performance monitoring and associated research. On the data level this
approach abstracts three levels of data:

\begin{itemize}
\tightlist
\item
  source data
\item
  analytical data
\item
  results (and performance related data)
\end{itemize}

The source data for this paper comprises the PRU Performance Reference
Trajectory and operator reported data under the airport operator data
flow. This study also uses some aeronautical information for the use
case airports accessible through AIP or community maintained open data
archives (e.g.~http://openaip.net, http://ourairports.com). These data
flows are processed and stored in the EUROCONTROL data Warehouse. The
approach recognises that the full source data may or cannot be provided
for a variety of reasons. These include volume of the source data sets,
technical constraints allowing access to the source data storage, and
data sharing policies or use limitations. For the latter cases, PRU
promotes the sharing of a so-called representative sample. The idea is
that with the provision of the associated processing code, other
researchers/practitioners are able to reconstruct a part of the analytic
data - and subsequently part of the reported results - on the basis of
the provided source data set.

Analytical data refers to the data sets established based on the
aforementioned processing code. An essential defining characteristic of
analytic data is that it is the data basis on which the reported results
can be produced. The abstraction between source data and analytic data
allows for the sharing of such data as limitations are typically only
appliable to the source data. We follow the notion of
{[}\protect\hyperlink{ref-marwick_2017}{10}{]} by referring to the
provision of a representative sample of 8 days accompanying this paper
as source data.

As depicted in Fig. \ref{fig:reproducible-research} the provision of the
data, code, and text allows to move to the right part of the
reproducibility spectrum, away from marketing and final results to
credible, transparent, and challengeable analysis. Accordingly, the
results (or performance related data) level presents the final
abstraction.

\subsection{Toolset}\label{toolset}

This work builds on the RStudio tools for the R language
{[}\protect\hyperlink{ref-rcoreteam_2018}{14}{]} including Git (and the
web-based repository managers GitHub and GitLab) as underlying version
control system. The R language was originally developed within the
statistical community supporting the task of statistical reporting by
providing routines for the statistical computing and visualisation.
Being open source, the R community is actively engaging and sharing
related software packages to enhance the core functionality. Without
limiting the impact of other packages, the development of knitr
{[}\protect\hyperlink{ref-R-knitr}{15}{]} and RMarkdown
{[}\protect\hyperlink{ref-xie2018}{16}{]}, ggplot
{[}\protect\hyperlink{ref-wickham_2016}{17}{]} for visualisation, and
the so-called tidyverse packages and RStudio IDE
{[}\protect\hyperlink{ref-rstudioteam_2015}{18}{]} provide the basis for
an open source data analysis ecosystem.

A key feature for the implementation of the reproducibility paradigm is
the fact that RMarkdown supports the combination of text, coding, and
visualisations in a single document. The latter can be rendered into
various output formats, e.g.~classical paper report-styles (pdf, MS
Word), html web-pages, and slides (both classical Powerpoint style or
html-based presentations). The fact that outputs can be rendered to html
pages also supports the embedding of interactive graphics. This also
facilitates the communication and outreach amongst researchers/analysts
as standard web technology becomes a standard outlet format. It follows
that changes to the code (and associated visualisations) or text can be
readily updated by re-rendering the adapted report. This allows to
develop standard templates for recurring reports based on varying input
data, but it also serves the iterative development of larger reports
(and studies).

A minimal example of RMarkdown document, a plain-text file with the
conventional extension \texttt{.Rmd} is shown below:

\footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\OtherTok{---}
\FunctionTok{title:}\AttributeTok{ }\StringTok{"R Markdown example"}
\FunctionTok{author:}\AttributeTok{ }\StringTok{"John Doe"}
\FunctionTok{date:}\AttributeTok{ }\StringTok{"2019-02-17"}
\FunctionTok{output:}\AttributeTok{ html_document}
\OtherTok{---}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{Some math }
\NormalTok{$\textbackslash{}frac\{1\}\{N\} \textbackslash{}sum_\{i=1\}^N (x_i -\textbackslash{}mu)^2$,}
\NormalTok{followed by numbered list:}

\NormalTok{1. }\FloatTok{*italics* and **bold** text}
\FloatTok{1. then bullets}
\FloatTok{    - simple, eh?}

\NormalTok{and a code chunk:}

\NormalTok{```\{r\}}
\NormalTok{library(ggplot2)}
\NormalTok{fit = lm(mpg ~ wt, data = mtcars)}
\NormalTok{b   = coef(fit)}
\NormalTok{ggplot(mtcars, aes(x = wt, y = mpg)) +}
\NormalTok{  geom_point() + }
\NormalTok{  geom_smooth(method = lm, se = TRUE)}
\NormalTok{```}

\NormalTok{The slope of the regression is }\BaseNTok{`r b[1]`}\NormalTok{.}
\NormalTok{```}
\end{Highlighting}
\end{Shaded}

\normalsize

\begin{figure}[htb]

{\centering \includegraphics[width=1\linewidth]{figures/rmd-example}

}

\caption{The output of a minimalist example of RMarkdown.}\label{fig:rmd-example}
\end{figure}

There are three components of an R Markdown document: the metadata, the
text and the code. The metadata, written between the pair of three
dashes \texttt{-\/-\/-}, helps in the production of output document
(title, type of document {[}web page, PDF, MS Word{]}, \ldots{}) The
text, which follows the metadata, is witten using Markdown, a simplified
formatting syntax. There are two types of (computer) code:

\begin{itemize}
\item
  A code chunk starts with three backticks and \texttt{r} indicates the
  language name,\footnote{R Markdown supports 50 other languages.} and
  ends with three backticks.
\item
  An inline R code expression starts with backtick r and ends with a
  backtick.
\end{itemize}

Code portions are executed when the document is rendered.

The result of rendering the R Markdown above is shown in Fig.
\ref{fig:rmd-example}.

A particular strong point is that the R/Rstudio ecosystem is open source
software. This includes also its extensibility with other packages.
There exists a wide variety of packages for data analytical tasks and
techniques. This mechanism can also be useful for the communication of
results. Researchers/analysts are able to provide their code and
supporting documentation as a package for other researchers to build on.
For example, throughout the last year, PRU has developed certain utility
functions for reoccurring tasks of the data analytical pipeline. It is
now planned to wrap these functions into a package saving other
researchers/analysts of recoding certain steps. For this paper, the
functions and data transformations are reported as part of the RMarkdown
and supporting R files. All these files are available via the project
webpage.

\subsection{Capabilities: Performance Reference
Trajectory}\label{capabilities-performance-reference-trajectory}

Throughout the recent year PRU developed the data analytical
capabilities to fuse correlated position reports and ADSB surveillance
data. This work has been reported in
{[}\protect\hyperlink{ref-koelle_open_2017}{8}{]} and
{[}\protect\hyperlink{ref-spinielli_2018}{5}{]}. The idea is to
establish a pan-European air situation picture openly available to
research and practitioners. Conceptually, the fused performance
reference trajectory will be a key milestone in combatting marketing
myths. For example, operational improvements can be directly assessed by
interested analysts with a view to validate the observable performance
improvements. This allows for the application of commonly accepted
methods and models to express performance benefits and - ultimately -
challenge related publications on deployments and achieved improvements.

End of 2018 the development project was closed and PRU is now
investigating options for deploying and maintaining the processing
modules on a cloud platform. The goal is to establish and operate the
underlying data processing capabilities throughout 2019 and launch the
regular provision of the performance reference trajectory as Open Data.
It is also planned to establish a network and complement the air
situation picture by deploying additional sensors or determining
non-ADSB derived positions as a community effort. As a mid-term goal
associated data bases and models for performance shall be established
and jointly maintained.

\section{Results}\label{results}

\subsection{Data Preparation}\label{data-preparation}

\begin{figure}[htb]

{\centering \includegraphics[width=1\linewidth]{figures/data-preparation-pipeline}

}

\caption{Data preparation pipeline.}\label{fig:data-preparation-pipeline}
\end{figure}

The raw data sources for the use cases of this paper have been processed
by a pipeline of R scripts. For example, the additional ASMA time
indicator and the anticipated further breakdown of the arrival phase can
be described by a set of tranistion points:

\begin{itemize}
\tightlist
\item
  segment I: entry into the ASMA area to entry into the
  holding/vectoring segment
\item
  segment II: holding/vectoring segment; for the use case reported
  holding stacks or point merge procedure
\item
  segment III: final approach alignment and approach
\end{itemize}

Figure \ref{fig:data-preparation-pipeline} shows how the transition
points (40 NM, entry/exit from holding/point-merge area, landing) for
the use case airports have been extracted. Due to the size of the daily
reference trajectory files, the pipeline operates on one date at the
time. Assuming the relevant RT flight and position files are downloaded
from GCP as described in
{[}\protect\hyperlink{ref-performancereviewunit_2019}{19}{]}, the steps
and the relevant scripts (available in the companion repository for this
paper) used are:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  the script \texttt{R/extract-arrivals.R} extracts arrivals to the
  \texttt{APT} airport and generates:

  \begin{itemize}
  \tightlist
  \item
    a file with flight details
  \item
    a file with all the arrivals' positions
  \item
    a file with all arrivals' positions within 40 NM the airport's
    reference position (ARP)
  \end{itemize}
\item
  the script \texttt{R/extract-transition-points.R} extracts transition
  points in the terminal area and generates a file containing per each
  flight:

  \begin{itemize}
  \tightlist
  \item
    the first point within 40 NM, \texttt{P40}
  \item
    the first and last point within the holding/point-merge area,
    \texttt{PHOLD}
  \item
    an estimated landing point, \texttt{PLAND}
  \end{itemize}
\end{enumerate}

\begin{figure}[hbt]

{\centering \includegraphics[width=1\linewidth]{figures/eidw-arrival-chart-east}

}

\caption{Excerpt of Irish AIP showing point merge
arc layout East of Dublin Airport.}\label{fig:eidw-arrival-chart-east}
\end{figure}

This stage makes use of data extracted/extrapolated from the relevant
AIP, see for example Fig. \ref{fig:eidw-arrival-chart-east}, such as
ARP, runway name and threshold position/elevation, polygons for holding
stacks or point-merge areas. Specifically the polygons have been
heuristically defined and enlarged by 2NM in order to cater for the
spread of real flown trajectories, see the scripts
\texttt{R/egll-data.R} and \texttt{R/eidw-data.R}.

Each position mentioned above consist of at least

\begin{itemize}
\tightlist
\item
  flight ID
\item
  4D position (timestamp \(T\), longitude, latitude, altitude)
\item
  flown distance \(D\)
\item
  sequence number
\item
  distance from ARP
\end{itemize}

These allow to calculate for each flight

\begin{itemize}
\tightlist
\item
  the time spent before the holding/point-merge area,
  \(T_{PHOLD|first} - T_{P40}\)
\item
  the distance flown before the holding/point-merge area,
  \(D_{PHOLD | first} - D_{P40}\)
\item
  the time spent within the holding/point-merge area,
  \(T_{PHOLD | last} - T_{PHOLD|first}\)
\item
  the distance flown within the holding/point-merge area,
  \(D_{PHOLD | last} - D_{PHOLD | first}\)
\item
  the time spent out of the holding/point-merge area till landing,
  \(T_{PLAND} - T_{PHOLD | last}\)
\item
  the distance flown out of the holding/point-merge area till landing,
  \(D_{PLAND} - D_{PHOLD | last}\)
\end{itemize}

\subsection{Data Coverage}\label{data-coverage}

The previous sections described the data preparatory action. For this
paper, a reproducible data set for the analysis of arrival management
techniques at London Heathrow and Dublin airport have been extracted.
The data set comprises flights for the period of 1.-8. August 2017. It
can be inferred from the traffic mix at London Heathrow (major
international hub) that only a limited subset of flights is not equipped
with ADSB, while a slightly higher level of non-equipped flights is
expected at Dublin (higher share of intra-European flights).

\begin{figure}[hbt]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/coverage-egll-1}

}

\caption{Number of arrivals in data sets, i.e.~1.-8. July
2017.}\label{fig:coverage-egll}
\end{figure}

Fig. \ref{fig:coverage-egll} shows the overlap of the data between the
arrivals extracted from the airport operator data flow (APDF) and the
reference trajectory (RTRJ) for London Heathrow (EGLL) and Dublin
(EIDW).

For the time horizon of this paper, the coverage of the reference
trajectory derived data is for Heathrow (EGLL) 100\% and for Dublin
97\%.

\subsection{\texorpdfstring{``replicate'' ASMA
(time-based)}{replicate ASMA (time-based)}}\label{replicate-asma-time-based}

The additional time in the arrival sequencing and metering area (ASMA)
is a globally recognised measure to assess flight efficiency during the
arrival phase. The ASMA metric calculates the additional time based on a
reference time which is determined for a family of flights sharing
similar characteristics. These are the entry sector into the arrival
airspace, weight turbulence category and engine type, and the actual
landing runway. For each flight \(i\) the additional ASMA time is

\[{add.~ASMA~time}_i = {actual~travel~time}_i - {reference~time}_i\]

The reference time \(i\) is determined for flights with similar arrival
characteristics (c.f.
\url{http://ansperformance.eu/references/definition/additional_asma_time.html}).
These include 1.) the arrival entry sector at 40NM from the airport, 2.)
the aircraft weigth turbulence category and engine type, and 3.) the
actual landing runway. The reference times are built for a longer time
horizon and extracted from the PRU data base as additional source data.

\begin{figure}[hbt]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/ASMA-travel-egll-1}

}

\caption{Distribution of actual ASMA travel times for EGLL
(1.-8. July 2017).}\label{fig:ASMA-travel-egll}
\end{figure}

As an example, the distribution of the actual arrival transit times for
arrivals into Heathrow (EGLL) are shown in Fig.
\ref{fig:ASMA-travel-egll}. Heathrow operates holding stacks to
tactically optimise the arrival throughput. Dependent on the traffic
situation, aircraft may experience longer holding times (i.e.~long tails
of boxplots in Fig. \ref{fig:ASMA-travel-egll}).

Fig. \ref{fig:ASMA-compare} compares the additional ASMA time extracted
from the PRU performance dashboard, the trajectory based calculation
with the dashboard derived reference time, and a 20th percentile
approach purely based on the trajectory data.

\begin{figure}[hbt]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/ASMA-compare-1}

}

\caption{Comparison of determined ASMA values, i.e.~1.-8. July
2017.}\label{fig:ASMA-compare}
\end{figure}

Fig. \ref{fig:ASMA-compare} shows the variation of the timebased
additional ASMA times determined for our sample. APDFNM refers to the
variant used for the EUROCONTROL Performance Review System. The
reference time is build on the basis of a one-year sammple. Applying
this reference time to the trajectory based travel times (i.e.~TRJ-REF),
there is a strong increase of the additional ASMA time for Heathrow
(EGLL), while in the case of Dublin (EIDW) reasonable decrease takes
place. Calculating the additional ASMA time exclusively with the
trajectory sample and applying a 20th percentile approach (TRJ-20PCT),
shows a significant drop in EGLL and a moderate increase at Dublin. The
high variation for Heathrow shows that the additional ASMA time is
strongly dependent on the time horizon. The pure trajectory bases
measure - by definition - includes the effect of the operational concept
(stack holdings) and distorts the measure. In the case of Dublin it can
be argued that the TRJ-20PCT measure reflects the time based ASMA
measure suffiently.

\subsection{Distance-based ASMA}\label{distance-based-asma}

conceptually, measuring flight inefficiency in terms of flight time
accounts for the airspace user perspective. The previous section has
also shown the high dependency of the time based measure on the sample
size or reference time attribution. Flight time is linked to engine time
and accordingly, assuming maneuvering for operational separation and
sequencing reasons, fuel burn. From a service provision perspective,
efficiency relates to the capability to provide the least distance flown
within the ASMA area. In that respect, the data analytical capability to
determine an associated additional distance flown metric addresses an
open gap.

\begin{figure}[H]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/ASMA-time-dist-1}

}

\caption{Relationship travel time and distance flown,
i.e.~1.-8. July 2017.}\label{fig:ASMA-time-dist}
\end{figure}

Fig. \ref{fig:ASMA-time-dist} shows the relationship between the actual
distance flown and the actual time flown. The variation of travel times
is influenced by varying airspeeds or weather phenomena (e.g.~wind).
This paper applies a simple 15th percentile approach to the distance
flown per arrival sector, aircraft type/engine, and landing runway.

\begin{table}[t]

\caption{\label{tab:tbl-asma-dist}Average additional distance flown (15th pct)}
\centering
\begin{tabular}{l|r|r|r}
\hline
AIRPORT & TOT\_ADD\_DIST & N\_ARRS & AVG\_ADD\_DIST\\
\hline
EGLL & 117037.79 & 5249 & 22.30\\
\hline
EIDW & 33079.22 & 2634 & 12.56\\
\hline
\end{tabular}
\end{table}

Tab. \ref{tab:tbl-asma-dist} lists the determined average additional
distance at the study airports.

\subsection{Arrival Management
Practices}\label{arrival-management-practices}

For an initial drill down, this paper uses arrival practices at London
Heathrow and Dublin. The approach to managing the arrival flow differs
significantly between both airports. Conceptually, London Heathrow
employs the ``pressure on the runway'' concept with a focus on achieving
and balancing a continuously high level of runway throughput
utilisation. This is achieved by controlling the demand through 4
arrival stacks that allow for the tactical sequencing on the last
portion of the approach. Dublin is one of the airports that pioneered
the implementation of ``point merge''. Point merge aims at replacing
circular holdings by longitudal holdings to establish the sequence based
on predefined legs.

From a performance perspective it is therefore interesting to identify
and measure the impact of the operational concept on the overall flight
efficiency during the arrival phase of flight.

\begin{figure}[H]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/egll-trajectory-example-1}

}

\caption{EGLL - long trajectory flown on 1. July
2017.}\label{fig:egll-trajectory-example}
\end{figure}

\begin{figure}[H]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/eidw-trajectory-example-1}

}

\caption{EIDW - long trajectory flown on 1. July
2017.}\label{fig:eidw-trajectory-example}
\end{figure}

Fig. \ref{fig:egll-trajectory-example} and Fig.
\ref{fig:eidw-trajectory-example} show examples for long distances flown
on 1. July 2017. The trajectory data allows for the identification of
the holdings and/or sequencing area. This allows to separate the effect
of the operational concepts.

\begin{figure}[H]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/distance-flown-segment-1}

}

\caption{Actual distances flown for each segment of
the arrival phase (i.e.~EGLL, 1.-8. July 2017).}\label{fig:distance-flown-segment}
\end{figure}

\begin{figure}[H]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/distance-flown-segment-eidw-1}

}

\caption{Actual distances flown for each
segment of the arrival phase (i.e.~EIDW, 1.-8. July 2017).}\label{fig:distance-flown-segment-eidw}
\end{figure}

Fig. \ref{fig:distance-flown-segment} depicts the flown distances per
segment of the arrival phase. In the case of London Heathrow the
classical stack based holding can be seen in the significant variations
of the actual distance flown per arrival direction. Only limited
sequencing takes place during the first segment (arrival within 40 NM
and the stacks). Fig. \ref{fig:distance-flown-segment-eidw} shows
different footprint. It follows from the distribution of the initial
approach segment that this portion is already used for sequencing and
synchronisation by Dublin approach. The middle segment shows a wider
variation and less uniform distribution than the approach at London
Heathrow.

\begin{figure}[H]

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/eidw-saturation-example-1}

}

\caption{EIDW - arrival throughput saturation
(1.-8. July 2017).}\label{fig:eidw-saturation-example}
\end{figure}

Fig. \ref{fig:eidw-saturation-example} depicts the variation of the
arrival segment throughput at Dublin (EIDW). The median for every set of
flights in the ASMA area versus flights landing is highlighted. The
behavour is a classical saturation curve, i.e.~up to an arrival bunching
of 10-11 flights per 20 min (30-33 flights per hour), the point merge
procedure seems to be able to support the tactical absorption and
sequencing. It is noteworthy that the saturation curve falls for higher
arrival throughputs. As no mearurements for 12 through 14 landings had
been made in the sample, this behaviour could be caused by the chosen
time horizon.

\section{Conclusions}\label{conclusions}

The work reported establishes a fully reproducible paper for ANS
performance in Europe. It is the one of the first attempts to address
the reproducibility crisis within ANS performance by providing the
artifacts that allow an independent researcher or professional to
reproduce the results and verify upstream processing steps.

This paper follows up on previous work
{[}\protect\hyperlink{ref-spinielli_2018}{5}{]},
{[}\protect\hyperlink{ref-koelle_open_2017}{8}{]},
{[}\protect\hyperlink{ref-spinielli_2017}{20}{]} and forms part of the
roll-out of the performance reference trajectory by the PRU. As such the
paper demonstrates the feasibility to the approach, the data analytical
capabilities and preprocessing, and the reporting via the R/RStudio
ecosystem. This represents an essential milestone moving towards the
envisaged target environment. However, further work is required to
establish the reproducible environment for ANS Performance.

To overcome these limitations, PRU will be looking into deploying the
trajectory processing on a cloud-based platform including the regular
provision of the identified analytical data levels.

The move to open data for ANS Performance is a \emph{giant leap} in
terms of transparency. This will foster a new level of cooperation and
discussion amongst researchers and professionals interested in this
topic. It removes the burden to acquire and preprocess large-scale
surveillance data and associated aircraft data or aeronautical
information. PRU envisages a strong collaboration through a network of
research and practitioners. The departure point for this development was
to address the lack and hurdle of verifying results. A further advantage
is that novel applications, methods and algorithms (e.g.~machine
learning), can be applied to a harmonised dataset. It is hoped that
development efforts can accelerate while offering a platform for
improvements to existing processing methods and algorithms.

The major application of the paper results will inform the performance
reference trajectory roll-out project. This goes in hand with the
iterative move of the operational performance measures to trajectory
derived data. This can take on the form of quality assurance measures
(e.g.~verifying the order of magnitude, complementing missing reported
data), but also making the reference trajectory the primary means of
data collection. The latter will support the further development of the
operational performance measures responding to the ICAO GANP.

This paper documents a key milestone in the project. As such the future
work will address:

\begin{itemize}
\tightlist
\item
  roll-out and establishment of the performance reference trajectory:
  This includes the organisational and technical work to deploy the
  processing modules and arrange for a regular processing of the source
  data and its provision as Open Data.
\item
  establishing the current performance reporting based on the
  trajectory: Since more than 2 years PRU addresses the need for Open
  Data by providing monthly updates for the key performance indicators
  used under the EUROCONTROL Performance Review System and the SES
  Performance Scheme. Underlying analytic data is made available via
  \url{http://ansperformance.eu}. As part of the roll-out, it is
  foreseen to move associated indicators to a trajectory derived or at
  least verified process and expand on the analytical data level.
\item
  establishment of a network: The underlying idea is that various levels
  of future work can be orchestrated in close collaboration with
  interested researchers and professionals. This can span work
  addressing the improvement of the actual data processing (e.g.~less
  resource intensive computing) to the further development and
  validation of novel techniques for measuring operational performance
  (e.g.~application of machine learning, time series analysis).
\end{itemize}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\vspace{6pt} \setlength{\parskip}{0pt} \footnotesize
\setlength{\parindent}{0cm}

\hypertarget{refs}{}
\hypertarget{ref-kolata_2011}{}
{[}1{]} G. Kolata, ``How a New Hope in Cancer Fell Apart,'' \emph{The
New York Times: Health}, 07-Jul-2011.

\hypertarget{ref-ritchie_2012}{}
{[}2{]} S. J. Ritchie, R. Wiseman, and C. C. French, ``Failing the
future: Three unsuccessful attempts to replicate Bem's `Retroactive
Facilitation of Recall'Effect,'' \emph{PloS one}, vol. 7, no. 3, p.
e33423, 2012.

\hypertarget{ref-fraser_etal_2018}{}
{[}3{]} H. Fraser, T. Parker, S. Nakagawa, A. Barnett, and F. Fidler,
``Questionable research practices in ecology and evolution,'' \emph{PloS
one}, vol. 13, no. 7, p. e0200303, 2018.

\hypertarget{ref-fidler_wilcox_2018}{}
{[}4{]} F. Fidler and J. Wilcox, ``Reproducibility of Scientific
Results,'' in \emph{The Stanford Encyclopedia of Philosophy}, Winter
2018., E. N. Zalta, Ed. Metaphysics Research Lab, Stanford University,
2018.

\hypertarget{ref-spinielli_2018}{}
{[}5{]} E. Spinielli, R. Koelle, K. Barker, and N. Korbey, ``Open Flight
Trajectories for Reproducible ANS Performance Review,'' presented at the
SESAR Innovation Days 2018, 2018, p. 8.

\hypertarget{ref-gandrud_2015}{}
{[}6{]} C. Gandrud, \emph{Reproducible Research with R and R Studio,
Second Edition (Chapman \& Hall/CRC The R Series)}. Routledge, 2015.

\hypertarget{ref-peteraldhous_2016}{}
{[}7{]} C. S. Peter Aldhous, ``Spies in the Skies,'' \emph{See Maps
Showing Where FBI Planes Are Watching From Above}, 06-Apr-2016.
{[}Online{]}. Available:
\url{https://www.buzzfeed.com/peteraldhous/spies-in-the-skies}.

\hypertarget{ref-koelle_open_2017}{}
{[}8{]} R. Koelle, ``Open Source Software and Crowd Sourced Data for
Operational Performance Analysis,'' presented at the USA/Europe ATM R\&D
Seminar, 2017.

\hypertarget{ref-daniel_nust_2018_2361304}{}
{[}9{]} D. Nüst, M. Schutzeichel, and M. Konkol, ``Opening Reproducible
Research: A research project website and blog,'' Nov. 2018.

\hypertarget{ref-marwick_2017}{}
{[}10{]} B. Marwick, ``Open Science in Archaeology,'' \emph{Open Science
Framework}, Jan. 2017.

\hypertarget{ref-vandewalle_2009}{}
{[}11{]} P. Vandewalle, J. Kovacevic, and M. Vetterli, ``Reproducible
research in signal processing,'' \emph{IEEE Signal Processing Magazine},
vol. 26, no. 3, pp. 37--47, May 2009.

\hypertarget{ref-Peng1226}{}
{[}12{]} R. D. Peng, ``Reproducible Research in Computational Science,''
\emph{Science}, vol. 334, no. 6060, pp. 1226--1227, 2011.

\hypertarget{ref-leek_2015}{}
{[}13{]} J. T. Leek and R. D. Peng, ``Opinion: Reproducible research can
still be wrong: Adopting a prevention approach,'' \emph{Proc Natl Acad
Sci USA}, vol. 112, no. 6, p. 1645, Feb. 2015.

\hypertarget{ref-rcoreteam_2018}{}
{[}14{]} R Core Team, \emph{R: A Language and Environment for
Statistical Computing}. Vienna, Austria: R Foundation for Statistical
Computing, 2018.

\hypertarget{ref-R-knitr}{}
{[}15{]} Y. Xie, \emph{Knitr: A General-Purpose Package for Dynamic
Report Generation in R}. 2018.

\hypertarget{ref-xie2018}{}
{[}16{]} Y. Xie, J. Allaire, and G. Grolemund, \emph{R Markdown: The
Definitive Guide}. Boca Raton, Florida: Chapman and Hall/CRC, 2018.

\hypertarget{ref-wickham_2016}{}
{[}17{]} H. Wickham, \emph{Ggplot2: Elegant Graphics for Data Analysis}.
Springer-Verlag New York, 2016.

\hypertarget{ref-rstudioteam_2015}{}
{[}18{]} RStudio Team, \emph{RStudio: Integrated Development Environment
for R}. Boston, MA: RStudio, Inc., 2015.

\hypertarget{ref-performancereviewunit_2019}{}
{[}19{]} Performance Review Unit, ``Reference trajectories for
Performance Review of European ANS,'' 2019. {[}Online{]}. Available:
\url{https://doi.org/10.5281/zenodo.2566844}.

\hypertarget{ref-spinielli_2017}{}
{[}20{]} E. Spinielli, R. Koelle, M. Zanin, and S. Belkoura, ``Initial
Implementation of Reference Trajectories for Performance Review,''
\emph{SESAR Innovation Days}, no. 7, p. 8, 2017.

\setlength{\parskip}{6pt}

\vspace*{1\baselineskip}
\hypertarget{biographies}{%
\section*{Authors' Biographies}\label{biographies}}
\addcontentsline{toc}{section}{Authors' Biographies}

\vspace*{-3\baselineskip}

\begin{IEEEbiographynophoto}{\normalsize Enrico Spinielli}\normalsize
is the Data Manager of the Performance Review Unit at EUROCONTROL, Brussels.
He holds a Master Degree in Electronic Engineering from the University of Pisa, Italy.
His intellectual interests include data analytics, data visualization and
map projections.
\end{IEEEbiographynophoto}

\vspace*{-2.5\baselineskip}

\begin{IEEEbiographynophoto}{\normalsize Rainer Koelle}\normalsize
heads the Safety and Quality of Service section
of the Performance Review Unit at EUROCONTROL, Brussels.
He holds a PhD from the University of Lancaster, 2013, in Communication Systems.
His research interests include data analytics for operational performance analysis.
\end{IEEEbiographynophoto}

\vfill

\end{document}

### 2019-02-18 R/analyse-transition-points.R

@@ -1,0 +1,67 @@
+library(fs)
+library(here)
+library(purrr)
+library(readr)
+library(dplyr)
+library(skimr)
+
+egll_transitions <- dir_ls(path = here("data"),
+                           regexp = "egll_transition_points_.*.csv") %>%
+  map_dfr(~ read_csv(.))
+
+
+# TODO: add assert
+# # just one pair P40 -- PLAND per flight!
+# egll_transitions %>%
+# dplyr::group_by(flight_id) %>%
+# arrange(flight_id, timestamp) %>%
+# select(flight_id, distance_flown, type) %>%
+# filter(type %in% c("P40", "PLAND")) %>% count() %>% filter(n > 2)
+
+
+flown_tma <- egll_transitions %>%
+  dplyr::group_by(flight_id) %>%
+  arrange(flight_id, timestamp) %>%
+  select(flight_id, distance_flown, type) %>%
+  filter(type %in% c("P40", "PLAND")) %>%
+  mutate(flown_tma = distance_flown - lag(distance_flown)) %>%
+  filter(!is.na(flown_tma)) %>%
+  select(-type, -distance_flown) %>%
+  ungroup()
+
+flown_holding <- egll_transitions %>%
+  filter(type %in% c("PHOLD")) %>%
+  group_by(flight_id, holding_id) %>%
+  arrange(flight_id, timestamp) %>%
+  select(flight_id, timestamp, holding_id, distance_flown, type) %>%
+  mutate(flown_holding = distance_flown - lag(distance_flown),
+         duration_holding = timestamp - lag(timestamp)) %>%
+  filter(!is.na(flown_holding)) %>%
+  select(-type, -distance_flown, -timestamp) %>%
+  ungroup()
+
+# TODO: in 8 days only few instances, still...
+# # flights crossing multiple holding stacks
+# flown_holding %>%
+#   arrange(flight_id) %>%
+#   group_by(flight_id) %>%
+#   filter(n() > 2)
+
+
+# TODO: decide what happens in case of crossing multiple holdings
+# from last PHOLD to PLAND, i.e. PLAND and the preceeding one, see slice()
+flown_to_runway <- egll_transitions %>%
+  group_by(flight_id) %>%
+  arrange(flight_id, timestamp) %>%
+  # take PLAND and the preceeding one
+  slice((n()-1):n()) %>%
+  select(flight_id, timestamp, holding_id, distance_flown, type) %>%
[…]

### 2019-02-18 R/asma.R

@@ -1,0 +1,102 @@
+# Utility function for the ASMA part of the paper
+# 
+# The functions fall broadly into X categories
+# 
+# 1. read functions for the analytical data, including variable renaming and type coercions
+# 
+
+
+# ================= READ IN ANALYTICAL DATA =======================
+
+read_arrs_apdf <- function(apt, .path = "./data/"){
+  data_file <- paste(.path, apt, "_arrs_fact_2017.csv", sep = "")
+  ds <- readr::read_csv(
+    data_file
+    ,col_types = cols(
+      CALLSIGN = col_character()
+      ,REG = col_character()
+      ,ADEP = col_character()
+      ,ADES = col_character()
+      ,ALDT = col_datetime(format = "")
+      ,AIBT = col_datetime(format = "")
+      ,STA = col_datetime(format = "")
+      ,TYP = col_character()
+      ,WTC = col_character()
+      ,C40_TIME = col_datetime(format = "")
+      ,C40_LAT = col_double()
+      ,C40_LON = col_double()
+      ,C40_ALT = col_double()
+      )
+    )
+  return(ds)
+}
+
+
+read_flts_trj <- function(apt, .path = "./data/"){
+  data_path <- .path
+  data_files<- dir(data_path, pattern = paste(apt, "_flt_rt", sep = ""))
+  
+  flts <- data_files %>%
+    purrr::map(~ read_csv(file.path(data_path, .),
+                   col_types = cols(
+                     period_start = col_datetime(format = "")
+                     ,period_finish = col_datetime(format = "")
+                   ))
+    ) %>%
+    dplyr::bind_rows()
+  return(flts)
+}
+
+
+# ================= COVERAGE DATA PREP =======================
+
+prep_cov_apdf <- function(ds){
+  ds <- ds %>% 
+    dplyr::mutate(
+      ARR_TIME = if_else(!is.na(AIBT), AIBT, STA) 
+      ,DOF     = lubridate::date(ARR_TIME)
+      ,SOURCE  = "APDF") %>% 
+    select(REG, TYP, ADEP, ADES, ARR_TIME, DOF, SOURCE)
+  return(ds)
+}
+
+
+prep_cov_trj <- function(ds){
+  ds <- ds %>% 
+    # rename variables
+    dplyr::select( REG  = aircraft_reg, TYP = aircraft_type
+                  ,ADEP = adep, ADES = ades
+           , period_start, period_finish) %>%
[…]

### 2019-02-18 R/egll-data.R

@@ -1,0 +1,332 @@
+library(tibble)
+library(dplyr)
+library(sf)
+
+# do not use trrrj yet, just copied needed functions in utils.R ...
+#library(trrrj)
+# source(here::here("R", "utils.R"))
+
+
+######### Airport Reference Point
+# Heathrow (from https://skyvector.com/airport/EGLL/London-Heathrow-Airport )
+# Location Information for EGLL
+# Coordinates: N51°28.65' / W0°27.68'
+# View all Airports in England, United Kingdom.
+# Elevation is 83.0 feet MSL.
+# Magnetic Variation is 0° West
+
+########## Runways
+# Runway 09L/27R
+# Dimensions:	12802 x 164 feet / 3902 x 50 meters
+# Surface:	Hard
+# Runway 09L	Runway 27R
+# Coordinates:	N51°28.65' / W0°29.10'	N51°28.66' / W0°26.00'
+# Elevation: 	79 	78
+# Runway Heading: 	091° 	271°
+# Displaced Threshold: 	1004 Feet
+#
+# Runway 09R/27L
+# Dimensions:	12008 x 164 feet / 3660 x 50 meters
+# Surface:	Unknown
+# Runway 09R	Runway 27L
+# Coordinates:	N51°27.89' / W0°28.94'	N51°27.90' / W0°26.04'
+# Elevation: 	75 	77
+# Runway Heading: 	091° 	271°
+# Displaced Threshold: 	1007 Feet
+
+########## VOR-DME's for HOLDING STACKS
+# EGLL arrival stacks, from http://worldaerodata.com/wad.cgi?nav=<NAV NAME>
+# NAV NAME = BOVINGDON, BIGGIN, ...
+
+# BOVINGDON
+# Type 	ID 	Channel 	Freq 	Country 	State
+# VOR-DME 	BNN 	084Y 	113.75 	United Kingdom 	-
+#   Latitude 	Longitude 	Airport
+# 51.726111
+# 51° 43' 34.00" N 	-0.549722
+# 000° 32' 59.00" W
+
+# BIGGIN
+# Type 	ID 	Channel 	Freq 	Country 	State
+# VOR-DME 	BIG 	098X 	115.1 	United Kingdom 	-
+#   Latitude 	Longitude 	Airport
[…]

### 2019-02-18 R/eidw-data.R

@@ -1,0 +1,130 @@
+library(tibble)
+library(dplyr)
+library(sf)
+
+# do not use trrrj yet, just copied needed functions in utils.R ...
+# library(trrrj)
+#source(here::here("R", "utils.R"))
+
+######### Airport Reference Point
+# Dublin (from https://skyvector.com/airport/EIDW/Dublin-Airport )
+# Location Information for EIDW
+# Coordinates: N53°25.28' / W6°16.20'
+# View all Airports in Dublin, Ireland.
+# Elevation is 242.0 feet MSL.
+# Magnetic Variation is 3° West
+
+########## Runways
+# Runway 10/28
+# Dimensions:	8652 x 148 feet / 2637 x 45 meters
+# Surface:	Hard
+# Runway 10	Runway 28
+# Coordinates:	N53°25.35' / W6°17.40'	N53°25.22' / W6°15.03'
+# Elevation: 	242 	202
+# Runway Heading: 	101° 	281°
+#
+# Runway 16/34
+# Dimensions:	6800 x 200 feet / 2073 x 61 meters
+# Surface:	Unknown
+# Runway 16	Runway 34
+# Coordinates:	N53°26.22' / W6°15.72'	N53°25.19' / W6°14.98'
+# Elevation: 	217 	202
+# Runway Heading: 	162° 	342°
+
+
+eidw_apt <- tibble::tribble(
+  ~latitude,     ~longitude, ~elevation, ~heading,  ~icao,        ~id, ~type,       ~name,
+  "N53°25.28'", "W6°16.20'",        242,       NA, "EIDW",     "EIDW", "ARP",  "EIDW ARP",
+  "N53°25.35'", "W6°17.40'",        242,      101, "EIDW", "RWY10/28", "RWY", "Runway 10",
+  "N53°25.22'", "W6°15.03'",        202,      281, "EIDW", "RWY10/28", "RWY", "Runway 28",
+  "N53°26.22'", "W6°15.72'",        217,      162, "EIDW", "RWY16/34", "RWY", "Runway 16",
[…]

### 2019-02-18 R/extract-arrivals.R

@@ -1,0 +1,103 @@
+# you can RUN this from command line (once you set the dates of interest
+# below _AND_ bunzip2 the relevant position files) as foloows:
+#
+# date; Rscript R/extract-arrivals.R; date
+#
+# for me it took:
+# extract_beg <- "Fri Jan  4 14:43:25 CET 2019" %>%
+#   as_datetime(tz = "CET", format = "%a %b %e %T CET %Y") %>%
+#   as_datetime()
+# extract_end <- "Fri Jan  4 15:05:45 CET 2019" %>%
+#   as_datetime(tz = "CET", format = "%a %b %e %T CET %Y") %>%
+#   as_datetime()
+#
+# extract_end - extract_beg
+# Time difference of 22.33333 mins
+
+library(readr)
+library(dplyr)
+library(purrr)
+library(geosphere)
+library(rlang)
+library(stringr)
+library(fs)
+
+source(here::here("R", "utils.R"))
+
+days_of_data <- c(
+  "2017-08-03"
+)
+
+apts <- c(
+  "eidw",
+  "egll")
+
+
+# extract flights arrivals at airport apt on date
+flts_for_apt_on_date <- function(apt, date) {
+  flights_bz2 <- fs::path(
+    here::here("data-raw/"),
+    str_glue("cpr_fr24_flights_{d}", d = date),
+    ext = "csv.bz2")
+  extract_arrival_flights(rawdatafile = flights_bz2, apt = apt)
+}
+
+purrr::pwalk(
+  tidyr::crossing(apts, days_of_data),
+  ~ flts_for_apt_on_date(.x, .y))
+
+
+# extract position reports for flight arrivals at airport apt on date
+poss_for_apt_on_date <- function(apt, date) {
+  flights_apt_csv <- fs::path(
+    here::here("data"),
+    str_glue("{apt}_flt_rt_{d}", apt = apt, d = date),
+    ext = "csv")
+  flt_rt <- readr::read_csv(flights_apt_csv)
+  flt_ids <- flt_rt %>% dplyr::pull(flight_id)
+
+  source(
+    here::here(
+      "R",
+      stringr::str_glue("{apt}-data.R", apt = apt)))
+  apt_data <- get(paste0(apt, "_apt"))
+  arp <- get(paste0(apt, "_arp"))
+  arp <- c(arp$longitude, arp$latitude)
+
+  poss_bz2 <- fs::path(
[…]

### 2019-02-18 R/extract-transition-points.R

@@ -1,0 +1,166 @@
+library(readr)
+library(dplyr)
+library(sf)
+library(stringr)
+
+date_of_interest <- "2017-08-08"
+buffer_miles <- 2
+# # EGLL
+# apt <- "egll"
+# lower_ft <- 6700
+# EIDW
+apt <- "eidw"
+lower_ft <- 2700
+
+
+# get airport relevant details
+source(here::here("R", str_glue("{apt}-data.R", apt = apt)))
+arp <- get(str_glue("{apt}_arp", apt = apt))
+
+# use st_buffer to enlarge the "manually" defined holding polygons
+# see https://stackoverflow.com/a/46706245/963575
+library(units)
+
+# buffer for holding polygon: `buffer_miles` nautical miles
+buf <- set_units(buffer_miles, nmile) %>%
+  set_units(m)
+
+holdings <- get(str_glue("{apt}_holdings_sf", apt = apt)) %>%
+  st_transform(29902) %>%
+  st_buffer(buf) %>%
+  st_transform(4326)
+
+
+# read flight info
+flt_rt_file <- str_glue("data/{apt}_flt_rt_{d}.csv", apt = apt, d = date_of_interest)
+cols <- cols(
+  flight_id = col_character(),
+  callsign = col_character(),
+  aircraft_reg = col_character(),
+  aircraft_type = col_character(),
+  aircraft_address = col_character(),
+  adep = col_character(),
+  ades = col_character(),
+  ssr_codes = col_character(),
+  period_start = col_datetime(format = ""),
+  period_finish = col_datetime(format = ""),
+  source_ids = col_character()
+)
+
+flt_rt <- read_csv(flt_rt_file)
+flt_rt_no_dups <- flt_rt %>%
+  group_by(callsign) %>%
+  mutate(n = n()) %>%
+  ungroup() %>%
+  filter(n == 1 | (n >= 2 & !is.na(aircraft_type)))
+
+# read positions within 40 NM
+cols <- cols(
+  flight_id = col_character(),
+  timestamp = col_datetime(format = ""),
+  latitude = col_double(),
+  longitude = col_double(),
+  altitude = col_double(),
+  speed_gnd = col_double(),
+  track_gnd = col_double(),
+  vert_speed = col_double(),
+  on_ground = col_logical(),
+  distance = col_double(),
+  distance_arp = col_double()
+)
+pos_rt_40NM_file <- str_glue(
[…]

### 2019-02-18 R/holding-definition.R

@@ -1,0 +1,107 @@
+library(sf)
+library(geosphere)
+
+#' calculate turning radius
+#'
+#' @param tas true airspeed (kt)
+#' @param ba  bank angle (degrees) [default = 30].
+#'
+#' @return turning radius (ft)
+#'
+turning_radius <- function(tas, ba = 30) {
+  tas * tas * tan(ba * pi / 180) / 11.26
+}
+
+#' calculate rate of turn
+#'
+#' @param tas true airspeed (kt)
+#' @param ba  bank angle (degrees) [default = 30].
+#'
+#' @return turning radius (ft)
+#'
+turn_rate <- function(tas, ba) {
+  1.091 * tan(ba * pi / 180) / tas
+}
+
+# st_as_sfc("CIRCULARSTRING(2 0, 1 1, 0 0)") %>% st_cast("LINESTRING") %>% st_cast("POINT") %>% plot()
+
+holding_fix <- st_sfc(st_point(cbind(-0.4472222, 51.305)), crs=4326)
+inbound_direction <- 329
+holding_side <- "right"
+tas_max <- 220
+
+P <- st_coordinates(holding_fix) %>%
+  as.vector() %>%
+  setNames(c("lon", "lat"))
+r <- turning_radius(tas_max)
+
+# center of upper circle
+Cu <- geosphere::destPoint(P, inbound_direction + 90, 0.3048 * r) %>%
+  as.vector() %>%
+  setNames(c("lon", "lat"))
+
+# top of upper turn
+T <- geosphere::destPoint(Cu, inbound_direction, 0.3048 * r) %>%
+  as.vector() %>%
+  setNames(c("lon", "lat"))
+
+
+# opposite to holding fix
+Q <- geosphere::destPoint(P, inbound_direction + 90, 0.3048 * 2 * r) %>%
+  as.vector() %>%
+  setNames(c("lon", "lat"))
+
+
+# first straight leg
+R <- geosphere::destPoint(Q, inbound_direction + 180, 0.514444 * tas_max * 60) %>%
+  as.vector() %>%
+  setNames(c("lon", "lat"))
+
+# point before holding fix
+S <- geosphere::destPoint(P, inbound_direction + 180, 0.514444 * tas_max * 60) %>%
+  as.vector() %>%
+  setNames(c("lon", "lat"))
+
+# center of lower circle
+Cl <- geosphere::destPoint(S, inbound_direction + 90, 0.3048 * r) %>%
[…]

### 2019-02-18 R/lszh-data.R

@@ -1,0 +1,53 @@
+
+######### Airport Reference Point
+# Zurich ( from https://skyvector.com/airport/LSZH/Zurich-Airport )
+#
+# Location Information for LSZH
+# Coordinates: N47°27.48' / E8°32.88'
+# View all Airports in Zürich, Switzerland.
+# Elevation is 1417.0 feet MSL.
+# Magnetic Variation is 2° East
+
+########## Runways
+# Runway 16/34
+# Dimensions:	12139 x 197 feet / 3700 x 60 meters
+# Surface:	Hard
+# Runway 16	Runway 34
+# Coordinates:	N47°28.54' / E8°32.16'	N47°26.96' / E8°33.25'
+# Elevation: 	1390 	1388
+# Runway Heading: 	154° 	334°
+# Displaced Threshold: 		1542 Feet
+
+# Runway 14/32
+# Dimensions:	10827 x 197 feet / 3300 x 60 meters
+# Surface:	Unknown
+# Runway 14	Runway 32
+# Coordinates:	N47°28.93' / E8°32.16'	N47°27.68' / E8°33.87'
+# Elevation: 	1402 	1402
+# Runway Heading: 	136° 	316°
+# Displaced Threshold: 	492 Feet
+
+# Runway 10/28
+# Dimensions:	8202 x 197 feet / 2500 x 60 meters
+# Surface:	Unknown
+# Runway 10	Runway 28
+# Coordinates:	N47°27.54' / E8°32.25'	N47°27.40' / E8°34.23'
+# Elevation: 	1391 	1416
+# Runway Heading: 	095° 	275°
+
+
+lszh_apt <- tribble(
+  ~latitude,  ~longitude, ~elevation,  ~icao,          ~id, ~type,        ~name,
+  "N47°27.48'", "E8°32.88'",       1417, "LSZH",       "LSZH", "ARP",   "LSZH ARP",
+  "N47°28.54'", "E8°32.16'",       1390, "LSZH",   "RWY16/34", "RWY", "Runway 16",
+  "N47°26.96'", "E8°33.25'",       1388, "LSZH",   "RWY16/34", "RWY", "Runway 34",
+  "N47°28.93'", "E8°32.16'",       1402, "LSZH",   "RWY14/32", "RWY", "Runway 14",
+  "N47°27.68'", "E8°33.87'",       1402, "LSZH",   "RWY14/32", "RWY", "Runway 32",
[…]

### 2019-02-18 R/plot-egll-tma.R

@@ -1,0 +1,63 @@
+# in order to use it source this file and then invoke the function.
+# you get a ggplot object back
+plot_egll_tma <- function() {
+  # FIXME: super nasty
+  library(ggplot2)
+  library(here)
+  library(readr)
+
+
+  source(here("R", "utils.R"), encoding = 'UTF-8', local = TRUE)
+  source(here("R", "egll-data.R"), encoding = 'UTF-8', local = TRUE)
+
+  stacks <- egll_vor %>%
+    dplyr::mutate(hjust = 0, vjust = 0) %>%
+    dplyr::mutate(
+      hjust = ifelse(id == "BNN",  -0.2, hjust),
+      vjust = ifelse(id == "BNN",  1.5, vjust)) %>%
+    dplyr::mutate(
+      hjust = ifelse(id == "BIG",  1.2, hjust),
+      vjust = ifelse(id == "BIG", -0.7, vjust)) %>%
+    dplyr::mutate(
+      hjust = ifelse(id == "LAM",  1.2, hjust),
+      vjust = ifelse(id == "LAM",  1.5, vjust)) %>%
+    dplyr::mutate(
+      hjust = ifelse(id == "OCL",  -0.2, hjust),
+      vjust = ifelse(id == "OCL", -0.7, vjust))
+
+  gvor <- ggplot2::geom_point(data = egll_vor,
+    ggplot2::aes(x = longitude,
+      y = latitude),
+    colour = "blue",
+    size = 2)
+  gvorl <- ggplot2::geom_text(data = stacks,
+    ggplot2::aes(x = longitude,
+      y = latitude,
+      label = id,
+      hjust = hjust,
+      vjust = vjust))
+
+
+  gbox <- ggplot2::geom_sf(data = egll_holdings_sf, alpha = 0)
+
+
+  g <- ggplot2::ggplot()
+
+  grw_sf <- ggplot2::geom_sf(data = egll_rw_sf, size = 1.2)
+
+  rw <- egll_apt %>%
+    dplyr::filter(
+      stringr::str_detect(name, stringr::fixed("Runway", ignore_case = TRUE))) %>%
+    group_by(id) %>%
+    mutate(latitude2 = lag(latitude), longitude2 = lag(longitude)) %>%
+    ungroup() %>%
+    filter(!is.na(latitude2)) %>%
+    select(-name)
+
+  grw <- ggplot2::geom_segment(
+    data = rw,
[…]

### 2019-02-18 R/plot-eidw-tma.R

@@ -1,0 +1,67 @@
+# in order to use it source this file and then invoke the function.
+# you get a ggplot object back
+plot_eidw_tma <- function() {
+  # FIXME: super nasty
+  library(ggplot2)
+  library(here)
+  library(readr)
+
+  source(here("R", "utils.R"), encoding = 'UTF-8', local = TRUE)
+  source(here("R", "eidw-data.R"), encoding = 'UTF-8', local = TRUE)
+
+  eidw_vor <- eidw_vor %>%
+    dplyr::filter(!id %in% c("BOYNE", "BAGSO", "SORIN"))
+
+
+  stacks <- eidw_vor %>%
+    dplyr::mutate(hjust = 0, vjust = -0.7) %>%
+    # TODO: adapt to EIDW
+    dplyr::mutate(
+      hjust = ifelse(id == "PEKOK",  1.0, hjust),
+      vjust = ifelse(id == "PEKOK",  1.5, vjust)) %>%
+    dplyr::mutate(
+      hjust = ifelse(id == "EPIDU",  1.2, hjust),
+      vjust = ifelse(id == "EPIDU", 0.7, vjust)) %>%
+    dplyr::mutate(
+      hjust = ifelse(id == "KERAV",  1.2, hjust),
+      vjust = ifelse(id == "KERAV",  1.5, vjust)) %>%
+    dplyr::mutate(
+      hjust = ifelse(id == "OCL",  -0.2, hjust),
+      vjust = ifelse(id == "OCL", -0.7, vjust))
+
+  gvor <- ggplot2::geom_point(data = eidw_vor,
+    ggplot2::aes(x = longitude,
+      y = latitude),
+    colour = "blue",
+    size = 2)
+  gvorl <- ggplot2::geom_text(data = stacks,
+    ggplot2::aes(x = longitude,
+      y = latitude,
+      label = id,
+      hjust = hjust,
+      vjust = vjust))
+
+
+  gbox <- ggplot2::geom_sf(data = eidw_holdings_sf, alpha = 0)
+
+
+  g <- ggplot2::ggplot()
+
+  grw_sf <- ggplot2::geom_sf(data = eidw_rw_sf, size = 1.2)
+
+  rw <- eidw_apt %>%
+    dplyr::filter(
+      stringr::str_detect(name, stringr::fixed("Runway", ignore_case = TRUE))) %>%
+    group_by(id) %>%
[…]

### 2019-02-18 R/read-data.R

@@ -1,0 +1,45 @@
+read_trj_40NM <- function(apt, date = "2017-08-01"){
+  cols <- cols(
+    flight_id = col_character(),
+    timestamp = col_datetime(format = ""),
+    latitude = col_double(),
+    longitude = col_double(),
+    altitude = col_double(),
+    speed_gnd = col_double(),
+    track_gnd = col_double(),
+    vert_speed = col_double(),
+    on_ground = col_logical(),
+    distance = col_double(),
+    distance_arp = col_double(),
+    sequence_number = col_integer()
+  )
+
+  files <- stringr::str_glue("data/{apt}_pos_rt_40NM_{d}.csv", d = date)
+  files %>% purrr::map_df(~ readr::read_csv(., col_types = cols))
+}
+
+read_trj_trans <- function(apt, date = "2017-08-01"){
+  cols <- cols(
+    flight_id = col_character(),
+    timestamp = col_datetime(format = ""),
+    longitude = col_double(),
+    latitude = col_double(),
+    altitude = col_double(),
+    distance_flown = col_double(),
+    distance_arp = col_double(),
+    sequence_number = col_integer(),
+    type = col_character(),
+    holding_id = col_character(),
+    bearing = col_double(),
+    quadrant = col_character(),
+    callsign = col_character(),
+    aircraft_reg = col_character(),
+    aircraft_type = col_character(),
+    aircraft_address = col_character(),
+    adep = col_character(),
+    ades = col_character()
+  )
+
+  files <- stringr::str_glue("data/{apt}_transition_points_{d}.csv", d = date)
+  files %>% purrr::map_df(~ readr::read_csv(., col_types = cols))
+}

### 2019-02-18 R/utils.R

@@ -1,0 +1,436 @@
+library(dplyr)
+library(stringr)
+
+# copied from trrrj::ddm2dd
+ddm2dd <- function(value) {
+  # RegEx for DdM (eventually w/ spaces):
+  #   ([N|S|W|E]\s*)?(\d+)°\s*(\d+(?:\.\d+)?)'
+  pattern <- "([N|S|W|E]\\s*)?(\\d+)°\\s*(\\d+(?:\\.\\d+)?)'"
+  pieces <- value %>% stringr::str_subset(pattern) %>% stringr::str_match(pattern)
+  dd <- as.double(pieces[, 3]) + as.double(pieces[, 4]) / 60
+  dd <- ifelse(stringr::str_detect(pieces[, 2], "S|W"), -dd, dd)
+  # constrain longitude to (-180, 180]: ((lon - 180) %% 360) - 180
+  # constrain latitude to (-90, 90]: ((lat - 90) %% 180) - 90
+  dd <- ifelse(
+    stringr::str_detect(pieces[, 2], "W|E"), # if not lon, assume lat
+    ( (dd - 180) %% 360) - 180,
+    ( (dd -  90) %% 180) -  90)
+  dd
+}
+
+# copied from trrrj::dms2dd
+dms2dd <- function(value) {
+  # RegEx for DMS (eventually w/ spaces):
+  #   (\d+)°\s*(\d+)'\s*(\d+(?:\.\d+)?)\"\s*([N|S|W|E]?)
+  pattern <- "(\\d+)°\\s*(\\d+)'\\s*(\\d+(?:\\.\\d+)?)\"\\s*([N|S|W|E]?)"
+  pieces <- value %>% stringr::str_subset(pattern) %>% stringr::str_match(pattern)
+  dd <- ( (as.double(pieces[, 4]) / 60) + as.double(pieces[, 3])) / 60 + as.double(pieces[, 2])
+  dd <- ifelse(stringr::str_detect(pieces[, 5], "S|W"), -dd, dd)
+  # constrain longitude to (-180, 180]: ((lon - 180) %% 360) - 180
+  # constrain latitude to (-90, 90]: ((lat - 90) %% 180) - 90
+  dd <- ifelse(
+    stringr::str_detect(pieces[, 5], "W|E"), # if not lon, assume lat
+    ( (dd - 180) %% 360) - 180,
+    ( (dd - 90) %% 180)  -  90)
+  dd
+}
+
+
+# extract position reports at `distance` from position (lon,lat) arp
[…]

### 2019-02-18 analysis/data.Rmd
data description

@@
 
-Here we describe the
-* raw data, i.e. the Reference Trajectory data set
-* pre-processed data, i.e. the extracted arrivals and the transition points.
+# Data Files
+
+We have two directories for data
+
+1. `data-raw`: it contains the raw source data. For this project these are the Reference Trajectory
+   files as from Google Cloud Platform. See https://github.com/euctrl-pru/reftrj .
+   Due to their big size these are not duplicated/version controlled on this repo.
+1. `data`: here are all the files necessary for the analysis as performed in the paper.
+
+## From `data-raw` to `data`
+
+The raw data filenames are as follows:
+
+* flight information, i.e. `cpr_fr24_flights_2017-08-01.csv.bz2`
+* positions data, i.e. `mas_05_cpr_fr24_synth_positions_2017-08-01.csv.bz2`
+
+PRU's [Open Reference Trajectory dataset](https://github.com/euctrl-pru/reftrj)
+makes available the files needed for the analysis in this project,
+you can retrieve them either by saving from your browser the relevant linked
+flight and position files or use Google `gsutil` as follows:
+
+```{bash ,eval=FALSE}
+# enable relevant Python virtual environment (Python 2) if you have one, and you should.
+$ pyenv activate anaconda2-4.2.0
+$ cd data-raw
+$ gsutil cp \
+    gs://pru-trajectories/sources/merged/overnight_cpr_fr24/raw_cpr_fr24_positions_2017-08-08.csv.bz2 .
+```
+
+As explained in the paper, you can use `R/extract-arrivals.R` to get to CSV files
+containing only the flights and positions for your airport of initerest.
+You can edit the relevant configuration values at the beginning of the script before launching it.
+Yuo will get:
+
+* Flights details for arrivals at `<airport>` on `<date>`
+  are in a CSV file `<airport>_flt_rt_<date>.csv`.
+  For example `egll_flt_rt_2017-08-01.csv` is for arrivals to
+  London Heathrow Airport (ICAO's id EGLL) on Aug, 1 2017.
+  Key data items:
+    * flight_id     unique flight identification
+    * callsign      callsign flight identification of flight
+    * aircraft_reg  aircraft registration
+    * aircraft_type aircraft type, ICAO coding (e.g. A388, B772)
+    * adep / ades   aerodrome of departure / destination
+
+* Position reports for arrival at `<airport>` on `<date>`
+  are in a CSV file `<airport>_pos_rt_<date>.csv`.
+  For example `eidw_pos_rt_2017-08-01.csv` is for arrivals to
+  Dublin Airport (ICAO's id EIDW) on Aug, 1 2017.
+
+The focus of this project is for the terminal area, so we have further filtered the data
+and kept the last 40 NM from the airport reference position (ARP).
+
+The position reports within 40 nautical miles for arrival at
+`<airport>` on `<date>`  are in a CSV file 
+`<airport>_pos_rt_40NM_<date>.csv`.
+These are a subset from `<airport>_pos_rt_<date>.csv`.
+For example `eidw_pos_rt_40NM_2017-08-02.csv` is for arrivals'
+positions within 40 nautical miles to Dublin Airport
+(ICAO's id EIDW) on Aug, 2 2017.
+Key data items:
+  * flight_id     unique flight identification
+  * timestamp     position report time
+  * latitude      latitude coordinate (decimal notation)
+  * longitude     longitude coordinate (decimal notation)
+  * on_ground     on-ground / aloft flag
+  * distance      [not sure ~~] miles flown [?]
+  * distance_arp  distance position to aerodrome reference point
+
+Further we have extracted the transition points from the holding/point merge areas
+using the script `R/extract-transition-points.R` which produces
+the files `<airport>_transition_points_<date>.csv`.
+
+## Version controlled data
+
+We version control only the following data sets in `data`
+
+* `<airport>_flt_rt_<date>.csv`
+* `<airport>_pos_rt_40NM_<date>.csv`
+* `<airport>_transition_points_<date>.csv`

### 2019-02-18 analysis/data.Rmd
fixed few typos and markdown

@@
 PRU's [Open Reference Trajectory dataset](https://github.com/euctrl-pru/reftrj)
-makes available the files needed for the analysis in this project,
-you can retrieve them either by saving from your browser the relevant linked
-flight and position files or use Google `gsutil` as follows:
+makes available the files needed for the analysis in this project;
+you can retrieve them either by saving the relevant linked
+flight and position files from your browser or by using Google `gsutil` as follows:
 
@@
 
-As explained in the paper, you can use `R/extract-arrivals.R` to get to CSV files
+As explained in the paper, you can use `R/extract-arrivals.R` to produce CSV files
 containing only the flights and positions for your airport of initerest.
-You can edit the relevant configuration values at the beginning of the script before launching it.
-Yuo will get:
+You should edit the relevant configuration values at the beginning of the script before launching it.
+You will get:
 
@@
   Key data items:
-    * flight_id     unique flight identification
-    * callsign      callsign flight identification of flight
-    * aircraft_reg  aircraft registration
-    * aircraft_type aircraft type, ICAO coding (e.g. A388, B772)
-    * adep / ades   aerodrome of departure / destination
+    * `flight_id`     unique flight identification
+    * `callsign`      callsign flight identification of flight
+    * `aircraft_reg`  aircraft registration
+    * `aircraft_type` aircraft type, ICAO coding (e.g. A388, B772)
+    * `adep` / `ades`   aerodrome of departure / destination
 
@@
 Key data items:
-  * flight_id     unique flight identification
-  * timestamp     position report time
-  * latitude      latitude coordinate (decimal notation)
-  * longitude     longitude coordinate (decimal notation)
-  * on_ground     on-ground / aloft flag
-  * distance      [not sure ~~] miles flown [?]
-  * distance_arp  distance position to aerodrome reference point
+
+  * `flight_id`     unique flight identification
+  * `timestamp`     position report time
+  * `latitude`      latitude coordinate (decimal notation)
+  * `longitude`     longitude coordinate (decimal notation)
+  * `on_ground`     on-ground / aloft flag
+  * `distance`      flown distance
+  * `distance_arp`  distance position to aerodrome reference point
 
@@
 * `<airport>_transition_points_<date>.csv`
+
+No files are version controlled in `data-raw`.

### 2019-09-26 paper.Rmd
updated base_format according to bookdown book. Removed word_document and commented out parts of YAML

@@
 output:
-  rticles::ieee_article:
-    base_format: bookdown::pdf_book
+  bookdown::pdf_book:
+    base_format: rticles::ieee_article
     fig_caption: no
@@
   html_document: default
-  word_document:
-    toc: no
 bibliography: references.bib
-# biblatex: true
 color-links: yes
@@
 affiliation:
-  # author-columnar: yes
   institution-columnar: true
@@
 papersize: a4paper
-abstract: |
-  Air transportation is undergoing a fundamental transformation. On the political, strategic, and operational level work is directed to meet the future growth of air traffic.
-  To meet this goal, operational concepts and technological enablers are deployed with the promises of distinct performance improvements.
-These improvements are typically widely marketed, however, the underlying evidence is not made available.
-From that respect Air Navigation System Performance suffers from the same credibility crisis seen in other industries.
-  This paper implements the reproducibility paradigm by providing data, code, and results openly to enable another researcher or analyst to reproduce - and potentially - improve or build on our work.
-This work is part of a wider thread to establish Open Data and Open Source based data analytical capabilities for Air Navigation System Performance in Europe.
-  The implementation of the paradigm is demonstrated on the basis of the performance reference trajectory by replicating a classical performance measure, and then moving into investigating the application of trajectory based information for enhancing the current state of performance monitoring for the arrival phase.
-This paper applies the concept and approach as an use case analysis for 2 European airports with different strategies for sequencing arrival traffic (i.e.
-stack holding and point merge).
-  The work reported demonstrates the feasibility of the approach and reflects a stepping stone in the roll-out of the performance reference trajectory within the European context.
-The work also builds the basis for novel developments on the basis of the implemented reproducibility workflow as this paper is accompanied by a webpage that supports the impartial evaluation of the work by providing the data, code, and how this reports had been produced.
-
+abstract: "Air transportation is undergoing a fundamental transformation. On the political,
+  strategic, and operational level work is directed to meet the future growth of air
+  traffic.\nTo meet this goal, operational concepts and technological enablers are
+  deployed with the promises of distinct performance improvements. These improvements
+  are typically widely marketed, however, the underlying evidence is not made available.
+  From that respect Air Navigation System Performance suffers from the same credibility
+  crisis seen in other industries. \nThis paper implements the reproducibility paradigm
+  by providing data, code, and results openly to enable another researcher or analyst
+  to reproduce - and potentially - improve or build on our work. This work is part
+  of a wider thread to establish Open Data and Open Source based data analytical capabilities
+  for Air Navigation System Performance in Europe.\nThe implementation of the paradigm
+  is demonstrated on the basis of the performance reference trajectory by replicating
+  a classical performance measure, and then moving into investigating the application
+  of trajectory based information for enhancing the current state of performance monitoring
+  for the arrival phase. This paper applies the concept and approach as an use case
+  analysis for 2 European airports with different strategies for sequencing arrival
+  traffic (i.e. stack holding and point merge).\nThe work reported demonstrates the
+  feasibility of the approach and reflects a stepping stone in the roll-out of the
+  performance reference trajectory within the European context. The work also builds
+  the basis for novel developments on the basis of the implemented reproducibility
+  workflow as this paper is accompanied by a webpage that supports the impartial evaluation
+  of the work by providing the data, code, and how this reports had been produced.\n"
 title-meta: Busting the Myth! - Reproducibility in ANS Performance

### 2019-09-26 analysis/setup.Rmd
Improved installation instructions for rticles

@@
 * `rticles`: from [`rstudio`](https://github.com/rstudio/rticles) if
-  [PR #207](https://github.com/rstudio/rticles/pull/207) has been merged,
-  otherwise from [`espinielli`'s branch](https://github.com/espinielli/rticles/tree/add-support-keywords)
+  [PR #247](https://github.com/rstudio/rticles/pull/247) has been merged
+  ```{r, eval=FALSE}
+  install.packages("rticles")
+  ```
+  otherwise directly from [PR #247](https://github.com/rstudio/rticles/pull/247)
+  ```{r, eval=FALSE}
+  remotes::install_github("rstudio/rticles#247")
+  ```
 
@@
 
-  * References on second or more line not correctly indented. (maybe re-try BibLaTeX)
+  * References on second and subsequent lines not correctly indented.

### 2019-09-26 analysis/setup.Rmd
Update due to merge of PR in rticles

@@
 * `rticles`: from [`rstudio`](https://github.com/rstudio/rticles) if
-  [PR #247](https://github.com/rstudio/rticles/pull/247) has been merged
+  [PR #247](https://github.com/rstudio/rticles/pull/247) has been merged and released on CRAN
   ```{r, eval=FALSE}
@@
   ```
-  otherwise directly from [PR #247](https://github.com/rstudio/rticles/pull/247)
+  otherwise directly from [Github](https://github.com/rstudio/rticles/)
   ```{r, eval=FALSE}
-  remotes::install_github("rstudio/rticles#247")
+  remotes::install_github("rstudio/rticles")
   ```

### 2019-09-26 analysis/data.Rmd
correct data filename on google

@@
 $ gsutil cp \
-    gs://pru-trajectories/sources/merged/overnight_cpr_fr24/raw_cpr_fr24_positions_2017-08-08.csv.bz2 .
+    gs://pru-trajectories/products/synth_positions/cpr_fr24/mas_05_cpr_fr24_synth_positions_2017-08-03.csv.bz2 .
 ```

### 2019-09-26 analysis/website.Rmd
Added info on paper and website production

---
title: "Website"
output:
  workflowr::wflow_html:
    toc: false
---

## Build the companion website

The companion website to this project is generated by the `workflowr` package as follows:

```{r website-build,eval=FALSE}
wflow_build()
```

This will 
1. generate the relevant HTML pages in `docs` out of the R markdown files in the `analysis`
   directory.
1. preview the website in the Viewer panel in RStudio

The `docs` directory is automatically served by GitHub, so the website will be browsable on
the Internet at https://euctrl-pru.github.io/reproducible-ans-performance-paper/.

## Publish the website

Publishing and versioning the project website us accomplished via `wflow_publish()`.
This will build and version control the HTML pages.

### 2019-10-03 R/egll-data.R
improved data extraction scripts

@@ -71,12 +71,12 @@ library(sf)
 # 000° 26' 50.00" W
 
 egll_apt <- tibble::tribble(
-     ~latitude,  ~longitude, ~elevation,  ~icao,          ~id, ~type,        ~name,
-  "N51°28.65'", "W0°27.68'",       83.0, "EGLL",       "EGLL", "ARP",   "EGLL ARP",
-  "N51°28.65'", "W0°29.10'",         79, "EGLL", "RWY09L/27R", "RWY", "Runway 09L",
-  "N51°28.66'", "W0°26.00'",         78, "EGLL", "RWY09L/27R", "RWY", "Runway 27R",
-  "N51°27.89'", "W0°28.94'",         75, "EGLL", "RWY09R/27L", "RWY", "Runway 09R",
-  "N51°27.90'", "W0°26.04'",         77, "EGLL", "RWY09R/27L", "RWY", "Runway 27L"
+     ~latitude,  ~longitude, ~elevation, ~heading,  ~icao,          ~id, ~type,        ~name,
+  "N51°28.65'", "W0°27.68'",       83.0,       NA, "EGLL",       "EGLL", "ARP",   "EGLL ARP",
+  "N51°28.65'", "W0°29.10'",         79,       91, "EGLL", "RWY09L/27R", "RWY", "Runway 09L",
+  "N51°28.66'", "W0°26.00'",         78,      271, "EGLL", "RWY09L/27R", "RWY", "Runway 27R",
+  "N51°27.89'", "W0°28.94'",         75,       91, "EGLL", "RWY09R/27L", "RWY", "Runway 09R",
+  "N51°27.90'", "W0°26.04'",         77,      271, "EGLL", "RWY09R/27L", "RWY", "Runway 27L"
 ) %>%
   dplyr::mutate(latitude = ddm2dd(latitude), longitude = ddm2dd(longitude))
 

### 2019-10-03 R/egll-holdings.R

@@ -7,7 +7,7 @@ library(stringr)
 date_of_interest <- "2017-08-01"
 
 # get airport relevant details
-source(here::here("R", "egll-data.R"))
+source(here::here("R", "egll-data.R"), encoding = 'UTF-8', local = TRUE)
 arp <- get("egll_arp")
 
 # use st_buffer to enlarge the "manually" defined holding polygons

### 2019-10-03 R/eidw-data.R

@@ -4,7 +4,7 @@ library(sf)
 
 # do not use trrrj yet, just copied needed functions in utils.R ...
 # library(trrrj)
-#source(here::here("R", "utils.R"))
+#source(here::here("R", "utils.R"), encoding = 'UTF-8', local = TRUE)
 
 ######### Airport Reference Point
 # Dublin (from https://skyvector.com/airport/EIDW/Dublin-Airport )

### 2019-10-03 R/extract-arrivals.R

@@ -23,19 +23,28 @@ library(stringr)
 library(fs)
 library(vroom)
 
-source(here::here("R", "utils.R"))
+source(here::here("R", "utils.R"), encoding = 'UTF-8', local = TRUE)
 
 days_of_data <- c(
+  "2017-08-01",
+  "2017-08-02",
+  "2017-08-03",
+  "2017-08-04",
+  "2017-08-05",
   "2017-08-06",
   "2017-08-07",
   "2017-08-08"
 )
 
 apts <- c(
-  # "eidw",
-  "egll")
+  "lszh",
+  "eidw",
+  "egll"
+)
 
+################ EXTRACT ARRIVALS #################
 
+######### Extract Flights #######
 # extract flights arrivals at airport apt on date
 flts_for_apt_on_date <- function(apt, date) {
   flights_bz2 <- fs::path(
@@ -49,7 +58,7 @@ purrr::pwalk(
   tidyr::crossing(apts, days_of_data),
   ~ flts_for_apt_on_date(.x, .y))
 
-
+############ Extract positions ###########
 # extract position reports for flight arrivals at airport apt on date
 poss_for_apt_on_date <- function(apt, date) {
   flights_apt_csv <- fs::path(
@@ -62,7 +71,9 @@ poss_for_apt_on_date <- function(apt, date) {
   source(
     here::here(
       "R",
-      stringr::str_glue("{apt}-data.R", apt = apt)))
+      stringr::str_glue("{apt}-data.R", apt = apt)),
+    encoding = 'UTF-8',
+    local = TRUE)
   apt_data <- get(paste0(apt, "_apt"))
   arp <- get(paste0(apt, "_arp"))
   arp <- c(arp$longitude, arp$latitude)
@@ -73,9 +84,6 @@ poss_for_apt_on_date <- function(apt, date) {
       "mas_05_cpr_fr24_synth_positions_{d}",
       d = date),
     ext = "csv.bz2")
-
-  # bunzip2
-  #R.utils::bunzip2(bz2file, csvfile, remove = FALSE, skip = TRUE)
 
   poss_csv <- fs::path_ext_remove(poss_bz2)
 
@@ -91,15 +99,26 @@ purrr::pwalk(
   ~ poss_for_apt_on_date(.x, .y))
 
 
-######################### SMOOTH #####################
+#################### SMOOTH #####################
 purrr::walk(
[…]

### 2019-10-03 R/extract-transition-points.R

@@ -2,165 +2,185 @@ library(readr)
 library(dplyr)
 library(sf)
 library(stringr)
-
-date_of_interest <- "2017-08-08"
-buffer_miles <- 2
-# # EGLL
-# apt <- "egll"
-# lower_ft <- 6700
-# EIDW
-apt <- "eidw"
-lower_ft <- 2700
-
-
-# get airport relevant details
-source(here::here("R", str_glue("{apt}-data.R", apt = apt)))
-arp <- get(str_glue("{apt}_arp", apt = apt))
-
-# use st_buffer to enlarge the "manually" defined holding polygons
-# see https://stackoverflow.com/a/46706245/963575
+library(trrrj)
 library(units)
 
-# buffer for holding polygon: `buffer_miles` nautical miles
-buf <- set_units(buffer_miles, nmile) %>%
-  set_units(m)
-
-holdings <- get(str_glue("{apt}_holdings_sf", apt = apt)) %>%
-  st_transform(29902) %>%
-  st_buffer(buf) %>%
-  st_transform(4326)
-
-
-# read flight info
-flt_rt_file <- str_glue("data/{apt}_flt_rt_{d}.csv", apt = apt, d = date_of_interest)
-cols <- cols(
-  flight_id = col_character(),
-  callsign = col_character(),
-  aircraft_reg = col_character(),
-  aircraft_type = col_character(),
-  aircraft_address = col_character(),
-  adep = col_character(),
-  ades = col_character(),
-  ssr_codes = col_character(),
-  period_start = col_datetime(format = ""),
-  period_finish = col_datetime(format = ""),
-  source_ids = col_character()
+days_of_data <- c(
+  "2017-08-01",
+  "2017-08-02",
+  "2017-08-03",
+  "2017-08-04",
+  "2017-08-05",
+  "2017-08-06",
+  "2017-08-07",
+  "2017-08-08"
 )
 
-flt_rt <- read_csv(flt_rt_file)
-flt_rt_no_dups <- flt_rt %>%
-  group_by(callsign) %>%
-  mutate(n = n()) %>%
-  ungroup() %>%
-  filter(n == 1 | (n >= 2 & !is.na(aircraft_type)))
-
-# read positions within 40 NM
-cols <- cols(
-  flight_id = col_character(),
-  timestamp = col_datetime(format = ""),
-  latitude = col_double(),
-  longitude = col_double(),
-  altitude = col_double(),
[…]

### 2019-10-03 R/lszh-data.R

@@ -1,3 +1,6 @@
+library(tibble)
+library(dplyr)
+library(sf)
 
 ######### Airport Reference Point
 # Zurich ( from https://skyvector.com/airport/LSZH/Zurich-Airport )
@@ -37,14 +40,14 @@
 
 
 lszh_apt <- tribble(
-  ~latitude,  ~longitude, ~elevation,  ~icao,          ~id, ~type,        ~name,
-  "N47°27.48'", "E8°32.88'",       1417, "LSZH",       "LSZH", "ARP",   "LSZH ARP",
-  "N47°28.54'", "E8°32.16'",       1390, "LSZH",   "RWY16/34", "RWY", "Runway 16",
-  "N47°26.96'", "E8°33.25'",       1388, "LSZH",   "RWY16/34", "RWY", "Runway 34",
-  "N47°28.93'", "E8°32.16'",       1402, "LSZH",   "RWY14/32", "RWY", "Runway 14",
-  "N47°27.68'", "E8°33.87'",       1402, "LSZH",   "RWY14/32", "RWY", "Runway 32",
-  "N47°27.54'", "E8°32.25'",       1391, "LSZH",   "RWY10/28", "RWY", "Runway 10",
-  "N47°27.40'", "E8°34.23'",       1416, "LSZH",   "RWY10/38", "RWY", "Runway 28"
+  ~latitude,     ~longitude, ~elevation, ~heading,  ~icao,          ~id, ~type,        ~name,
+  "N47°27.48'", "E8°32.88'",       1417,       NA, "LSZH",       "LSZH", "ARP",   "LSZH ARP",
+  "N47°28.54'", "E8°32.16'",       1390,      154, "LSZH",   "RWY16/34", "RWY", "Runway 16",
+  "N47°26.96'", "E8°33.25'",       1388,      334, "LSZH",   "RWY16/34", "RWY", "Runway 34",
+  "N47°28.93'", "E8°32.16'",       1402,      136, "LSZH",   "RWY14/32", "RWY", "Runway 14",
+  "N47°27.68'", "E8°33.87'",       1402,      316, "LSZH",   "RWY14/32", "RWY", "Runway 32",
+  "N47°27.54'", "E8°32.25'",       1391,       95, "LSZH",   "RWY10/28", "RWY", "Runway 10",
+  "N47°27.40'", "E8°34.23'",       1416,      275, "LSZH",   "RWY10/38", "RWY", "Runway 28"
[…]

### 2019-10-03 R/utils.R

@@ -63,9 +63,23 @@ extract_airport_distance <- function(rawdatafile, arp, distance) {
     fs::path_ext_set("csv")
   file <- fs::path_join(c(here::here(), "data", file))
 
+  cols <- cols(
+    flight_id = col_character(),
+    timestamp = col_datetime(format = ""),
+    latitude = col_double(),
+    longitude = col_double(),
+    altitude = col_double(),
+    speed_gnd = col_double(),
+    track_gnd = col_double(),
+    vert_speed = col_double(),
+    on_ground = col_logical(),
+    distance = col_double(),
+    distance_arp = col_double(),
+    sequence_number = col_double()
+  )
   d <- distance * 1852
   rawdatafile %>%
-    readr::read_csv() %>%
+    readr::read_csv(col_types = cols) %>%
     dplyr::filter(distance_arp <= d) %>%
     readr::write_csv(file)
 }
@@ -99,7 +113,7 @@ extract_arrival_flights <- function(rawdatafile, apt, prefix = "flt_rt") {
 
 #' extract positions for airport
 #'
-#' @param rawdatafile reference trajectory file
+#' @param rawdatafile reference trajectory bzipped file
 #' @param apt airport, i.e. "egll"
 #' @param arp airport reference point (lon/lat), i.e. `c(-0.461389, 51.4775)`
 #' @param ids flight ids to keep
@@ -107,11 +121,14 @@ extract_arrival_flights <- function(rawdatafile, apt, prefix = "flt_rt") {
 #'              The default will result in filename
 #'              egll_pos_rt_<date>.csv where <date> is extracted from the date portion
 #'              in the `rawdatafile` name
+#' @param postfix string for filename, default "raw".
 #' @return a data frame
 #'
 extract_arrival_positions <- function(rawdatafile, apt, arp, ids, prefix = "pos_rt", postfix = "raw") {
-  if(!fs::file_exists(rawdatafile)) {
-    stop(paste0("Error: file", rawdatafile, " does not exist."))
+  pp <- paste0(rawdatafile, ".bz2")
+
+  if(!fs::file_exists(pp)) {
+    stop(paste0("Error: file", pp, " does not exist."))
   }
   data_date <- rawdatafile %>%
     fs::path_file() %>%
@@ -121,10 +138,10 @@ extract_arrival_positions <- function(rawdatafile, apt, arp, ids, prefix = "pos_
[…]

### 2019-10-04 paper.Rmd
almost final for SID2019

@@
       keep_tex: yes
-      after_body: biographies.tex
     number_sections: yes
@@
   are typically widely marketed, however, the underlying evidence is not made available.
-  From that respect Air Navigation System Performance suffers from the same credibility
-  crisis seen in other industries. \nThis paper implements the reproducibility paradigm
+  From that respect Air Navigation System Performance faces varying levels of transparency
+  as seen in other industries. \nThis paper implements the reproducibility paradigm
   by providing data, code, and results openly to enable another researcher or analyst
@@
 
-Accordingly, ANS performance improvements are subject to a "credibility crisis" similar to reported problems in other domains, e.g.
+Accordingly, ANS performance improvements are subject to a "credibility crisis" (varying levels of transparency) similar to reported problems in other domains, e.g.
 medicine [@kolata_2011], psychology [@ritchie_2012], ecology [@fraser_etal_2018].
@@
 Independent (operational) ANS performance builds on transparent and robust performance monitoring and reporting.
-Within the European context EUROCONTROL's Performance Review Unit (PRU) is committed to provide impartial analyses of the performance of the European Air Navigation System based on verified data.
-With a view to enhance transparency, PRU is - inter alia - making performance-related data publicly available via its Portal, [http://ansperformance.eu](http://ansperformance.eu), since early 2017 combined with the methodological documentation for the calculation of the performance indicators used under the EUROCONTROL Performance Revew System and the European Union Performance Scheme.
+Within the European context, EUROCONTROL's Performance Review Unit (PRU) is committed to provide impartial analyses of the performance of the European Air Navigation System based on verified data.
+With a view to enhance transparency, PRU is - inter alia - making performance-related data publicly available via its portal, [http://ansperformance.eu](http://ansperformance.eu), since early 2017 combined with the methodological documentation for the calculation of the performance indicators used under the EUROCONTROL Performance Revew System and the European Union Performance Scheme.
 There has been recognition of using common and openly available tools and data in ATM research and operational validation.
@@
 One of the associated building blocks is the performance reference trajectory [@spinielli_2018].
-The performance reference trajectory will be deployed as Open Data throughout 2019.
+At the time of writing, the performance reference trajectory is in its design stage.
 The trajectory project builds on crowd sourced surveillance data fused with correlated position reports collected from European air navigation service providers.
 Based on the Open Data model it will be possible to augment the data processing (e.g.
-data coverage, time synchronisation), establishment of derived data bases (e,g fleet data base), associated performance models (e.g.
-aircraft performance model), but also the continual further development and validation of operational performance measures/indicators.
+data coverage, time synchronisation), establish derived data bases (e,g fleet data base), and derive associated performance models (e.g.
+aircraft performance model).
+This setup enables the continual further development and validation of operational performance measures/indicators.
 
@@
 # Background - Reproducibility
+
+## Scope and Definition
 
@@
 In our domain (i.e.
-operational ANS Performance and associated data analysis) this will require the access to data and derived summaries, how associated visualisations have been generated, the accompanying explanatory text, and conclusions drawn for the analysis.
+operational ANS Performance and associated data analysis) this will require the access to data and derived summaries, how associated visualisations have been generated, the accompanying explanatory text, and conclusions drawn from the analysis.
 
@@
 This can be contrasted by emerging practices like data journalism.
-One interesting example is buzzFeed's article on the use of surveillance planes operated by law enforcement and military for a variety of purposes.
+One interesting example is buzzFeed's article on the use of surveillance planes operated by law enforcement and military.
 The authors were able to demonstrate based on Open Data that these purposes range from tracking drug trafficking to testing new surveillance technology [@peteraldhous_2016].
-The credibility of the article comes from the fact that all artifacts are made available to reproduce the findings or study the chosen approach and data transformations made leading to the conclusions of the article.
+The credibility of the article comes from the fact that all artifacts are made available to reproduce the findings or study the approach and data transformations leading to the conclusions of the article.
 
@@
 Though the license restrictions did not allow for the publication of the underlying aircraft position data, the authors made available the analytical data set and documented their analysis from the data preparation through to the publication stage.
-
-The reproducibility discussion focuses strongly on the recreation of results by other researchers/analysts.
-A related problem is the re-generating of already completed (and potentially published) results with the - previously - available same input data. 
-This is a fundamental issue that can be summarised as "can a future me re-do what I had done in the past"?
-This problem space ranges across the complete data analysis life-cyle, including archiving the orignially used input data or the steps to generate the study data, documenting the data cleaning and transformation steps including the parameters chosen, etc.
-By definition such a capability is a precondition for reproduction or replication by others. 
-This implies that the analytical steps including the steps to produce summaries or estimates, visualisations, and associated conclusions are documented and archived. 
-In order to enable others to recreate the own findings, such steps and documentation would need to be shareable.
-This consideration has a strong bearing on the toolset chosen for this paper.
-
-<!-- check if this can be merged with above -->
-Some of the reasons why researchers/analysts do not publish data and computational procedures that allow reproduction will be hard to change [@daniel_nust_2018_2361304].
-The predominant technical reasons have been listed above.
-However, there are also personal reasons such as the fear for floating ideas or loss of competitive advantage.
-Other reasons deal with more technical problems, e.g.
-lack of capabilities, procedures.
-
-Another incentive for the adoption of an open and reproducible research practice can be seen in the wider recognition through increased citations, media attention, discussion with the community and collaborators, and ultimately funding opportunities.
-The recent years have seen a strong use of research blogs and community platforms supporting the publication of smaller steps or intermediate results.
-The combination of web presentation and discussions offers a good basis for an initial dialogue.
-Platforms like google scholar or researchgate.net offer additional features for outreach and communication.
-This equally applies for paper achives such as arxiv.org for pre-prints.
-Based on the credibility crisis discussion a number of journals are now promoting - or even requiring - the provision of analytical data in support of the paper publication. 
-There is also an increasing number of requirements for national or European funded research with respect to data management and Open Data access to the research.
 
@@
 
+## Practical Challanges
+
+The reproducibility discussion focuses strongly on the recreation of results by other researchers/analysts.
+A related problem is the re-generating of already completed (and potentially published) results with the - previously - available same input data. 
+This is a fundamental issue that can be summarised as "can a future me re-do what I had done in the past"?
+This problem space ranges across the complete data analysis life-cyle, including archiving the orignially used input data or the steps to generate the study data, documenting the data cleaning and transformation steps including the parameters chosen, etc.
+By definition such a capability is a precondition for reproduction or replication by others. 
+This implies that the analytical steps including the steps to produce summaries or estimates, visualisations, and associated conclusions are documented and archived. 
+In order to enable others to recreate the own findings, such steps and documentation would need to be shareable.
+This consideration has a strong bearing on the toolset chosen for this paper.
+
+<!-- check if this can be merged with above -->
+Some of the reasons why researchers/analysts do not publish data and computational procedures that allow reproduction will be hard to change [@daniel_nust_2018_2361304].
+The predominant technical reasons have been listed above.
+However, there are also personal reasons such as the fear for floating ideas or loss of competitive advantage.
+Other reasons deal with more technical problems, e.g.
+lack of capabilities, procedures.
+
+Another incentive for the adoption of an open and reproducible research practice can be seen in the wider recognition through increased citations, media attention, discussion with the community and collaborators, and ultimately funding opportunities.
+The recent years have seen a strong use of research blogs and community platforms supporting the publication of smaller steps or intermediate results.
+The combination of web presentation and discussions offers a good basis for an initial dialogue.
+Platforms like google scholar or researchgate.net offer additional features for outreach and communication.
+This equally applies for paper archives such as arxiv.org for pre-prints.
+Based on the credibility crisis discussion a number of journals are now promoting - or even requiring - the provision of analytical data in support of the paper publication. 
+There is also an increasing number of requirements for national or European funded research with respect to data management and Open Data access to the research.
+
 (ref:reproducible-research) Reproducible research, Fig. 2 in [@marwick_2017].
@@
 
+## Reproducibility Spectrum
+
 Across the literature, researchers have identified a "reproducibility spectrum" ranging from the extreme, i.e.
-publication only ( advertising and marketing of the done work), to almost or even fully replicable research.
+publication only (advertising and marketing of the done work), to almost or even fully replicable research.
 The work of Marwick et al.
-(2017) [@marwick_2017] and Peng (2011) [@Peng1226] uses the analogy of a spectrum that can be mapped against the scoring of Vandewalle (2009) describing the level of reproducibility on a 0 to 5 scale [@marwick_2017].
+[@marwick_2017] and Peng [@Peng1226] uses the analogy of a spectrum that can be mapped against the scoring of Vandewalle (2009) describing the level of reproducibility on a 0 to 5 scale [@marwick_2017].
 In light of [@leek_2015], this paper conceptualises reproducibility as the "ability to recompute data analytic results given an observed dataset and knowledge of the data analysis pipeline" and accordingly "replicability of a study is the chance that an independent experiment targeting the same scientific question will produce a consistent result".
@@
 http://openaip.net, http://ourairports.com).
-These data flows are processed and stored in the EUROCONTROL data Warehouse.
+These data flows are processed and stored in the EUROCONTROL data warehouse.
 The approach recognises that the full source data may or cannot be provided for a variety of reasons. 
@@
 Being open source, the R community is actively engaging and sharing related software packages to enhance the core functionality.
-Without limiting the impact of other packages, the development of knitr [@R-knitr] and RMarkdown [@xie2018], ggplot [@wickham_2016] for visualisation, and the so-called tidyverse packages and RStudio IDE [@rstudioteam_2015] provide the basis for an open source data analysis ecosystem.
+Without limiting the impact of other packages, the development of knitr [@R-knitr] and RMarkdown [@xie2018], ggplot [@wickham_2016] for visualisation, and the so-called tidyverse packages and RStudio IDE [@rstudioteam_2015] represent an open source ecosystem for data analysis.
 
@@
 
-A minimal example of RMarkdown document, a plain-text file with the conventional
+A minimal example of an RMarkdown document, a plain-text file with the conventional
 extension `.Rmd` is shown below:
@@
 
-(ref:rmd-example) The output of a minimalist example of RMarkdown.
-
-```{r rmd-example,echo=FALSE,fig.cap='(ref:rmd-example)',fig.pos='htb',fig.height=4}
-knitr::include_graphics('figures/rmd-example.png', dpi = NA)
-```
-
 There are three components of an R Markdown document: the metadata, the text and the code.
@@
 output document (title, type of document [web page, PDF, MS Word], ...)
-The text, which follows the metadata, is witten using Markdown, a simplified formatting syntax.
+The text, which follows the metadata, is witten using Markdown, a simplified formatting syntax
+which can include prose and code.
 There are two types of (computer) code:
@@
 Code portions are executed when the document is rendered.
-
-The result of rendering the R Markdown above is shown in Fig. \@ref(fig:rmd-example).
-
-A particular strong point is that the R/Rstudio ecosystem is open source software.
+Fig. \@ref(fig:rmd-example) shows how the simple R Markdown example is rendered.
+
+(ref:rmd-example) The output of a minimalist example of RMarkdown.
+
+```{r rmd-example,echo=FALSE,fig.cap='(ref:rmd-example)',fig.pos='htb',fig.height=3}
+knitr::include_graphics('figures/rmd-example.png', dpi = NA)
+```
+
+A particular strong point is that the R/RStudio ecosystem is open source software.
 This includes also its extensibility with other packages.
@@
 This work has been reported in [@koelle_open_2017] and [@spinielli_2018]. The idea is to establish a pan-European air situation picture openly available to research and practitioners.
-Conceptually, the fused performance reference trajectory will be a key milestone in combatting marketing myths.
-For example, operational improvements can be directly assessed by interested analysts with a view to validate the observable performance improvements.
+Conceptually, the fused performance reference trajectory will be a key milestone in combatting the "marketing myths."
+For example, operational improvements can be directly assessed by interested analysts as access to the Open Data enables the validation of the observable performance improvements.
 This allows for the application of commonly accepted methods and models to express performance benefits and  - ultimately - challenge related publications on deployments and achieved improvements.
 
-End of 2018 the development project was closed and PRU is now investigating options for deploying and maintaining the processing modules on a cloud platform.
-The goal is to establish and operate the underlying data processing capabilities throughout 2019 and launch the regular provision of the performance reference trajectory as Open Data.
+At the end of 2018 the initial development project phase was closed and PRU is now investigating options for deploying and maintaining the processing modules on a cloud platform.
+The goal is to establish and operate the underlying data processing capabilities throughout 2019 and launch the regular provision of the performance reference trajectory as Open Data in 2020.
 It is also planned to establish a network and complement the air situation picture by deploying additional sensors or determining non-ADSB derived positions as a community effort.
@@
     * a file with flight details
-    * a file with all the arrivals' positions
-    * a file with all arrivals' positions within 40 NM the airport's reference position (ARP)
+    * a file with all the arrivals' positions (`raw` trajectories)
+    * a file with all the arrivals' positions (smoothed trajectories)
+    * a file with all (smoothed) arrivals' positions within 40 NM of the airport's reference position (ARP)
 
@@
   as ARP, runway name and threshold position/elevation, polygons for holding
-  stacks or point-merge areas. Specifically the polygons have been heuristically
-  defined and enlarged by 2NM in order to cater for the spread of real flown trajectories,
+  stacks or point-merge areas. Specifically the polygons have been heuristically defined from the
+  point merge points in the AIP and enlarged by 2NM in order to cater for the spread of real flown
+  trajectories,
   see the scripts `R/egll-data.R` and `R/eidw-data.R`.
 
-Each position mentioned above consist of at least
-
-* flight ID
+Each transition position mentioned above consist of at least
+
+* aircraft/flight info (ID, callsign, registration, ADEP/ADES, ICAO 24 bit address)
 * 4D position (timestamp $T$, longitude, latitude, altitude)
-* flown distance $D$
+* flown distance $D$ from aerodrome of departure (ADEP)
 * sequence number
+* point info (sequence number, type of transition point [`P40`, `PHOLD`, `PLAND`])
 * distance from ARP
@@
 
-(ref:coverage-egll) Number of arrivals in data sets, i.e. 1.-8. July 2017.
+(ref:coverage-egll) Number of arrivals in data sets, i.e. 1.-8. Aug 2017.
 
@@
 The ASMA metric calculates the additional time based on a reference time which is determined for a family of flights sharing similar characteristics. 
-These are the entry sector into the arrival airspace, weight turbulence category and engine type, and the actual landing runway.
 For each flight $i$ the additional ASMA time is
@@
 
-(ref:ASMA-travel-egll) Distribution of actual ASMA travel times for EGLL (1.-8. July 2017).
+(ref:ASMA-travel-egll) Distribution of actual ASMA travel times for EGLL (1.-8. Aug 2017).
 
@@
 ```
+
 As an example, the distribution of the actual arrival transit times for arrivals into Heathrow (EGLL) are shown in Fig.
@@
 
-(ref:ASMA-compare) Comparison of determined ASMA values, i.e. 1.-8. July 2017.
+(ref:ASMA-compare) Comparison of determined ASMA values, i.e. 1.-8. Aug 2017.
 
@@
 
-Fig. \@ref(fig:ASMA-compare) shows the variation of the timebased additional ASMA times determined for our sample.
+In Fig. \@ref(fig:ASMA-compare) 
 APDFNM refers to the variant used for the EUROCONTROL Performance Review System.
-The reference time is build on the basis of a one-year sammple.
+The reference time is build on the basis of a one-year sample.
 Applying this reference time to the trajectory based travel times (i.e.
-TRJ-REF), there is a strong increase of the additional ASMA time for Heathrow (EGLL), while in the case of Dublin (EIDW) reasonable decrease takes place.
+TRJ-REF), there is a strong increase of the additional ASMA time for Heathrow (EGLL), while in the case of Dublin (EIDW) a reasonable decrease takes place.
 Calculating the additional ASMA time exclusively with the trajectory sample and applying a 20th percentile approach (TRJ-20PCT), shows a significant drop in EGLL and a moderate increase at Dublin.
 The high variation for Heathrow shows that the additional ASMA time is strongly dependent on the time horizon.
-The pure trajectory bases measure - by definition - includes the effect of the operational concept (stack holdings) and distorts the measure.
+The pure trajectory based measure - by definition - includes the effect of the operational concept (stack holdings) and distorts the measure.
 In the case of Dublin it can be argued that the TRJ-20PCT measure reflects the time based ASMA measure suffiently.
@@
 
-(ref:ASMA-time-dist) Relationship travel time and distance flown, i.e. 1.-8. July 2017.
+(ref:ASMA-time-dist) Relationship travel time and distance flown, i.e. 1.-8. Aug 2017.
 
@@
 The variation of travel times is influenced by varying airspeeds or weather phenomena (e.g. wind).
-This paper applies a simple 15th percentile approach to the distance flown per arrival sector, aircraft type/engine, and landing runway.
+This paper applies a simple 15-th percentile approach to the distance flown per arrival sector, aircraft type/engine, and landing runway.
 
@@
 Tab. \@ref(tab:tbl-asma-dist) lists the determined average additional distance at the study airports.
+The total additional distance is determined as the aggregated sum of the differences in each segmnent
+with 15-th percentile reference distance.
 
@@
 
-(ref:egll-trajectory-example) EGLL - long trajectory flown on 1. July 2017.
+(ref:egll-trajectory-example) EGLL - long trajectory flown on 1. Aug 2017.
 
@@
 
-(ref:eidw-trajectory-example) EIDW - long trajectory flown on 1. July 2017.
+(ref:eidw-trajectory-example) EIDW - long trajectory flown on 1. Aug 2017.
 
@@
 
-Fig. \@ref(fig:egll-trajectory-example) and Fig. \@ref(fig:eidw-trajectory-example) show examples for long distances flown on 1. July 2017.
+Fig. \@ref(fig:egll-trajectory-example) and Fig. \@ref(fig:eidw-trajectory-example) show examples
+for long distances flown on 1. Aug 2017.
 The trajectory data allows for the identification of the holdings and/or sequencing area.
@@
 
-(ref:distance-flown-segment) Actual distances flown for each segment of the arrival phase (i.e. EGLL, 1.-8. July 2017).
+(ref:distance-flown-segment) Actual distances flown for each segment of the arrival phase (i.e. EGLL, 1.-8. Aug 2017).
 
@@
 
-(ref:distance-flown-segment-eidw) Actual distances flown for each segment of the arrival phase (i.e. EIDW, 1.-8. July 2017).
+(ref:distance-flown-segment-eidw) Actual distances flown for each segment of the arrival phase (i.e. EIDW, 1.-8. Aug 2017).
 
@@
 
-(ref:eidw-saturation-example) EIDW - arrival throughput saturation (1.-8. July 2017).
+(ref:eidw-saturation-example) EIDW - arrival throughput saturation (1.-8. Aug 2017).
 

### 2019-10-04 paper.Rmd
final code version

@@
 ---
-title: "Busting the Myth! - Reproducibility in ANS Performance"
+title: Towards Reproducibility in ANS Performance
 author:
@@
   of trajectory based information for enhancing the current state of performance monitoring
-  for the arrival phase. This paper applies the concept and approach as an use case
+  for the arrival phase. This paper applies the concept and approach as a use case
   analysis for 2 European airports with different strategies for sequencing arrival
-  traffic (i.e. stack holding and point merge).\nThe work reported demonstrates the
-  feasibility of the approach and reflects a stepping stone in the roll-out of the
-  performance reference trajectory within the European context. The work also builds
-  the basis for novel developments on the basis of the implemented reproducibility
-  workflow as this paper is accompanied by a webpage that supports the impartial evaluation
-  of the work by providing the data, code, and how this reports had been produced.\n"
-title-meta: Busting the Myth! - Reproducibility in ANS Performance
+  traffic (i.e. stack holding and point merge).\nThe research reported demonstrates the
+  feasibility of the reproducibility approach. This work introduces a novel workflow by making available a
+  companion web repository that provides data, code, and instructions to fully reproduce the
+  paper.\n"
+title-meta: Towards Reproducibility in ANS Performance
 ---
@@
 
-Air transportation is undergoing a fundamental transformation.
 To meet the future growth of air traffic, political decision makers, strategic planers, practitioners and researchers aim to enhance the performance of air navigation services (ANS).
-At the time being there is no consistent and regionally harmonised deployment programme.
-Typically, the deployment programmes are revolving around public-private partnerships and regional funding programmes resulting in a wide mix of consortia and stakeholder subsets.
+Typically, deployment programmes are revolving around public-private partnerships and regional funding programmes resulting in a wide mix of consortia and stakeholder subsets.
 The actual deployment of novel concepts or technological enablers is then driven by the programmes and to a lesser extent by operational network wide priorities following a piece meal approach.
 
-Moreover, the success of these deployments is typically supported by promising business cases and operational evaluations - often expressed as operational performance improvements. 
+Moreover, the success of these deployments is typically supported by promising business cases and operational performance improvements. 
 While these improvements are marketed, the underlying data and associated analyses in support of the results are seldomly made available. 
 Explanations range from data sensitivity, commercial interests and property rights, through to complexity of the data analyses or simulations, and the associated volume of data.
-
-Accordingly, ANS performance improvements are subject to a "credibility crisis" (varying levels of transparency) similar to reported problems in other domains, e.g.
+Similar practise is reported in other domains, e.g.
 medicine [@kolata_2011], psychology [@ritchie_2012], ecology [@fraser_etal_2018].
-Fidler and Wilcox (2018) provides a recent review of the reproducibility discussion [@fidler_wilcox_2018].
-Part of the crisis sensation was driven by the diffusion of reported results as other researchers/analysts were able to discover substantial flaws in the data analyses.
-However, a crisis also offers room for self-correction.
-Throughout the recent years, this lead to revisitng the concept of reproducible research.
+Fidler and Wilcox [@fidler_wilcox_2018] provides a recent review of the reproducibility discussion.
+Throughout the recent years, this lead to revisiting the concept of reproducible research.
 This concept is based on the premises that an independent researcher/analyst is enabled to reproduce the reported results.
@@
 Independent (operational) ANS performance builds on transparent and robust performance monitoring and reporting.
-Within the European context, EUROCONTROL's Performance Review Unit (PRU) is committed to provide impartial analyses of the performance of the European Air Navigation System based on verified data.
-With a view to enhance transparency, PRU is - inter alia - making performance-related data publicly available via its portal, [http://ansperformance.eu](http://ansperformance.eu), since early 2017 combined with the methodological documentation for the calculation of the performance indicators used under the EUROCONTROL Performance Revew System and the European Union Performance Scheme.
+Within the European context, EUROCONTROL's Performance Review Unit (PRU) is committed to provide impartial analyses of the ANS performance.
+With a view to enhance transparency, PRU is - inter alia - making performance-related data publicly available via its portal, [http://ansperformance.eu](http://ansperformance.eu), since early 2017 combined with the methodological documentation for the calculation of the performance indicators used under the EUROCONTROL Performance Review System and the European Union Performance Scheme.
 There has been recognition of using common and openly available tools and data in ATM research and operational validation.
@@
 This enables the independent review and validation of the work reported.
-Conceptually, an interested researcher/analyst can reproduce this paper by using the same toolset or validate the results by using the source data or interim data products if the complete reproduction of the study would be too time/resource consuming.
-
-```{r}
-#- Purpose and significance of the paper 
-#- contribution of paper
-```
-
-The contributions of this paper are as follows:
-
-* first fully reproducible ANS Performance related paper: The paper is written in RMarkdown allowing for the analysis of how text, code chunks, and visualisations are produced (including associated parameter settings) on the basis of the "source text".
-* transparent data analyses and result generation through supporting web-page: All data used in this paper, intermediate data artifacts, and results including the associated transformation are publicly available for inspection, validation, or reproduction.
-* use-case analysis of ANS Performance during the arrival phase for 2 European airports: As an example the classical 'additional ASMA time' performance indicator is implemented on the basis of the provided data with showcasing trajectory derived capabilities to further study the efficiency of the arrival phase (e.g.
-impact of sequencing strategies [stack holdings, point merge]).
-
-<!-- note: make sure that Section I does not significantly exceed first page -->
 
@@
 Within the medical world cases like the Duke ovarian cancer scandal [@kolata_2011] are not single incidents, but also demonstrate how errors during the data analysis - in this case during the data preparation phase using spread sheets - can discredit the work.
-Despite the increasing awareness, the adaption of reproducibility or moving to other open scholarship/investigation techniques is not yet achieved. 
+Despite the increasing awareness, the adoption of reproducibility or moving to other open scholarship/investigation techniques is not yet achieved. 
 This can be contrasted by emerging practices like data journalism.
@@
 The authors were able to demonstrate based on Open Data that these purposes range from tracking drug trafficking to testing new surveillance technology [@peteraldhous_2016].
-The credibility of the article comes from the fact that all artifacts are made available to reproduce the findings or study the approach and data transformations leading to the conclusions of the article.
+The credibility of the article comes from the fact that all artefacts are made available to reproduce the findings or study the approach and data transformations leading to the conclusions of the article.
 
@@
 (2017) focusses on computational reproducibility in archaeological research by providing basic principles.
-The case study of their implementation uses open source software (R/Rmarkdown/git approach)  [@marwick_2017].
+The case study of their implementation uses open source software (R/RMarkdown/git approach)  [@marwick_2017].
 The paper advocates the publishing (analytical) data in support of the paper.
 This should not be an issue, even if there are constraints on the overall study data.
-* engineering / signal processing: Vandewalle (2009) discuss the motivation and approach to reproducibility in signal processing (i.e.
+* engineering / signal processing: Vandewalle [@vandewalle_2009] discuss the motivation and approach to reproducibility in signal processing (i.e.
 what, why, and how).
-Also here the case is made to open access to the code and data for validation purposes [@vandewalle_2009].
-* operational ANS Performance monitoring: Spinielli et al. (2018) highlight the development and use of Open Flight Trajectories for reproducible ANS [@spinielli_2018].
-
-## Practical Challanges
-
-The reproducibility discussion focuses strongly on the recreation of results by other researchers/analysts.
-A related problem is the re-generating of already completed (and potentially published) results with the - previously - available same input data. 
-This is a fundamental issue that can be summarised as "can a future me re-do what I had done in the past"?
-This problem space ranges across the complete data analysis life-cyle, including archiving the orignially used input data or the steps to generate the study data, documenting the data cleaning and transformation steps including the parameters chosen, etc.
-By definition such a capability is a precondition for reproduction or replication by others. 
-This implies that the analytical steps including the steps to produce summaries or estimates, visualisations, and associated conclusions are documented and archived. 
-In order to enable others to recreate the own findings, such steps and documentation would need to be shareable.
-This consideration has a strong bearing on the toolset chosen for this paper.
-
-<!-- check if this can be merged with above -->
-Some of the reasons why researchers/analysts do not publish data and computational procedures that allow reproduction will be hard to change [@daniel_nust_2018_2361304].
-The predominant technical reasons have been listed above.
-However, there are also personal reasons such as the fear for floating ideas or loss of competitive advantage.
-Other reasons deal with more technical problems, e.g.
-lack of capabilities, procedures.
-
-Another incentive for the adoption of an open and reproducible research practice can be seen in the wider recognition through increased citations, media attention, discussion with the community and collaborators, and ultimately funding opportunities.
-The recent years have seen a strong use of research blogs and community platforms supporting the publication of smaller steps or intermediate results.
-The combination of web presentation and discussions offers a good basis for an initial dialogue.
-Platforms like google scholar or researchgate.net offer additional features for outreach and communication.
-This equally applies for paper archives such as arxiv.org for pre-prints.
-Based on the credibility crisis discussion a number of journals are now promoting - or even requiring - the provision of analytical data in support of the paper publication. 
-There is also an increasing number of requirements for national or European funded research with respect to data management and Open Data access to the research.
+Also here the case is made to open access to the code and data for validation purposes .
+* operational ANS Performance monitoring: Spinielli et al. [@spinielli_2018] highlight the development and use of Open Flight Trajectories for reproducible ANS .
 
@@
 The work of Marwick et al.
-[@marwick_2017] and Peng [@Peng1226] uses the analogy of a spectrum that can be mapped against the scoring of Vandewalle (2009) describing the level of reproducibility on a 0 to 5 scale [@marwick_2017].
+[@marwick_2017] and Peng [@Peng1226] uses the analogy of a spectrum that can be mapped against the scoring of Vandewalle [@vandewalle_2009] describing the level of reproducibility on a 0 to 5 scale [@marwick_2017].
 In light of [@leek_2015], this paper conceptualises reproducibility as the "ability to recompute data analytic results given an observed dataset and knowledge of the data analysis pipeline" and accordingly "replicability of a study is the chance that an independent experiment targeting the same scientific question will produce a consistent result".
@@
 Last but not least, the data for this paper is made available in conjunction with this text as Open Data.
-This also includes intermediate data artifacts which allows for the reproduction of the complete work or portions of it.
+This also includes intermediate data artefacts which allows for the reproduction of the complete work or portions of it.
 The conceptual approach and its building blocks will be presented in more detail in the next section.
@@
 An essential defining characteristic of analytic data is that it is the data basis on which the reported results can be produced.
-The abstraction between source data and analytic data allows for the sharing of such data as limitations are typically only appliable to the source data.
+The abstraction between source data and analytic data allows for the sharing of such data as limitations are typically only applicable to the source data.
 We follow the notion of [@marwick_2017] by referring to the provision of a representative sample of 8 days accompanying this paper as source data.
@@
 Without limiting the impact of other packages, the development of knitr [@R-knitr] and RMarkdown [@xie2018], ggplot [@wickham_2016] for visualisation, and the so-called tidyverse packages and RStudio IDE [@rstudioteam_2015] represent an open source ecosystem for data analysis.
-
-A key feature for the implementation of the reproducibility paradigm is the fact that RMarkdown supports the combination of text, coding, and visualisations in a single document. 
-The latter can be rendered into various output formats, e.g. classical paper report-styles (pdf, MS Word), html web-pages, and slides (both classical Powerpoint style or html-based presentations).
-The fact that outputs can be rendered to html pages also supports the embedding of interactive graphics.
-This also facilitates the communication and outreach amongst researchers/analysts as standard web technology becomes a standard outlet format.
-It follows that changes to the code (and associated visualisations) or text can be readily updated by re-rendering the adapted report. 
-This allows to develop standard templates for recurring reports based on varying input data, but it also serves the iterative development of larger reports (and studies).
-
+A key feature for the implementation of the reproducibility paradigm is the fact that RMarkdown supports the combination of text, analytical code, and visualisations in a single document. 
 A minimal example of an RMarkdown document, a plain-text file with the conventional
@@
 output document (title, type of document [web page, PDF, MS Word], ...)
-The text, which follows the metadata, is witten using Markdown, a simplified formatting syntax
+The text, which follows the metadata, is written using Markdown, a simplified formatting syntax
 which can include prose and code.
@@
 - A code chunk starts with three backticks and `r` indicates the
-  language name,^[R Markdown supports 50 other languages.] and ends with
+  language name,^[R Markdown supports 50 other computer languages.] and ends with
   three backticks.
@@
 
-```{r rmd-example,echo=FALSE,fig.cap='(ref:rmd-example)',fig.pos='htb',fig.height=3}
+```{r rmd-example,echo=FALSE,fig.cap='(ref:rmd-example)',fig.pos='htb',fig.height=2}
 knitr::include_graphics('figures/rmd-example.png', dpi = NA)
@@
 
-A particular strong point is that the R/RStudio ecosystem is open source software.
-This includes also its extensibility with other packages.
-There exists a wide variety of packages for data analytical tasks and techniques.
-This mechanism can also be useful for the communication of results.
-Researchers/analysts are able to provide their code and supporting documentation as a package for other researchers to build on.
-For example, throughout the last year, PRU has developed certain utility functions for reoccurring tasks of the data analytical pipeline.
-It is now planned to wrap these functions into a package saving other researchers/analysts of recoding certain steps.
-For this paper, the functions and data transformations are reported as part of the RMarkdown and supporting R files.
-All these files are available via the project webpage.
-
 ## Capabilities: Performance Reference Trajectory
@@
 Throughout the recent year PRU developed the data analytical capabilities to fuse correlated position reports and ADSB surveillance data.
-This work has been reported in [@koelle_open_2017] and [@spinielli_2018]. The idea is to establish a pan-European air situation picture openly available to research and practitioners.
-Conceptually, the fused performance reference trajectory will be a key milestone in combatting the "marketing myths."
+This work has been reported in [@koelle_open_2017] and [@spinielli_2018]. The idea is to establish a pan-European air situation picture openly available to researchers and practitioners.
+Conceptually, the performance reference trajectory will be a key milestone in combatting the "marketing myths."
 For example, operational improvements can be directly assessed by interested analysts as access to the Open Data enables the validation of the observable performance improvements.
-This allows for the application of commonly accepted methods and models to express performance benefits and  - ultimately - challenge related publications on deployments and achieved improvements.
+This allows for the application of commonly accepted methods and models to express performance benefits and  ultimately challenge related publications on deployments and achieved improvements.
 
@@
 The goal is to establish and operate the underlying data processing capabilities throughout 2019 and launch the regular provision of the performance reference trajectory as Open Data in 2020.
-It is also planned to establish a network and complement the air situation picture by deploying additional sensors or determining non-ADSB derived positions as a community effort.
-As a mid-term goal associated data bases and models for performance shall be established and jointly maintained.
-
-<!-- not word by word - but the info is contained elsewhere 
-## Application Use Cases: ASMA, Point Merge, holdings
-
-This paper revolves around advancing the state of the art by demonstrating the implementation of the reproducibility paradigm.
-Accordingly, this paper builds on the concepts behind Fig. \@ref(fig:reproducible-research) and Fig. \@ref(fig:concept-approach).
-This includes providing the data, code, and results.
-All of this is version controlled and archived on <<LINK TO ARCHIVE OR WEB PAGE>>.
-
-The approach and processing pipeline described in this paper builds on several data sets.
-...
-
-As a practical application use case, this paper replicates the additional ASMA time algorithm.
-... describe AMSA here ... which data sets used ... how compared ... point to next section for details of this.
-
-In a second step, the ASMA principle is moved to a distance based measure.
-... describe what will be done ... how does this link to what we want to achieve ... point to next section for more details.
-
-As examples for further developments based on the aforementioned capabilities, this paper presents a closer look at arrival flow management strategies.
-
--->
 
@@
 
-(ref:data-preparation-pipeline) Data preparation pipeline.
-
-```{r data-preparation-pipeline,fig.cap='(ref:data-preparation-pipeline)',fig.pos='htb',fig.height=4}
-knitr::include_graphics("figures/data-preparation-pipeline.png")
-```
-
 The raw data sources for the use cases of this paper have been processed by a pipeline of R scripts.
-For example, the additional ASMA time indicator and the anticipated further breakdown of the arrival phase can be described by a set of tranistion points:
+For example, the additional ASMA time indicator and the anticipated further breakdown of the arrival phase can be described by a set of transition points:
 
@@
 
+(ref:data-preparation-pipeline) Data preparation pipeline.
+
+```{r data-preparation-pipeline,fig.cap='(ref:data-preparation-pipeline)',fig.pos='htb',fig.height=2.5}
+knitr::include_graphics("figures/data-preparation-pipeline.png")
+```
+
 Figure \@ref(fig:data-preparation-pipeline) shows how the transition points 
@@
 have been extracted.
-Due to the size of the daily reference trajectory files, the pipeline operates
-on one date at the time.
-Assuming the relevant RT flight and position files are downloaded
-from GCP as described in [@performancereviewunit_2019],
-the steps and the relevant scripts (available in the companion repository for this paper) used are:
-
-1. the script `R/extract-arrivals.R` extracts arrivals to the `APT` airport and generates:
-
-    * a file with flight details
-    * a file with all the arrivals' positions (`raw` trajectories)
-    * a file with all the arrivals' positions (smoothed trajectories)
-    * a file with all (smoothed) arrivals' positions within 40 NM of the airport's reference position (ARP)
+The following steps and scripts comprise [@performancereviewunit_2019]:
+
+1. the script `R/extract-arrivals.R` extracts arrivals to the `APT` airport and generates files with:
+
+    * flight details
+    * arrivals' positions (`raw`  and smoothed trajectories)
+    * (smoothed) arrivals' positions within 40 NM of the airport's reference position (ARP)
 
@@
 
-(ref:eidw-arrival-chart-east) Excerpt of Irish AIP showing point merge arc layout East of Dublin Airport.
-
-```{r eidw-arrival-chart-east,fig.cap='(ref:eidw-arrival-chart-east)',fig.pos='hbt',fig.height=4}
-knitr::include_graphics("figures/eidw-arrival-chart-east.png")
-```
-
-  This stage makes use of data extracted/extrapolated from the relevant AIP, see
-  for example Fig. \@ref(fig:eidw-arrival-chart-east), such
+  This stage makes use of data extracted/extrapolated from the relevant AIP such
   as ARP, runway name and threshold position/elevation, polygons for holding
@@
 
-* aircraft/flight info (ID, callsign, registration, ADEP/ADES, ICAO 24 bit address)
-* 4D position (timestamp $T$, longitude, latitude, altitude)
-* flown distance $D$ from aerodrome of departure (ADEP)
-* sequence number
-* point info (sequence number, type of transition point [`P40`, `PHOLD`, `PLAND`])
-* distance from ARP
+* aircraft/flight info: ID, callsign, registration, ADEP/ADES, ICAO 24 bit address;
+* 4D position: timestamp $T$, longitude, latitude, altitude;
+* position info: flown distance $D$ from aerodrome of departure (ADEP), sequence number,
+  distance to ARP as a proxy of aerodrome of destination (ADES);
+* point info: sequence number, type of transition point [`P40`, `PHOLD`, `PLAND`]
 
@@
 
-The previous sections described the data preparatory action.
 For this paper, a reproducible data set for the analysis of arrival management techniques at London Heathrow and Dublin airport have been extracted. 
@@
     #legend.position = "bottom"
-    legend.position = c(1, 0.65), legend.justification = c(1, 0)
+    legend.position = c(0.95, 0.60), legend.justification = c(1, 0)
     , axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5)
@@
 ```
-
-Fig.
-\@ref(fig:coverage-egll) shows the overlap of the data between the arrivals extracted from the airport operator data flow (APDF) and the reference trajectory (RTRJ) for London Heathrow (EGLL) and Dublin (EIDW).
 
@@
 
+Fig.
+\@ref(fig:coverage-egll) shows the overlap of the data between the arrivals extracted from the airport operator data flow (APDF) and the reference trajectory (RTRJ) for London Heathrow (EGLL) and Dublin (EIDW).
 For the time horizon of this paper, the coverage of the reference trajectory derived data is for Heathrow (EGLL) `r round(cov_perc$avg[cov_perc$ADES == "EGLL"], 2) * 100`% and for Dublin `r round(cov_perc$avg[cov_perc$ADES == "EIDW"], 2) * 100`%.
@@
 [http://ansperformance.eu/references/definition/additional_asma_time.html](http://ansperformance.eu/references/definition/additional_asma_time.html)).
-These include 1.) the arrival entry sector at 40NM from the airport, 2.) the aircraft weigth turbulence category and engine type, and 3.) the actual landing runway.
+These include 1.) the arrival entry sector at 40NM from the airport, 2.) the aircraft weight turbulence category and engine type, and 3.) the actual landing runway.
 The reference times are built for a longer time horizon and extracted from the PRU data base as additional source data.
@@
 Heathrow operates holding stacks to tactically optimise the arrival throughput.
-Dependent on the traffic situation, aircraft may experience longer holding times (i.e. long tails of boxplots in Fig. \@ref(fig:ASMA-travel-egll)).
+Dependent on the traffic situation, aircraft may experience longer holding times (i.e. long tails of box plots in Fig. \@ref(fig:ASMA-travel-egll)).
 
@@
 Fig.
-\@ref(fig:ASMA-compare) compares the additional ASMA time extracted from the PRU performance dashboard, the trajectory based calculation with the dashboard derived reference time, and a 20th percentile approach purely based on the trajectory data.
+\@ref(fig:ASMA-compare) compares the additional ASMA time extracted from the PRU performance dashboard, the trajectory based calculation with the dashboard derived reference time, and a 20-th percentile approach purely based on the trajectory data.
 
@@
 TRJ-REF), there is a strong increase of the additional ASMA time for Heathrow (EGLL), while in the case of Dublin (EIDW) a reasonable decrease takes place.
-Calculating the additional ASMA time exclusively with the trajectory sample and applying a 20th percentile approach (TRJ-20PCT), shows a significant drop in EGLL and a moderate increase at Dublin.
+Calculating the additional ASMA time exclusively with the trajectory sample and applying a 20-th percentile approach (TRJ-20PCT), shows a significant drop in EGLL and a moderate increase at Dublin.
 The high variation for Heathrow shows that the additional ASMA time is strongly dependent on the time horizon.
 The pure trajectory based measure - by definition - includes the effect of the operational concept (stack holdings) and distorts the measure.
-In the case of Dublin it can be argued that the TRJ-20PCT measure reflects the time based ASMA measure suffiently.
+In the case of Dublin it can be argued that the TRJ-20PCT measure reflects the time based ASMA measure sufficiently.
 
@@
 
-conceptually, measuring flight inefficiency in terms of flight time accounts for the airspace user perspective.
+Conceptually, measuring flight inefficiency in terms of flight time accounts for the airspace user perspective.
 The previous section has also shown the high dependency of the time based measure on the sample size or reference time attribution.
-Flight time is linked to engine time and accordingly, assuming maneuvering for operational separation and sequencing reasons, fuel burn.
+Flight time is linked to engine time and accordingly, assuming manoeuvring for operational separation and sequencing reasons, fuel burn.
 From a service provision perspective, efficiency relates to the capability to provide the least distance flown within the ASMA area.
@@
 
-```{r ASMA-time-dist, fig.cap='(ref:ASMA-time-dist)', message = FALSE}
+```{r ASMA-time-dist, fig.cap='(ref:ASMA-time-dist)', message = FALSE,fig.pos='hbt',fig.height=4}
 prep_asma_dist <- function(apt) {
@@
 This paper applies a simple 15-th percentile approach to the distance flown per arrival sector, aircraft type/engine, and landing runway.
-
-```{r tbl-asma-dist}
+Tab. \@ref(tab:tbl-asma-dist) lists the determined average additional distance at the study airports.
+The total additional distance is determined as the aggregated sum of the differences in each segment
+with 15-th percentile reference distance.
+
+```{r tbl-asma-dist,fig.pos='hbt'}
 asma_dist <- function(ds_pts, ds_dist){
@@
 knitr::kable(dist_asma, digits = 2
-             ,caption = "Average additional distance flown (15th pct)")
-```
-
-Tab. \@ref(tab:tbl-asma-dist) lists the determined average additional distance at the study airports.
-The total additional distance is determined as the aggregated sum of the differences in each segmnent
-with 15-th percentile reference distance.
+             ,caption = "Average additional distance flown (15-th pct)")
+```
 
 ## Arrival Management Practices
-
-For an initial drill down, this paper uses arrival practices at London Heathrow and Dublin.
-The approach to managing the arrival flow differs significantly between both airports.
-Conceptually, London Heathrow employs the "pressure on the runway" concept with a focus on achieving and balancing a continuously high level of runway throughput utilisation. 
-This is achieved by controlling the demand through 4 arrival stacks that allow for the tactical sequencing on the last portion of the approach.
-Dublin is one of the airports that pioneered the implementation of "point merge".
-Point merge aims at replacing circular holdings by longitudal holdings to establish the sequence based on predefined legs. 
-<!-- ADD A FEW MORE WORDS ON THE CONCEPTS IF NEEDED -->
-
-From a performance perspective it is therefore interesting to identify and measure the impact of the operational concept on the overall flight efficiency during the arrival phase of flight.
 
@@
 
+For an initial drill down, this paper uses arrival practices at London Heathrow and Dublin.
+The approach to managing the arrival flow differs significantly between these airports.
+Conceptually, London Heathrow employs the "pressure on the runway" concept with a focus on achieving and balancing a continuously high level of runway throughput utilisation. 
+This is accomplished by controlling the demand through 4 arrival stacks that allow for the tactical sequencing on the last portion of the approach.
+Instead Dublin is one of the airports that pioneered the implementation of "point merge".
+Point merge aims at replacing circular holdings by longitudinal holdings to establish the sequence based on predefined legs.
+
 (ref:egll-trajectory-example) EGLL - long trajectory flown on 1. Aug 2017.
 
-```{r egll-trajectory-example, fig.cap='(ref:egll-trajectory-example)'}
+```{r egll-trajectory-example, fig.cap='(ref:egll-trajectory-example)',fig.pos='htb',fig.height=4}
 plot_egll_tma() + geom_point(
@@
 
-```{r eidw-trajectory-example, fig.cap='(ref:eidw-trajectory-example)'}
+```{r eidw-trajectory-example,fig.cap='(ref:eidw-trajectory-example)',fig.pos='bht',fig.height=6.5}
 eidw_long_dist <- asma_dist2(eidw_pts, eidw_dist) %>% 
@@
 
-eidw_long_dist <- eidw_long_dist %>% mutate(IS_DAY1 = ifelse(flight_id %in% eidw_day1$flight_id, 1, 0)) %>%
-  filter(IS_DAY1 == 1) %>% filter(ADD_DIST < 50) %>% 
+eidw_long_dist <- eidw_long_dist %>%
+  mutate(IS_DAY1 = ifelse(flight_id %in% eidw_day1$flight_id, 1, 0)) %>%
+  filter(IS_DAY1 == 1) %>%
+  filter(ADD_DIST < 50) %>% 
   filter(flight_id %in% c(
@@
 
-source(here::here("R/plot-eidw-tma.R"))
-
-plot_eidw_tma() + geom_point(
-  data = eidw_day1 %>% select(flight_id, latitude, longitude)
-  , mapping = aes(x = longitude, y = latitude, colour = flight_id)
-  ) +
-  theme_bw() +
-  theme(legend.position = "none")
-```
+library(dplyr)
+library(sf)
+library(ggplot2)
+library(units)
+
+fids <- c(
+  "80788782-6fa7-42f5-b636-b5eee6822aab"
+  ,"bbef0049-6748-49dc-ada3-8d058a7a03a6"
+  #  ,"8df967dd-6880-4fda-86c3-a573a32abb19"
+  ,"2db1b8a1-28ce-4dbf-9115-7404940bc832"
+  ,"4d3496e0-06ef-4601-9339-1bc2367fe18f"
+)
+
+ff <- "data/eidw_flt_rt_2017-08-01.csv" %>% 
+  vroom::vroom()
+
+pp40 <- "data/eidw_pos_rt_40NM_2017-08-01.csv" %>% 
+  vroom::vroom() %>%
+  group_by(flight_id) %>% 
+  filter(flight_id %in% fids) %>% 
+  dplyr::arrange(timestamp) %>%
+  dplyr::left_join(ff, by = "flight_id")
+
+source(here("R", "utils.R"), encoding = 'UTF-8', local = TRUE)
+source(here("R", "eidw-data.R"), encoding = 'UTF-8', local = TRUE)
+
+buffer <-  10
+
+bbox <- pp40 %>%
+  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
+  sf::st_convex_hull() %>%
+  sf::st_transform(3857) %>%
+  sf::st_buffer(buffer) %>%
+  sf::st_transform(4326) %>%
+  sf::st_bbox() %>%
+  # TODO: use 'round away from zero' eventually
+  #       https://shrektan.com/post/2018/04/25/human-round-round-to-nearest-ties-away-from-zero/
+  # round() %>%
+  as.numeric()
+
+names(bbox) <- c("left", "bottom", "right", "top")
+
+eidw_vor <- eidw_vor %>%
+  dplyr::filter(!id %in% c("BOYNE", "BAGSO", "SORIN"))
+
+stacks <- eidw_vor %>%
+  dplyr::mutate(hjust = 0, vjust = -0.7) %>%
+  # TODO: adapt to EIDW
+  dplyr::mutate(
+    hjust = ifelse(id == "PEKOK",  1.0, hjust),
+    vjust = ifelse(id == "PEKOK",  1.5, vjust)) %>%
+  dplyr::mutate(
+    hjust = ifelse(id == "EPIDU",  1.2, hjust),
+    vjust = ifelse(id == "EPIDU", 0.7, vjust)) %>%
+  dplyr::mutate(
+    hjust = ifelse(id == "KERAV",  1.2, hjust),
+    vjust = ifelse(id == "KERAV",  1.5, vjust)) %>%
+  dplyr::mutate(
+    hjust = ifelse(id == "OCL",  -0.2, hjust),
+    vjust = ifelse(id == "OCL", -0.7, vjust))
+
+buf <- set_units(2, nmile) %>%
+  set_units(m)
+
+gvor <- ggplot2::geom_point(data = eidw_vor,
+                            ggplot2::aes(x = longitude,
+                                         y = latitude),
+                            colour = "blue",
+                            size = 2)
+gvorl <- ggplot2::geom_text(data = stacks,
+                            ggplot2::aes(x = longitude,
+                                         y = latitude,
+                                         label = id,
+                                         hjust = hjust,
+                                         vjust = vjust))
+
+eidw_holdings_sf_buf <- eidw_holdings_sf %>%
+  st_transform(29902) %>%
+  st_buffer(buf) %>%
+  st_transform(4326)
+
+grw_sf <- ggplot2::geom_sf(data = eidw_rw_sf, size = 1.2)
+
+world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
+
+ggplot2::ggplot(pp40) +
+  ggplot2::geom_sf(data = world) +
+  ggplot2::geom_path(
+    data = pp40,
+    mapping = ggplot2::aes_(
+      x = quote(longitude), y = quote(latitude),
+      colour = quote(callsign),
+      group = quote(callsign)
+    ),
+    size = 1.4, alpha = .3, lineend = "round") +
+  ggplot2::geom_sf(data = eidw_holdings_sf_buf, alpha = 0) +
+  grw_sf +
+  gvor + gvorl +
+  ggplot2::coord_sf(xlim = c(bbox["left"], bbox["right"]),
+                    ylim = c(bbox["bottom"], bbox["top"]),
+                    expand = TRUE) +
+  ggplot2::theme_minimal() +
+  ggplot2::theme(panel.background = ggplot2::element_rect(fill = "aliceblue"),
+                 legend.position = "none")
+
+```
+
+<!-- ADD A FEW MORE WORDS ON THE CONCEPTS IF NEEDED -->
+
+From a performance perspective it is therefore interesting to identify and measure the impact of the operational concept on the overall flight efficiency during the arrival phase of flight.
 
@@
 for long distances flown on 1. Aug 2017.
-The trajectory data allows for the identification of the holdings and/or sequencing area.
-This allows to separate the effect of the operational concepts.
+The trajectory data allows for the identification of the holdings and/or sequencing area so as
+to separate the effect of the operational concepts.
 
@@
 
-<!-- TO DO: clean figure - work out bearings -- do it for Dublin as well -->
-
-```{r distance-flown-segment, fig.cap="(ref:distance-flown-segment)"}
+```{r distance-flown-segment, fig.cap="(ref:distance-flown-segment)",fig.height=4}
 segment_approach <- function(ds) {
@@
   theme(axis.text.x = element_text(angle = 45, hjust = 0.5, vjust = 0.5)
-        ,legend.position = c(1, 0.65), legend.justification = c(1, 0)) +
+        ,legend.position = c(0.95, 0.60), legend.justification = c(1, 0)) +
   labs(x = "arrival entry sectors [radial from ARP]"
@@
 ```
+
+Fig. \@ref(fig:distance-flown-segment) depicts the flown distances per segment of the arrival phase.
+In the case of London Heathrow the classical stack based holding can be seen in the significant variations of the actual distance flown per arrival direction.
+Only limited sequencing takes place during the first segment (arrival within 40 NM and the stacks).
 
@@
 
-Fig. \@ref(fig:distance-flown-segment) depicts the flown distances per segment of the arrival phase.
-In the case of London Heathrow the classical stack based holding can be seen in the significant variations of the actual distance flown per arrival direction.
-Only limited sequencing takes place during the first segment (arrival within 40 NM and the stacks).
 Fig.
@@
 The middle segment shows a wider variation and less uniform distribution than the approach at London Heathrow.
+
+Fig. \@ref(fig:eidw-saturation-example) depicts the variation of the arrival segment throughput at Dublin (EIDW).
+The median for every set of flights in the ASMA area versus flights landing is highlighted.
+The behaviour is a classical saturation curve, i.e.
+up to an arrival bunching of 10-11 flights per 20 min (30-33 flights per hour), the point merge procedure seems to be able to support the tactical absorption and sequencing.
+It is noteworthy that the saturation curve falls for higher arrival throughputs.
+As no measurements for 12 through 14 landings had been made in the sample, this behaviour could be caused by the chosen time horizon.
 
@@
 
-```{r eidw-saturation-example, fig.cap='(ref:eidw-saturation-example)'}
+```{r eidw-saturation-example, fig.cap='(ref:eidw-saturation-example)',fig.pos='hb'}
 g <- ggplot(data = rq %>% na.omit, mapping = aes(x = as.factor(THRU_NG), y = THRU_N2 + THRU_N3)) + 
@@
 ```
-Fig. \@ref(fig:eidw-saturation-example) depicts the variation of the arrival segment throughput at Dublin (EIDW).
-The median for every set of flights in the ASMA area versus flights landing is highlighted.
-The behavour is a classical saturation curve, i.e.
-up to an arrival bunching of 10-11 flights per 20 min (30-33 flights per hour), the point merge procedure seems to be able to support the tactical absorption and sequencing.
-It is noteworthy that the saturation curve falls for higher arrival throughputs.
-As no mearurements for 12 through 14 landings had been made in the sample, this behaviour could be caused by the chosen time horizon.
 
@@
 The work reported establishes a fully reproducible paper for ANS performance in Europe.
-It is the one of the first attempts to address the reproducibility crisis within ANS performance by providing the artifacts that allow an independent researcher or professional to reproduce the results and verify upstream processing steps.
-
-```{r}
-# Limitations and advantages.
-```
+It is the one of the first attempt to provide the artefacts that allow an independent researcher or professional to reproduce the results and verify upstream processing steps.
 This paper follows up on previous work [@koelle_open_2017; @spinielli_2017; @spinielli_2018] and forms part of the roll-out of the performance reference trajectory by the PRU.
@@
 
-To overcome these limitations, PRU will be looking into deploying the trajectory processing on a cloud-based platform including the regular provision of the identified analytical data levels.
-
 The move to open data for ANS Performance is a *giant leap* in terms of transparency.
@@
 It removes the burden to acquire and preprocess large-scale surveillance data and associated aircraft data or aeronautical information.
-PRU envisages a strong collaboration through a network of research and practitioners.
 The departure point for this development was to address the lack and hurdle of verifying results.
 A further advantage is that novel applications, methods and algorithms (e.g. machine learning), can be applied to a harmonised dataset.
-It is hoped that development efforts can accelerate while offering a platform for improvements to existing processing methods and algorithms.
-
-```{r}
-# Applications of the results.
-```
-The major application of the paper results will inform the performance reference trajectory roll-out project.
+
+The major application of the paper results will inform the performance reference trajectory project.
 This goes in hand with the iterative move of the operational performance measures to trajectory derived data.
@@
 This paper documents a key milestone in the project. 
-As such the future work will address:
-
-* roll-out and establishment of the performance reference trajectory: This includes the organisational and technical work to deploy the processing modules and arrange for a regular processing of the source data and its provision as Open Data.
-* establishing the current performance reporting based on the trajectory: Since more than 2 years PRU addresses the need for Open Data by providing monthly updates for the key performance indicators used under the EUROCONTROL Performance Review System and the SES Performance Scheme.
-Underlying analytic data is made available via [http://ansperformance.eu](http://ansperformance.eu).
-As part of the roll-out, it is foreseen to move associated indicators to a trajectory derived or at least verified process and expand on the analytical data level.
-* establishment of a network: The underlying idea is that various levels of future work can be orchestrated in close collaboration with interested researchers and professionals.
-This can span work addressing the improvement of the actual data processing (e.g.
-less resource intensive computing) to the further development and validation of novel techniques for measuring operational performance (e.g.
-application of machine learning, time series analysis).
+As such the future work will address the roll-out of the performance reference trajectory,
+the establishment of the current performance reporting based on the trajectory (via [http://ansperformance.eu](http://ansperformance.eu), and setup of a network with interested researchers and professionals.
 

### 2019-10-04 paper.tex
final code version

@@
 \hypersetup{
-            pdftitle={Busting the Myth! - Reproducibility in ANS Performance},
+            pdftitle={Towards Reproducibility in ANS Performance},
             pdfauthor={Enrico Spinielli, Rainer Koelle},
@@
 \newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
 \newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
 \newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
 \newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
 \newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
 \newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
 \newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
 
@@
 % Do not put math or special symbols in the title.
-\title{Busting the Myth! - Reproducibility in ANS Performance}
+\title{Towards Reproducibility in ANS Performance}
 
@@
 \begin{abstract}
-Air transportation is undergoing a fundamental transformation. On the
-political, strategic, and operational level work is directed to meet the
-future growth of air traffic. To meet this goal, operational concepts
-and technological enablers are deployed with the promises of distinct
-performance improvements. These improvements are typically widely
-marketed, however, the underlying evidence is not made available. From
-that respect Air Navigation System Performance suffers from the same
-credibility crisis seen in other industries. This paper implements the
-reproducibility paradigm by providing data, code, and results openly to
-enable another researcher or analyst to reproduce - and potentially -
-improve or build on our work. This work is part of a wider thread to
-establish Open Data and Open Source based data analytical capabilities
-for Air Navigation System Performance in Europe. The implementation of
-the paradigm is demonstrated on the basis of the performance reference
-trajectory by replicating a classical performance measure, and then
-moving into investigating the application of trajectory based
-information for enhancing the current state of performance monitoring
-for the arrival phase. This paper applies the concept and approach as an
-use case analysis for 2 European airports with different strategies for
-sequencing arrival traffic (i.e.~stack holding and point merge). The
-work reported demonstrates the feasibility of the approach and reflects
-a stepping stone in the roll-out of the performance reference trajectory
-within the European context. The work also builds the basis for novel
-developments on the basis of the implemented reproducibility workflow as
-this paper is accompanied by a webpage that supports the impartial
-evaluation of the work by providing the data, code, and how this reports
-had been produced.
+Air transportation is undergoing a fundamental transformation. On the political, strategic, and operational level work is directed to meet the future growth of air traffic.
+To meet this goal, operational concepts and technological enablers are deployed with the promises of distinct performance improvements.
+These improvements are typically widely marketed, however, the underlying evidence is not made available.
+From that respect Air Navigation System Performance faces varying levels of transparency as seen in other industries.
+This paper implements the reproducibility paradigm by providing data, code, and results openly to enable another researcher or analyst to reproduce - and potentially - improve or build on our work.
+This work is part of a wider thread to establish Open Data and Open Source based data analytical capabilities for Air Navigation System Performance in Europe.
+The implementation of the paradigm is demonstrated on the basis of the performance reference trajectory by replicating a classical performance measure, and then moving into investigating the application of trajectory based information for enhancing the current state of performance monitoring for the arrival phase.
+This paper applies the concept and approach as a use case analysis for 2 European airports with different strategies for sequencing arrival traffic (i.e.~stack holding and point merge).
+The research reported demonstrates the feasibility of the reproducibility approach.
+This work introduces a novel workflow by making available a companion web repository that provides data, code, and instructions to fully reproduce the paper.
 \end{abstract}
@@
 \begin{IEEEkeywords}
-\textit{%
-reproducibility; ANS performance; flight trajectory; open data; performance indicator}
+reproducibility; ANS performance; flight trajectory; open data; performance indicator
 \end{IEEEkeywords}
@@
 \maketitle
+
+% no keywords
 
@@
 
-\section{Introduction}\label{introduction}
-
-Air transportation is undergoing a fundamental transformation. To meet
-the future growth of air traffic, political decision makers, strategic
-planers, practitioners and researchers aim to enhance the performance of
-air navigation services (ANS). At the time being there is no consistent
-and regionally harmonised deployment programme. Typically, the
-deployment programmes are revolving around public-private partnerships
-and regional funding programmes resulting in a wide mix of consortia and
-stakeholder subsets. The actual deployment of novel concepts or
-technological enablers is then driven by the programmes and to a lesser
-extent by operational network wide priorities following a piece meal
-approach.
-
-Moreover, the success of these deployments is typically supported by
-promising business cases and operational evaluations - often expressed
-as operational performance improvements. While these improvements are
-marketed, the underlying data and associated analyses in support of the
-results are seldomly made available. Explanations range from data
-sensitivity, commercial interests and property rights, through to
-complexity of the data analyses or simulations, and the associated
-volume of data.
-
-Accordingly, ANS performance improvements are subject to a ``credibility
-crisis'' similar to reported problems in other domains, e.g.~medicine
-{[}\protect\hyperlink{ref-kolata_2011}{1}{]}, psychology
-{[}\protect\hyperlink{ref-ritchie_2012}{2}{]}, ecology
-{[}\protect\hyperlink{ref-fraser_etal_2018}{3}{]}. Fidler and Wilcox
-(2018) provides a recent review of the reproducibility discussion
-{[}\protect\hyperlink{ref-fidler_wilcox_2018}{4}{]}. Part of the crisis
-sensation was driven by the diffusion of reported results as other
-researchers/analysts were able to discover substantial flaws in the data
-analyses. However, a crisis also offers room for self-correction.
-Throughout the recent years, this lead to revisitng the concept of
-reproducible research. This concept is based on the premises that an
-independent researcher/analyst is enabled to reproduce the reported
-results. Reproducibility covers therefore a spectrum and mix of method,
-data, and analytical code/scripts per se openly available for scrutiny
-by others.
-
-Independent (operational) ANS performance builds on transparent and
-robust performance monitoring and reporting. Within the European context
-EUROCONTROL's Performance Review Unit (PRU) is committed to provide
-impartial analyses of the performance of the European Air Navigation
-System based on verified data. With a view to enhance transparency, PRU
-is - inter alia - making performance-related data publicly available via
-its Portal, \url{http://ansperformance.eu}, since early 2017 combined
-with the methodological documentation for the calculation of the
-performance indicators used under the EUROCONTROL Performance Revew
-System and the European Union Performance Scheme. There has been
-recognition of using common and openly available tools and data in ATM
-research and operational validation. With the advent of crowd sourced
-data and aviation databases there is an opportunity to advance the state
-of the art by combining open data and government organisation collected
-data to establish a curated platform supporting the quest for
-reproducibility in ANS Performance. One of the associated building
-blocks is the performance reference trajectory
-{[}\protect\hyperlink{ref-spinielli_2018}{5}{]}. The performance
-reference trajectory will be deployed as Open Data throughout 2019. The
-trajectory project builds on crowd sourced surveillance data fused with
-correlated position reports collected from European air navigation
-service providers. Based on the Open Data model it will be possible to
-augment the data processing (e.g.~data coverage, time synchronisation),
-establishment of derived data bases (e,g fleet data base), associated
-performance models (e.g.~aircraft performance model), but also the
-continual further development and validation of operational performance
-measures/indicators.
-
-This paper builds on the previous work and implements the
-reproducibility paradigm. As a use-case example, a fully reproducible
-paper is devised assessing the performance in the arrival phase for 2
-European airports. The underlying data, supporting data transformation
-and analysis, and the paper ``source text'' are made openly available at
-the paper companion web repository
-\url{https://github.com/euctrl-pru/reproducible-ans-performance-paper}.
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+To meet the future growth of air traffic, political decision makers, strategic planers, practitioners and researchers aim to enhance the performance of air navigation services (ANS).
+Typically, deployment programmes are revolving around public-private partnerships and regional funding programmes resulting in a wide mix of consortia and stakeholder subsets.
+The actual deployment of novel concepts or technological enablers is then driven by the programmes and to a lesser extent by operational network wide priorities following a piece meal approach.
+
+Moreover, the success of these deployments is typically supported by promising business cases and operational performance improvements.
+While these improvements are marketed, the underlying data and associated analyses in support of the results are seldomly made available.
+Explanations range from data sensitivity, commercial interests and property rights, through to complexity of the data analyses or simulations, and the associated volume of data.
+Similar practise is reported in other domains, e.g.~medicine {[}\protect\hyperlink{ref-kolata_2011}{1}{]}, psychology {[}\protect\hyperlink{ref-ritchie_2012}{2}{]}, ecology {[}\protect\hyperlink{ref-fraser_etal_2018}{3}{]}.
+Fidler and Wilcox {[}\protect\hyperlink{ref-fidler_wilcox_2018}{4}{]} provides a recent review of the reproducibility discussion.
+Throughout the recent years, this lead to revisiting the concept of reproducible research.
+This concept is based on the premises that an independent researcher/analyst is enabled to reproduce the reported results.
+Reproducibility covers therefore a spectrum and mix of method, data, and analytical code/scripts per se openly available for scrutiny by others.
+
+Independent (operational) ANS performance builds on transparent and robust performance monitoring and reporting.
+Within the European context, EUROCONTROL's Performance Review Unit (PRU) is committed to provide impartial analyses of the ANS performance.
+With a view to enhance transparency, PRU is - inter alia - making performance-related data publicly available via its portal, \url{http://ansperformance.eu}, since early 2017 combined with the methodological documentation for the calculation of the performance indicators used under the EUROCONTROL Performance Review System and the European Union Performance Scheme.
+There has been recognition of using common and openly available tools and data in ATM research and operational validation.
+With the advent of crowd sourced data and aviation databases there is an opportunity to advance the state of the art by combining open data and government organisation collected data to establish a curated platform supporting the quest for reproducibility in ANS Performance.
+One of the associated building blocks is the performance reference trajectory {[}\protect\hyperlink{ref-spinielli_2018}{5}{]}.
+At the time of writing, the performance reference trajectory is in its design stage.
+The trajectory project builds on crowd sourced surveillance data fused with correlated position reports collected from European air navigation service providers.
+Based on the Open Data model it will be possible to augment the data processing (e.g.~data coverage, time synchronisation), establish derived data bases (e,g fleet data base), and derive associated performance models (e.g.~aircraft performance model).
+This setup enables the continual further development and validation of operational performance measures/indicators.
+
+This paper builds on the previous work and implements the reproducibility paradigm.
+As a use-case example, a fully reproducible paper is devised assessing the performance in the arrival phase for 2 European airports.
+The underlying data, supporting data transformation and analysis, and the paper ``source text'' are made openly available at the paper companion web repository \url{https://github.com/euctrl-pru/reproducible-ans-performance-paper}.
 This enables the independent review and validation of the work reported.
-Conceptually, an interested researcher/analyst can reproduce this paper
-by using the same toolset or validate the results by using the source
-data or interim data products if the complete reproduction of the study
-would be too time/resource consuming.
-
-The contributions of this paper are as follows:
+
+\hypertarget{background---reproducibility}{%
+\section{Background - Reproducibility}\label{background---reproducibility}}
+
+\hypertarget{scope-and-definition}{%
+\subsection{Scope and Definition}\label{scope-and-definition}}
+
+Conceptually, reproducibility is not a new topic.
+It is as old as the scientific method and at the heart of the verification/validation of results.
+The consistent and robust interpretation of results by independent researchers/analysts forms the evidence and the inherent body of knowledge in support of the applied methods or conclusions.
+This principle is also relevant for the further development on the basis of such results.
+
+The terms replication and reproducibility are often used interchangeably and there is a debate about the necessary and sufficient conditions for these terms.
+Results are considered replicable, ``if there is sufficient information available for independent researchers to make the same {[}consistent{]} findings using the same procedures with new data'' ({[}\protect\hyperlink{ref-gandrud_2015}{6}{]}, p.~3f).
+Replicability, thus, does not aim to produce the ``same'' (aka identical) results.
+Here the goal is to establish confirmation of the approach and findings by deriving same conclusions.\footnote{For example, the same data collection procedures may not necessarily result in ``identical'' results as the time horizon might be different, (new or different) data for another study case show different result values, but the conclusions are consistent with the replicated approach.}
+Accordingly, results are reproducible, if an independent researcher can recreate the findings based on the given information.
+In our domain (i.e.~operational ANS Performance and associated data analysis) this will require the access to data and derived summaries, how associated visualisations have been generated, the accompanying explanatory text, and conclusions drawn from the analysis.
+
+Reproducibility has also to be seen in light of today's ``publish or perish'' context.
+It appears that higher emphasis is given to the frequency of publication than the validity of the produced results.
+Within the medical world cases like the Duke ovarian cancer scandal {[}\protect\hyperlink{ref-kolata_2011}{1}{]} are not single incidents, but also demonstrate how errors during the data analysis - in this case during the data preparation phase using spread sheets - can discredit the work.
+Despite the increasing awareness, the adoption of reproducibility or moving to other open scholarship/investigation techniques is not yet achieved.
+This can be contrasted by emerging practices like data journalism.
+One interesting example is buzzFeed's article on the use of surveillance planes operated by law enforcement and military.
+The authors were able to demonstrate based on Open Data that these purposes range from tracking drug trafficking to testing new surveillance technology {[}\protect\hyperlink{ref-peteraldhous_2016}{7}{]}.
+The credibility of the article comes from the fact that all artefacts are made available to reproduce the findings or study the approach and data transformations leading to the conclusions of the article.
+
+Arguably there are limits to the access to underlying source data.
+However, these limitations also offer an opportunity to address reproducibility {[}\protect\hyperlink{ref-koelle_open_2017}{8}{]}.
+For example, the data volume size might prohibit the further dissemination of data, its collection is too resource intensive, or there exists restrictive license agreements.
+Experience has shown that even in such cases, a limited sample of the source data can be made available for others to verify the initial data processing steps.
+For that purpose the creation of an analytical data set as the output of the data collection, gathering, and cleaning process stage can be understood as sufficient in the sense of reproducible research.
+One key transparency aspect is the documentation of the data preparation steps.
+The article Spies in the Skies by Peter Aldhous and Charles Seife {[}\protect\hyperlink{ref-peteraldhous_2016}{7}{]} is a good example for a reproducible analysis of the US government's airborne surveillance program.
+Though the license restrictions did not allow for the publication of the underlying aircraft position data, the authors made available the analytical data set and documented their analysis from the data preparation through to the publication stage.
+
+Reproducibility is discussed in various domains. The following examples further promote the use of open source software and/or data:
 
@@
 \item
-  first fully reproducible ANS Performance related paper: The paper is
-  written in RMarkdown allowing for the analysis of how text, code
-  chunks, and visualisations are produced (including associated
-  parameter settings) on the basis of the ``source text''.
-\item
-  transparent data analyses and result generation through supporting
-  web-page: All data used in this paper, intermediate data artifacts,
-  and results including the associated transformation are publicly
-  available for inspection, validation, or reproduction.
-\item
-  use-case analysis of ANS Performance during the arrival phase for 2
-  European airports: As an example the classical `additional ASMA time'
-  performance indicator is implemented on the basis of the provided data
-  with showcasing trajectory derived capabilities to further study the
-  efficiency of the arrival phase (e.g.~impact of sequencing strategies
-  {[}stack holdings, point merge{]}).
+  archaeology: Marwick et al.~(2017) focusses on computational reproducibility in archaeological research by providing basic principles.
+The case study of their implementation uses open source software (R/RMarkdown/git approach) {[}\protect\hyperlink{ref-marwick_2017}{9}{]}.
+The paper advocates the publishing (analytical) data in support of the paper.
+This should not be an issue, even if there are constraints on the overall study data.
+\item
+  engineering / signal processing: Vandewalle {[}\protect\hyperlink{ref-vandewalle_2009}{10}{]} discuss the motivation and approach to reproducibility in signal processing (i.e.~what, why, and how).
+Also here the case is made to open access to the code and data for validation purposes .
+\item
+  operational ANS Performance monitoring: Spinielli et al.~{[}\protect\hyperlink{ref-spinielli_2018}{5}{]} highlight the development and use of Open Flight Trajectories for reproducible ANS .
 \end{itemize}
 
-\section{Background -
-Reproducibility}\label{background---reproducibility}
-
-Conceptually, reproducibility is not a new topic. It is as old as the
-scientific method and at the heart of the verification/validation of
-results. The consistent and robust interpretation of results by
-independent researchers/analysts forms the evidence and the inherent
-body of knowledge in support of the applied methods or conclusions. This
-principle is also relevant for the further development on the basis of
-such results.
-
-The terms replication and reproducibility are often used interchangeably
-and there is a debate about the necessary and sufficient conditions for
-these terms. Results are considered replicable, ``if there is sufficient
-information available for independent researchers to make the same
-{[}consistent{]} findings using the same procedures with new data''
-({[}\protect\hyperlink{ref-gandrud_2015}{6}{]}, p.~3f). Replicability,
-thus, does not aim to produce the ``same'' (aka identical) results. Here
-the goal is to establish confirmation of the approach and findings by
-deriving same conclusions.\footnote{For example, the same data
-  collection procedures may not necessarily result in ``identical''
-  results as the time horizon might be different, (new or different)
-  data for another study case show different result values, but the
-  conclusions are consistent with the replicated approach.} Accordingly,
-results are reproducible, if an independent researcher can recreate the
-findings based on the given information. In our domain (i.e.~operational
-ANS Performance and associated data analysis) this will require the
-access to data and derived summaries, how associated visualisations have
-been generated, the accompanying explanatory text, and conclusions drawn
-for the analysis.
-
-Reproducibility has also to be seen in light of today's ``publish or
-perish'' context. It appears that higher emphasis is given to the
-frequency of publication than the validity of the produced results.
-Within the medical world cases like the Duke ovarian cancer scandal
-{[}\protect\hyperlink{ref-kolata_2011}{1}{]} are not single incidents,
-but also demonstrate how errors during the data analysis - in this case
-during the data preparation phase using spread sheets - can discredit
-the work. Despite the increasing awareness, the adaption of
-reproducibility or moving to other open scholarship/investigation
-techniques is not yet achieved. This can be contrasted by emerging
-practices like data journalism. One interesting example is buzzFeed's
-article on the use of surveillance planes operated by law enforcement
-and military for a variety of purposes. The authors were able to
-demonstrate based on Open Data that these purposes range from tracking
-drug trafficking to testing new surveillance technology
-{[}\protect\hyperlink{ref-peteraldhous_2016}{7}{]}. The credibility of
-the article comes from the fact that all artifacts are made available to
-reproduce the findings or study the chosen approach and data
-transformations made leading to the conclusions of the article.
-
-Arguably there are limits to the access to underlying source data.
-However, these limitations also offer an opportunity to address
-reproducibility {[}\protect\hyperlink{ref-koelle_open_2017}{8}{]}. For
-example, the data volume size might prohibit the further dissemination
-of data, its collection is too resource intensive, or there exists
-restrictive license agreements. Experience has shown that even in such
-cases, a limited sample of the source data can be made available for
-others to verify the initial data processing steps. For that purpose the
-creation of an analytical data set as the output of the data collection,
-gathering, and cleaning process stage can be understood as sufficient in
-the sense of reproducible research. One key transparency aspect is the
-documentation of the data preparation steps. The article Spies in the
-Skies by Peter Aldhous and Charles Seife
-{[}\protect\hyperlink{ref-peteraldhous_2016}{7}{]} is a good example for
-a reproducible analysis of the US government's airborne surveillance
-program. Though the license restrictions did not allow for the
-publication of the underlying aircraft position data, the authors made
-available the analytical data set and documented their analysis from the
-data preparation through to the publication stage.
-
-The reproducibility discussion focuses strongly on the recreation of
-results by other researchers/analysts. A related problem is the
-re-generating of already completed (and potentially published) results
-with the - previously - available same input data. This is a fundamental
-issue that can be summarised as ``can a future me re-do what I had done
-in the past''? This problem space ranges across the complete data
-analysis life-cyle, including archiving the orignially used input data
-or the steps to generate the study data, documenting the data cleaning
-and transformation steps including the parameters chosen, etc. By
-definition such a capability is a precondition for reproduction or
-replication by others. This implies that the analytical steps including
-the steps to produce summaries or estimates, visualisations, and
-associated conclusions are documented and archived. In order to enable
-others to recreate the own findings, such steps and documentation would
-need to be shareable. This consideration has a strong bearing on the
-toolset chosen for this paper.
-
-Some of the reasons why researchers/analysts do not publish data and
-computational procedures that allow reproduction will be hard to change
-{[}\protect\hyperlink{ref-daniel_nust_2018_2361304}{9}{]}. The
-predominant technical reasons have been listed above. However, there are
-also personal reasons such as the fear for floating ideas or loss of
-competitive advantage. Other reasons deal with more technical problems,
-e.g.~lack of capabilities, procedures.
-
-Another incentive for the adoption of an open and reproducible research
-practice can be seen in the wider recognition through increased
-citations, media attention, discussion with the community and
-collaborators, and ultimately funding opportunities. The recent years
-have seen a strong use of research blogs and community platforms
-supporting the publication of smaller steps or intermediate results. The
-combination of web presentation and discussions offers a good basis for
-an initial dialogue. Platforms like google scholar or researchgate.net
-offer additional features for outreach and communication. This equally
-applies for paper achives such as arxiv.org for pre-prints. Based on the
-credibility crisis discussion a number of journals are now promoting -
-or even requiring - the provision of analytical data in support of the
-paper publication. There is also an increasing number of requirements
-for national or European funded research with respect to data management
-and Open Data access to the research.
-
-Reproducibility is discussed in various domains. The following examples
-further promote the use of open source software and/or data:
+\begin{figure}[hbt]
+
+{\centering \includegraphics[width=1\linewidth]{figures/reproducible-research-spectrum}
+
+}
+
+\caption{Reproducible research, Fig. 2 in {[}\protect\hyperlink{ref-marwick_2017}{9}{]}.}\label{fig:reproducible-research}
+\end{figure}
+
+\hypertarget{reproducibility-spectrum}{%
+\subsection{Reproducibility Spectrum}\label{reproducibility-spectrum}}
+
+Across the literature, researchers have identified a ``reproducibility spectrum'' ranging from the extreme, i.e.~publication only (advertising and marketing of the done work), to almost or even fully replicable research.
+The work of Marwick et al.~{[}\protect\hyperlink{ref-marwick_2017}{9}{]} and Peng {[}\protect\hyperlink{ref-Peng1226}{11}{]} uses the analogy of a spectrum that can be mapped against the scoring of Vandewalle {[}\protect\hyperlink{ref-vandewalle_2009}{10}{]} describing the level of reproducibility on a 0 to 5 scale {[}\protect\hyperlink{ref-marwick_2017}{9}{]}.
+In light of {[}\protect\hyperlink{ref-leek_2015}{12}{]}, this paper conceptualises reproducibility as the ``ability to recompute data analytic results given an observed dataset and knowledge of the data analysis pipeline'' and accordingly ``replicability of a study is the chance that an independent experiment targeting the same scientific question will produce a consistent result''.
+
+With Fig. \ref{fig:reproducible-research} and the discussion in this section, this paper implements a fully reproducible workflow.
+This is achieved by establishing this paper with open source software which allows to combine text, code, and visualisations.
+The associated source text can also easily be read with a simple text editor/viewer. From that perspective, the text is fully inspectable.
+This includes the majority of visualisations which are generated based on the data.
+Accordingly, the parameters and settings chosen can be validated by an independent researcher/analyst.
+Last but not least, the data for this paper is made available in conjunction with this text as Open Data.
+This also includes intermediate data artefacts which allows for the reproduction of the complete work or portions of it.
+The conceptual approach and its building blocks will be presented in more detail in the next section.
+
+\hypertarget{method-approach}{%
+\section{Method / Approach}\label{method-approach}}
+
+\hypertarget{conceptual-approach}{%
+\subsection{Conceptual Approach}\label{conceptual-approach}}
+
+\begin{figure}[hbt]
+
+{\centering \includegraphics[width=1\linewidth]{figures/approach-stages}
+
+}
+
+\caption{Conceptual approach to reproducibility by PRU.}\label{fig:concept-approach}
+\end{figure}
+
+This paper builds on the conceptual approach presented in Fig. \ref{fig:concept-approach} to implement replicable data analyses for performance monitoring and associated research.
+On the data level this approach abstracts three levels of data:
 
@@
 \item
-  archaeology: Marwick et al. (2017) focusses on computational
-  reproducibility in archaeological research by providing basic
-  principles. The case study of their implementation uses open source
-  software (R/Rmarkdown/git approach)
-  {[}\protect\hyperlink{ref-marwick_2017}{10}{]}. The paper advocates
-  the publishing (analytical) data in support of the paper. This should
-  not be an issue, even if there are constraints on the overall study
-  data.
-\item
-  engineering / signal processing: Vandewalle (2009) discuss the
-  motivation and approach to reproducibility in signal processing
-  (i.e.~what, why, and how). Also here the case is made to open access
-  to the code and data for validation purposes
-  {[}\protect\hyperlink{ref-vandewalle_2009}{11}{]}.
-\item
-  operational ANS Performance monitoring: Spinielli et al. (2018)
-  highlight the development and use of Open Flight Trajectories for
-  reproducible ANS {[}\protect\hyperlink{ref-spinielli_2018}{5}{]}.
-\end{itemize}
-
-\begin{figure}[hbt]
-
-{\centering \includegraphics[width=1\linewidth]{figures/reproducible-research-spectrum}
-
-}
-
-\caption{Reproducible research, Fig. 2 in
-{[}\protect\hyperlink{ref-marwick_2017}{10}{]}.}\label{fig:reproducible-research}
-\end{figure}
-
-Across the literature, researchers have identified a ``reproducibility
-spectrum'' ranging from the extreme, i.e.~publication only ( advertising
-and marketing of the done work), to almost or even fully replicable
-research. The work of Marwick et al. (2017)
-{[}\protect\hyperlink{ref-marwick_2017}{10}{]} and Peng (2011)
-{[}\protect\hyperlink{ref-Peng1226}{12}{]} uses the analogy of a
-spectrum that can be mapped against the scoring of Vandewalle (2009)
-describing the level of reproducibility on a 0 to 5 scale
-{[}\protect\hyperlink{ref-marwick_2017}{10}{]}. In light of
-{[}\protect\hyperlink{ref-leek_2015}{13}{]}, this paper conceptualises
-reproducibility as the ``ability to recompute data analytic results
-given an observed dataset and knowledge of the data analysis pipeline''
-and accordingly ``replicability of a study is the chance that an
-independent experiment targeting the same scientific question will
-produce a consistent result''.
-
-With Fig. \ref{fig:reproducible-research} and the discussion in this
-section, this paper implements a fully reproducible workflow. This is
-achieved by establishing this paper with open source software which
-allows to combine text, code, and visualisations. The associated source
-text can also easily be read with a simple text editor/viewer. From that
-perspective, the text is fully inspectable. This includes the majority
-of visualisations which are generated based on the data. Accordingly,
-the parameters and settings chosen can be validated by an independent
-researcher/analyst. Last but not least, the data for this paper is made
-available in conjunction with this text as Open Data. This also includes
-intermediate data artifacts which allows for the reproduction of the
-complete work or portions of it. The conceptual approach and its
-building blocks will be presented in more detail in the next section.
-
-\section{Method / Approach}\label{method-approach}
-
-\subsection{Conceptual Approach}\label{conceptual-approach}
-
-\begin{figure}[hbt]
-
-{\centering \includegraphics[width=1\linewidth]{figures/approach-stages}
-
-}
-
-\caption{Conceptual approach to reproducibility by PRU.}\label{fig:concept-approach}
-\end{figure}
-
-This paper builds on the conceptual approach presented in Fig.
-\ref{fig:concept-approach} to implement replicable data analyses for
-performance monitoring and associated research. On the data level this
-approach abstracts three levels of data:
-
-\begin{itemize}
-\tightlist
-\item
   source data
@@
 
-The source data for this paper comprises the PRU Performance Reference
-Trajectory and operator reported data under the airport operator data
-flow. This study also uses some aeronautical information for the use
-case airports accessible through AIP or community maintained open data
-archives (e.g.~http://openaip.net, http://ourairports.com). These data
-flows are processed and stored in the EUROCONTROL data Warehouse. The
-approach recognises that the full source data may or cannot be provided
-for a variety of reasons. These include volume of the source data sets,
-technical constraints allowing access to the source data storage, and
-data sharing policies or use limitations. For the latter cases, PRU
-promotes the sharing of a so-called representative sample. The idea is
-that with the provision of the associated processing code, other
-researchers/practitioners are able to reconstruct a part of the analytic
-data - and subsequently part of the reported results - on the basis of
-the provided source data set.
-
-Analytical data refers to the data sets established based on the
-aforementioned processing code. An essential defining characteristic of
-analytic data is that it is the data basis on which the reported results
-can be produced. The abstraction between source data and analytic data
-allows for the sharing of such data as limitations are typically only
-appliable to the source data. We follow the notion of
-{[}\protect\hyperlink{ref-marwick_2017}{10}{]} by referring to the
-provision of a representative sample of 8 days accompanying this paper
-as source data.
-
-As depicted in Fig. \ref{fig:reproducible-research} the provision of the
-data, code, and text allows to move to the right part of the
-reproducibility spectrum, away from marketing and final results to
-credible, transparent, and challengeable analysis. Accordingly, the
-results (or performance related data) level presents the final
-abstraction.
-
-\subsection{Toolset}\label{toolset}
-
-This work builds on the RStudio tools for the R language
-{[}\protect\hyperlink{ref-rcoreteam_2018}{14}{]} including Git (and the
-web-based repository managers GitHub and GitLab) as underlying version
-control system. The R language was originally developed within the
-statistical community supporting the task of statistical reporting by
-providing routines for the statistical computing and visualisation.
-Being open source, the R community is actively engaging and sharing
-related software packages to enhance the core functionality. Without
-limiting the impact of other packages, the development of knitr
-{[}\protect\hyperlink{ref-R-knitr}{15}{]} and RMarkdown
-{[}\protect\hyperlink{ref-xie2018}{16}{]}, ggplot
-{[}\protect\hyperlink{ref-wickham_2016}{17}{]} for visualisation, and
-the so-called tidyverse packages and RStudio IDE
-{[}\protect\hyperlink{ref-rstudioteam_2015}{18}{]} provide the basis for
-an open source data analysis ecosystem.
-
-A key feature for the implementation of the reproducibility paradigm is
-the fact that RMarkdown supports the combination of text, coding, and
-visualisations in a single document. The latter can be rendered into
-various output formats, e.g.~classical paper report-styles (pdf, MS
-Word), html web-pages, and slides (both classical Powerpoint style or
-html-based presentations). The fact that outputs can be rendered to html
-pages also supports the embedding of interactive graphics. This also
-facilitates the communication and outreach amongst researchers/analysts
-as standard web technology becomes a standard outlet format. It follows
-that changes to the code (and associated visualisations) or text can be
-readily updated by re-rendering the adapted report. This allows to
-develop standard templates for recurring reports based on varying input
-data, but it also serves the iterative development of larger reports
-(and studies).
-
-A minimal example of RMarkdown document, a plain-text file with the
-conventional extension \texttt{.Rmd} is shown below:
+The source data for this paper comprises the PRU Performance Reference Trajectory and operator reported data under the airport operator data flow.
+This study also uses some aeronautical information for the use case airports accessible through AIP or community maintained open data archives (e.g.~http://openaip.net, http://ourairports.com).
+These data flows are processed and stored in the EUROCONTROL data warehouse.
+The approach recognises that the full source data may or cannot be provided for a variety of reasons.
+These include volume of the source data sets, technical constraints allowing access to the source data storage, and data sharing policies or use limitations.
+For the latter cases, PRU promotes the sharing of a so-called representative sample.
+The idea is that with the provision of the associated processing code, other researchers/practitioners are able to reconstruct a part of the analytic data - and subsequently part of the reported results - on the basis of the provided source data set.
+
+Analytical data refers to the data sets established based on the aforementioned processing code.
+An essential defining characteristic of analytic data is that it is the data basis on which the reported results can be produced.
+The abstraction between source data and analytic data allows for the sharing of such data as limitations are typically only applicable to the source data.
+We follow the notion of {[}\protect\hyperlink{ref-marwick_2017}{9}{]} by referring to the provision of a representative sample of 8 days accompanying this paper as source data.
+
+As depicted in Fig.
+\ref{fig:reproducible-research} the provision of the data, code, and text allows to move to the right part of the reproducibility spectrum, away from marketing and final results to credible, transparent, and challengeable analysis.
+Accordingly, the results (or performance related data) level presents the final abstraction.
+
+\hypertarget{toolset}{%
+\subsection{Toolset}\label{toolset}}
+
+This work builds on the RStudio tools for the R language {[}\protect\hyperlink{ref-rcoreteam_2018}{13}{]} including Git (and the web-based repository managers GitHub and GitLab) as underlying version control system.
+The R language was originally developed within the statistical community supporting the task of statistical reporting by providing routines for the statistical computing and visualisation.
+Being open source, the R community is actively engaging and sharing related software packages to enhance the core functionality.
+Without limiting the impact of other packages, the development of knitr {[}\protect\hyperlink{ref-R-knitr}{14}{]} and RMarkdown {[}\protect\hyperlink{ref-xie2018}{15}{]}, ggplot {[}\protect\hyperlink{ref-wickham_2016}{16}{]} for visualisation, and the so-called tidyverse packages and RStudio IDE {[}\protect\hyperlink{ref-rstudioteam_2015}{17}{]} represent an open source ecosystem for data analysis.
+A key feature for the implementation of the reproducibility paradigm is the fact that RMarkdown supports the combination of text, analytical code, and visualisations in a single document.
+A minimal example of an RMarkdown document, a plain-text file with the conventional
+extension \texttt{.Rmd} is shown below:
 
@@
 
-\NormalTok{```\{r\}}
-\NormalTok{library(ggplot2)}
-\NormalTok{fit = lm(mpg ~ wt, data = mtcars)}
-\NormalTok{b   = coef(fit)}
-\NormalTok{ggplot(mtcars, aes(x = wt, y = mpg)) +}
-\NormalTok{  geom_point() + }
-\NormalTok{  geom_smooth(method = lm, se = TRUE)}
-\NormalTok{```}
+\BaseNTok{```\{r\}}
+\BaseNTok{library(ggplot2)}
+\BaseNTok{fit = lm(mpg ~ wt, data = mtcars)}
+\BaseNTok{b   = coef(fit)}
+\BaseNTok{ggplot(mtcars, aes(x = wt, y = mpg)) +}
+\BaseNTok{  geom_point() + }
+\BaseNTok{  geom_smooth(method = lm, se = TRUE)}
+\BaseNTok{```}
 
 \NormalTok{The slope of the regression is }\BaseNTok{`r b[1]`}\NormalTok{.}
-\NormalTok{```}
+\BaseNTok{```}
 \end{Highlighting}
@@
 
+There are three components of an R Markdown document: the metadata, the text and the code.
+The metadata, written between the pair of three dashes \texttt{-\/-\/-}, helps in the production of
+output document (title, type of document {[}web page, PDF, MS Word{]}, \ldots)
+The text, which follows the metadata, is written using Markdown, a simplified formatting syntax
+which can include prose and code.
+There are two types of (computer) code:
+
+\begin{itemize}
+\item
+  A code chunk starts with three backticks and \texttt{r} indicates the
+  language name,\footnote{R Markdown supports 50 other computer languages.} and ends with
+  three backticks.
+\item
+  An inline R code expression starts with backtick r and ends with a backtick.
+\end{itemize}
+
+Code portions are executed when the document is rendered.
+Fig. \ref{fig:rmd-example} shows how the simple R Markdown example is rendered.
+
 \begin{figure}[htb]
@@
 
-There are three components of an R Markdown document: the metadata, the
-text and the code. The metadata, written between the pair of three
-dashes \texttt{-\/-\/-}, helps in the production of output document
-(title, type of document {[}web page, PDF, MS Word{]}, \ldots{}) The
-text, which follows the metadata, is witten using Markdown, a simplified
-formatting syntax. There are two types of (computer) code:
-
-\begin{itemize}
-\item
-  A code chunk starts with three backticks and \texttt{r} indicates the
-  language name,\footnote{R Markdown supports 50 other languages.} and
-  ends with three backticks.
-\item
-  An inline R code expression starts with backtick r and ends with a
-  backtick.
-\end{itemize}
-
-Code portions are executed when the document is rendered.
-
-The result of rendering the R Markdown above is shown in Fig.
-\ref{fig:rmd-example}.
-
-A particular strong point is that the R/Rstudio ecosystem is open source
-software. This includes also its extensibility with other packages.
-There exists a wide variety of packages for data analytical tasks and
-techniques. This mechanism can also be useful for the communication of
-results. Researchers/analysts are able to provide their code and
-supporting documentation as a package for other researchers to build on.
-For example, throughout the last year, PRU has developed certain utility
-functions for reoccurring tasks of the data analytical pipeline. It is
-now planned to wrap these functions into a package saving other
-researchers/analysts of recoding certain steps. For this paper, the
-functions and data transformations are reported as part of the RMarkdown
-and supporting R files. All these files are available via the project
-webpage.
-
-\subsection{Capabilities: Performance Reference
-Trajectory}\label{capabilities-performance-reference-trajectory}
-
-Throughout the recent year PRU developed the data analytical
-capabilities to fuse correlated position reports and ADSB surveillance
-data. This work has been reported in
-{[}\protect\hyperlink{ref-koelle_open_2017}{8}{]} and
-{[}\protect\hyperlink{ref-spinielli_2018}{5}{]}. The idea is to
-establish a pan-European air situation picture openly available to
-research and practitioners. Conceptually, the fused performance
-reference trajectory will be a key milestone in combatting marketing
-myths. For example, operational improvements can be directly assessed by
-interested analysts with a view to validate the observable performance
-improvements. This allows for the application of commonly accepted
-methods and models to express performance benefits and - ultimately -
-challenge related publications on deployments and achieved improvements.
-
-End of 2018 the development project was closed and PRU is now
-investigating options for deploying and maintaining the processing
-modules on a cloud platform. The goal is to establish and operate the
-underlying data processing capabilities throughout 2019 and launch the
-regular provision of the performance reference trajectory as Open Data.
-It is also planned to establish a network and complement the air
-situation picture by deploying additional sensors or determining
-non-ADSB derived positions as a community effort. As a mid-term goal
-associated data bases and models for performance shall be established
-and jointly maintained.
-
-\section{Results}\label{results}
-
-\subsection{Data Preparation}\label{data-preparation}
-
-\begin{figure}[htb]
-
-{\centering \includegraphics[width=1\linewidth]{figures/data-preparation-pipeline}
-
-}
-
-\caption{Data preparation pipeline.}\label{fig:data-preparation-pipeline}
-\end{figure}
-
-The raw data sources for the use cases of this paper have been processed
-by a pipeline of R scripts. For example, the additional ASMA time
-indicator and the anticipated further breakdown of the arrival phase can
-be described by a set of tranistion points:
+\hypertarget{capabilities-performance-reference-trajectory}{%
+\subsection{Capabilities: Performance Reference Trajectory}\label{capabilities-performance-reference-trajectory}}
+
+Throughout the recent year PRU developed the data analytical capabilities to fuse correlated position reports and ADSB surveillance data.
+This work has been reported in {[}\protect\hyperlink{ref-koelle_open_2017}{8}{]} and {[}\protect\hyperlink{ref-spinielli_2018}{5}{]}.
+The idea is to establish a pan-European air situation picture openly available to researchers and practitioners.
+Conceptually, the performance reference trajectory will be a key milestone in combatting the ``marketing myths.''
+For example, operational improvements can be directly assessed by interested analysts as access to the Open Data enables the validation of the observable performance improvements.
+This allows for the application of commonly accepted methods and models to express performance benefits and ultimately challenge related publications on deployments and achieved improvements.
+
+At the end of 2018 the initial development project phase was closed and PRU is now investigating options for deploying and maintaining the processing modules on a cloud platform.
+The goal is to establish and operate the underlying data processing capabilities throughout 2019 and launch the regular provision of the performance reference trajectory as Open Data in 2020.
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\hypertarget{data-preparation}{%
+\subsection{Data Preparation}\label{data-preparation}}
+
+The raw data sources for the use cases of this paper have been processed by a pipeline of R scripts.
+For example, the additional ASMA time indicator and the anticipated further breakdown of the arrival phase can be described by a set of transition points:
 
@@
 \item
-  segment I: entry into the ASMA area to entry into the
-  holding/vectoring segment
-\item
-  segment II: holding/vectoring segment; for the use case reported
-  holding stacks or point merge procedure
+  segment I: entry into the ASMA area to entry into the holding/vectoring segment
+\item
+  segment II: holding/vectoring segment; for the use case reported holding stacks or point merge procedure
 \item
@@
 
-Figure \ref{fig:data-preparation-pipeline} shows how the transition
-points (40 NM, entry/exit from holding/point-merge area, landing) for
-the use case airports have been extracted. Due to the size of the daily
-reference trajectory files, the pipeline operates on one date at the
-time. Assuming the relevant RT flight and position files are downloaded
-from GCP as described in
-{[}\protect\hyperlink{ref-performancereviewunit_2019}{19}{]}, the steps
-and the relevant scripts (available in the companion repository for this
-paper) used are:
+\begin{figure}[htb]
+
+{\centering \includegraphics[width=1\linewidth]{figures/data-preparation-pipeline}
+
+}
+
+\caption{Data preparation pipeline.}\label{fig:data-preparation-pipeline}
+\end{figure}
+
+Figure \ref{fig:data-preparation-pipeline} shows how the transition points
+(40 NM, entry/exit from holding/point-merge area, landing) for the use case airports
+have been extracted.
+The following steps and scripts comprise {[}\protect\hyperlink{ref-performancereviewunit_2019}{18}{]}:
 
@@
 \item
-  the script \texttt{R/extract-arrivals.R} extracts arrivals to the
-  \texttt{APT} airport and generates:
+  the script \texttt{R/extract-arrivals.R} extracts arrivals to the \texttt{APT} airport and generates files with:
 
@@
   \item
-    a file with flight details
-  \item
-    a file with all the arrivals' positions
-  \item
-    a file with all arrivals' positions within 40 NM the airport's
-    reference position (ARP)
+    flight details
+  \item
+    arrivals' positions (\texttt{raw} and smoothed trajectories)
+  \item
+    (smoothed) arrivals' positions within 40 NM of the airport's reference position (ARP)
   \end{itemize}
 \item
-  the script \texttt{R/extract-transition-points.R} extracts transition
-  points in the terminal area and generates a file containing per each
-  flight:
+  the script \texttt{R/extract-transition-points.R} extracts transition points in the terminal area
+  and generates a file containing per each flight:
 
@@
   \item
-    the first and last point within the holding/point-merge area,
-    \texttt{PHOLD}
+    the first and last point within the holding/point-merge area, \texttt{PHOLD}
   \item
@@
 
-\begin{figure}[hbt]
-
-{\centering \includegraphics[width=1\linewidth]{figures/eidw-arrival-chart-east}
-
-}
-
-\caption{Excerpt of Irish AIP showing point merge
-arc layout East of Dublin Airport.}\label{fig:eidw-arrival-chart-east}
-\end{figure}
-
-This stage makes use of data extracted/extrapolated from the relevant
-AIP, see for example Fig. \ref{fig:eidw-arrival-chart-east}, such as
-ARP, runway name and threshold position/elevation, polygons for holding
-stacks or point-merge areas. Specifically the polygons have been
-heuristically defined and enlarged by 2NM in order to cater for the
-spread of real flown trajectories, see the scripts
-\texttt{R/egll-data.R} and \texttt{R/eidw-data.R}.
-
-Each position mentioned above consist of at least
+This stage makes use of data extracted/extrapolated from the relevant AIP such
+as ARP, runway name and threshold position/elevation, polygons for holding
+stacks or point-merge areas. Specifically the polygons have been heuristically defined from the
+point merge points in the AIP and enlarged by 2NM in order to cater for the spread of real flown
+trajectories,
+see the scripts \texttt{R/egll-data.R} and \texttt{R/eidw-data.R}.
+
+Each transition position mentioned above consist of at least
 
@@
 \item
-  flight ID
-\item
-  4D position (timestamp \(T\), longitude, latitude, altitude)
-\item
-  flown distance \(D\)
-\item
-  sequence number
-\item
-  distance from ARP
+  aircraft/flight info: ID, callsign, registration, ADEP/ADES, ICAO 24 bit address;
+\item
+  4D position: timestamp \(T\), longitude, latitude, altitude;
+\item
+  position info: flown distance \(D\) from aerodrome of departure (ADEP), sequence number,
+  distance to ARP as a proxy of aerodrome of destination (ADES);
+\item
+  point info: sequence number, type of transition point {[}\texttt{P40}, \texttt{PHOLD}, \texttt{PLAND}{]}
 \end{itemize}
@@
 \item
-  the time spent before the holding/point-merge area,
-  \(T_{PHOLD|first} - T_{P40}\)
+  the time spent before the holding/point-merge area, \(T_{PHOLD|first} - T_{P40}\)
 \item
@@
 \item
-  the time spent within the holding/point-merge area,
-  \(T_{PHOLD | last} - T_{PHOLD|first}\)
+  the time spent within the holding/point-merge area, \(T_{PHOLD | last} - T_{PHOLD|first}\)
 \item
@@
 
-\subsection{Data Coverage}\label{data-coverage}
-
-The previous sections described the data preparatory action. For this
-paper, a reproducible data set for the analysis of arrival management
-techniques at London Heathrow and Dublin airport have been extracted.
-The data set comprises flights for the period of 1.-8. August 2017. It
-can be inferred from the traffic mix at London Heathrow (major
-international hub) that only a limited subset of flights is not equipped
-with ADSB, while a slightly higher level of non-equipped flights is
-expected at Dublin (higher share of intra-European flights).
+\hypertarget{data-coverage}{%
+\subsection{Data Coverage}\label{data-coverage}}
+
+For this paper, a reproducible data set for the analysis of arrival management techniques at London Heathrow and Dublin airport have been extracted.
+The data set comprises flights for the period of 1.-8. August 2017.
+It can be inferred from the traffic mix at London Heathrow (major international hub) that only a limited subset of flights is not equipped with ADSB, while a slightly higher level of non-equipped flights is expected at Dublin (higher share of intra-European flights).
 
@@
 
-\caption{Number of arrivals in data sets, i.e.~1.-8. July
-2017.}\label{fig:coverage-egll}
-\end{figure}
-
-Fig. \ref{fig:coverage-egll} shows the overlap of the data between the
-arrivals extracted from the airport operator data flow (APDF) and the
-reference trajectory (RTRJ) for London Heathrow (EGLL) and Dublin
-(EIDW).
-
-For the time horizon of this paper, the coverage of the reference
-trajectory derived data is for Heathrow (EGLL) 100\% and for Dublin
-97\%.
-
-\subsection{\texorpdfstring{``replicate'' ASMA
-(time-based)}{replicate ASMA (time-based)}}\label{replicate-asma-time-based}
-
-The additional time in the arrival sequencing and metering area (ASMA)
-is a globally recognised measure to assess flight efficiency during the
-arrival phase. The ASMA metric calculates the additional time based on a
-reference time which is determined for a family of flights sharing
-similar characteristics. These are the entry sector into the arrival
-airspace, weight turbulence category and engine type, and the actual
-landing runway. For each flight \(i\) the additional ASMA time is
+\caption{Number of arrivals in data sets, i.e.~1.-8. Aug 2017.}\label{fig:coverage-egll}
+\end{figure}
+
+Fig.
+\ref{fig:coverage-egll} shows the overlap of the data between the arrivals extracted from the airport operator data flow (APDF) and the reference trajectory (RTRJ) for London Heathrow (EGLL) and Dublin (EIDW).
+For the time horizon of this paper, the coverage of the reference trajectory derived data is for Heathrow (EGLL) 100\% and for Dublin 97\%.
+
+\hypertarget{replicate-asma-time-based}{%
+\subsection{``replicate'' ASMA (time-based)}\label{replicate-asma-time-based}}
+
+The additional time in the arrival sequencing and metering area (ASMA) is a globally recognised measure to assess flight efficiency during the arrival phase.
+The ASMA metric calculates the additional time based on a reference time which is determined for a family of flights sharing similar characteristics.
+For each flight \(i\) the additional ASMA time is
 
@@
 
-The reference time \(i\) is determined for flights with similar arrival
-characteristics (c.f.
+The reference time \(i\) is determined for flights with similar arrival characteristics (c.f.
 \url{http://ansperformance.eu/references/definition/additional_asma_time.html}).
-These include 1.) the arrival entry sector at 40NM from the airport, 2.)
-the aircraft weigth turbulence category and engine type, and 3.) the
-actual landing runway. The reference times are built for a longer time
-horizon and extracted from the PRU data base as additional source data.
+These include 1.) the arrival entry sector at 40NM from the airport, 2.) the aircraft weight turbulence category and engine type, and 3.) the actual landing runway.
+The reference times are built for a longer time horizon and extracted from the PRU data base as additional source data.
 
@@
 
-\caption{Distribution of actual ASMA travel times for EGLL
-(1.-8. July 2017).}\label{fig:ASMA-travel-egll}
-\end{figure}
-
-As an example, the distribution of the actual arrival transit times for
-arrivals into Heathrow (EGLL) are shown in Fig.
-\ref{fig:ASMA-travel-egll}. Heathrow operates holding stacks to
-tactically optimise the arrival throughput. Dependent on the traffic
-situation, aircraft may experience longer holding times (i.e.~long tails
-of boxplots in Fig. \ref{fig:ASMA-travel-egll}).
-
-Fig. \ref{fig:ASMA-compare} compares the additional ASMA time extracted
-from the PRU performance dashboard, the trajectory based calculation
-with the dashboard derived reference time, and a 20th percentile
-approach purely based on the trajectory data.
+\caption{Distribution of actual ASMA travel times for EGLL (1.-8. Aug 2017).}\label{fig:ASMA-travel-egll}
+\end{figure}
+
+As an example, the distribution of the actual arrival transit times for arrivals into Heathrow (EGLL) are shown in Fig.
+\ref{fig:ASMA-travel-egll}.
+Heathrow operates holding stacks to tactically optimise the arrival throughput.
+Dependent on the traffic situation, aircraft may experience longer holding times (i.e.~long tails of box plots in Fig. \ref{fig:ASMA-travel-egll}).
+
+Fig.
+\ref{fig:ASMA-compare} compares the additional ASMA time extracted from the PRU performance dashboard, the trajectory based calculation with the dashboard derived reference time, and a 20-th percentile approach purely based on the trajectory data.
 
@@
 
-\caption{Comparison of determined ASMA values, i.e.~1.-8. July
-2017.}\label{fig:ASMA-compare}
-\end{figure}
-
-Fig. \ref{fig:ASMA-compare} shows the variation of the timebased
-additional ASMA times determined for our sample. APDFNM refers to the
-variant used for the EUROCONTROL Performance Review System. The
-reference time is build on the basis of a one-year sammple. Applying
-this reference time to the trajectory based travel times (i.e.~TRJ-REF),
-there is a strong increase of the additional ASMA time for Heathrow
-(EGLL), while in the case of Dublin (EIDW) reasonable decrease takes
-place. Calculating the additional ASMA time exclusively with the
-trajectory sample and applying a 20th percentile approach (TRJ-20PCT),
-shows a significant drop in EGLL and a moderate increase at Dublin. The
-high variation for Heathrow shows that the additional ASMA time is
-strongly dependent on the time horizon. The pure trajectory bases
-measure - by definition - includes the effect of the operational concept
-(stack holdings) and distorts the measure. In the case of Dublin it can
-be argued that the TRJ-20PCT measure reflects the time based ASMA
-measure suffiently.
-
-\subsection{Distance-based ASMA}\label{distance-based-asma}
-
-conceptually, measuring flight inefficiency in terms of flight time
-accounts for the airspace user perspective. The previous section has
-also shown the high dependency of the time based measure on the sample
-size or reference time attribution. Flight time is linked to engine time
-and accordingly, assuming maneuvering for operational separation and
-sequencing reasons, fuel burn. From a service provision perspective,
-efficiency relates to the capability to provide the least distance flown
-within the ASMA area. In that respect, the data analytical capability to
-determine an associated additional distance flown metric addresses an
-open gap.
-
-\begin{figure}[H]
+\caption{Comparison of determined ASMA values, i.e.~1.-8. Aug 2017.}\label{fig:ASMA-compare}
+\end{figure}
+
+In Fig. \ref{fig:ASMA-compare}
+APDFNM refers to the variant used for the EUROCONTROL Performance Review System.
+The reference time is build on the basis of a one-year sample.
+Applying this reference time to the trajectory based travel times (i.e.~TRJ-REF), there is a strong increase of the additional ASMA time for Heathrow (EGLL), while in the case of Dublin (EIDW) a reasonable decrease takes place.
+Calculating the additional ASMA time exclusively with the trajectory sample and applying a 20-th percentile approach (TRJ-20PCT), shows a significant drop in EGLL and a moderate increase at Dublin.
+The high variation for Heathrow shows that the additional ASMA time is strongly dependent on the time horizon.
+The pure trajectory based measure - by definition - includes the effect of the operational concept (stack holdings) and distorts the measure.
+In the case of Dublin it can be argued that the TRJ-20PCT measure reflects the time based ASMA measure sufficiently.
+
+\hypertarget{distance-based-asma}{%
+\subsection{Distance-based ASMA}\label{distance-based-asma}}
+
+Conceptually, measuring flight inefficiency in terms of flight time accounts for the airspace user perspective.
+The previous section has also shown the high dependency of the time based measure on the sample size or reference time attribution.
+Flight time is linked to engine time and accordingly, assuming manoeuvring for operational separation and sequencing reasons, fuel burn.
+From a service provision perspective, efficiency relates to the capability to provide the least distance flown within the ASMA area.
+In that respect, the data analytical capability to determine an associated additional distance flown metric addresses an open gap.
+
+\begin{figure}[hbt]
 
@@
 
-\caption{Relationship travel time and distance flown,
-i.e.~1.-8. July 2017.}\label{fig:ASMA-time-dist}
-\end{figure}
-
-Fig. \ref{fig:ASMA-time-dist} shows the relationship between the actual
-distance flown and the actual time flown. The variation of travel times
-is influenced by varying airspeeds or weather phenomena (e.g.~wind).
-This paper applies a simple 15th percentile approach to the distance
-flown per arrival sector, aircraft type/engine, and landing runway.
-
-\begin{table}[t]
-
-\caption{\label{tab:tbl-asma-dist}Average additional distance flown (15th pct)}
+\caption{Relationship travel time and distance flown, i.e.~1.-8. Aug 2017.}\label{fig:ASMA-time-dist}
+\end{figure}
+
+Fig. \ref{fig:ASMA-time-dist} shows the relationship between the actual distance flown and the actual time flown.
+The variation of travel times is influenced by varying airspeeds or weather phenomena (e.g.~wind).
+This paper applies a simple 15-th percentile approach to the distance flown per arrival sector, aircraft type/engine, and landing runway.
+Tab. \ref{tab:tbl-asma-dist} lists the determined average additional distance at the study airports.
+The total additional distance is determined as the aggregated sum of the differences in each segment
+with 15-th percentile reference distance.
+
+\begin{table}
+
+\caption{\label{tab:tbl-asma-dist}Average additional distance flown (15-th pct)}
 \centering
-\begin{tabular}{l|r|r|r}
+\begin{tabular}[t]{l|r|r|r}
 \hline
@@
 \hline
-EGLL & 117037.79 & 5249 & 22.30\\
+EGLL & 117014.9 & 5249 & 22.29\\
 \hline
-EIDW & 33079.22 & 2634 & 12.56\\
+EIDW & 33081.9 & 2634 & 12.56\\
 \hline
@@
 
-Tab. \ref{tab:tbl-asma-dist} lists the determined average additional
-distance at the study airports.
-
-\subsection{Arrival Management
-Practices}\label{arrival-management-practices}
-
-For an initial drill down, this paper uses arrival practices at London
-Heathrow and Dublin. The approach to managing the arrival flow differs
-significantly between both airports. Conceptually, London Heathrow
-employs the ``pressure on the runway'' concept with a focus on achieving
-and balancing a continuously high level of runway throughput
-utilisation. This is achieved by controlling the demand through 4
-arrival stacks that allow for the tactical sequencing on the last
-portion of the approach. Dublin is one of the airports that pioneered
-the implementation of ``point merge''. Point merge aims at replacing
-circular holdings by longitudal holdings to establish the sequence based
-on predefined legs.
-
-From a performance perspective it is therefore interesting to identify
-and measure the impact of the operational concept on the overall flight
-efficiency during the arrival phase of flight.
+\hypertarget{arrival-management-practices}{%
+\subsection{Arrival Management Practices}\label{arrival-management-practices}}
+
+For an initial drill down, this paper uses arrival practices at London Heathrow and Dublin.
+The approach to managing the arrival flow differs significantly between these airports.
+Conceptually, London Heathrow employs the ``pressure on the runway'' concept with a focus on achieving and balancing a continuously high level of runway throughput utilisation.
+This is accomplished by controlling the demand through 4 arrival stacks that allow for the tactical sequencing on the last portion of the approach.
+Instead Dublin is one of the airports that pioneered the implementation of ``point merge''.
+Point merge aims at replacing circular holdings by longitudinal holdings to establish the sequence based on predefined legs.
+
+\begin{figure}[htb]
+
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/egll-trajectory-example-1}
+
+}
+
+\caption{EGLL - long trajectory flown on 1. Aug 2017.}\label{fig:egll-trajectory-example}
+\end{figure}
+
+\begin{figure}[bht]
+
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/eidw-trajectory-example-1}
+
+}
+
+\caption{EIDW - long trajectory flown on 1. Aug 2017.}\label{fig:eidw-trajectory-example}
+\end{figure}
+
+From a performance perspective it is therefore interesting to identify and measure the impact of the operational concept on the overall flight efficiency during the arrival phase of flight.
+
+Fig. \ref{fig:egll-trajectory-example} and Fig. \ref{fig:eidw-trajectory-example} show examples
+for long distances flown on 1. Aug 2017.
+The trajectory data allows for the identification of the holdings and/or sequencing area so as
+to separate the effect of the operational concepts.
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/egll-trajectory-example-1}
-
-}
-
-\caption{EGLL - long trajectory flown on 1. July
-2017.}\label{fig:egll-trajectory-example}
-\end{figure}
+{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/distance-flown-segment-1}
+
+}
+
+\caption{Actual distances flown for each segment of the arrival phase (i.e.~EGLL, 1.-8. Aug 2017).}\label{fig:distance-flown-segment}
+\end{figure}
+
+Fig. \ref{fig:distance-flown-segment} depicts the flown distances per segment of the arrival phase.
+In the case of London Heathrow the classical stack based holding can be seen in the significant variations of the actual distance flown per arrival direction.
+Only limited sequencing takes place during the first segment (arrival within 40 NM and the stacks).
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/eidw-trajectory-example-1}
-
-}
-
-\caption{EIDW - long trajectory flown on 1. July
-2017.}\label{fig:eidw-trajectory-example}
-\end{figure}
-
-Fig. \ref{fig:egll-trajectory-example} and Fig.
-\ref{fig:eidw-trajectory-example} show examples for long distances flown
-on 1. July 2017. The trajectory data allows for the identification of
-the holdings and/or sequencing area. This allows to separate the effect
-of the operational concepts.
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/distance-flown-segment-1}
-
-}
-
-\caption{Actual distances flown for each segment of
-the arrival phase (i.e.~EGLL, 1.-8. July 2017).}\label{fig:distance-flown-segment}
-\end{figure}
-
-\begin{figure}[H]
-
 {\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/distance-flown-segment-eidw-1}
@@
 
-\caption{Actual distances flown for each
-segment of the arrival phase (i.e.~EIDW, 1.-8. July 2017).}\label{fig:distance-flown-segment-eidw}
-\end{figure}
-
-Fig. \ref{fig:distance-flown-segment} depicts the flown distances per
-segment of the arrival phase. In the case of London Heathrow the
-classical stack based holding can be seen in the significant variations
-of the actual distance flown per arrival direction. Only limited
-sequencing takes place during the first segment (arrival within 40 NM
-and the stacks). Fig. \ref{fig:distance-flown-segment-eidw} shows
-different footprint. It follows from the distribution of the initial
-approach segment that this portion is already used for sequencing and
-synchronisation by Dublin approach. The middle segment shows a wider
-variation and less uniform distribution than the approach at London
-Heathrow.
-
-\begin{figure}[H]
+\caption{Actual distances flown for each segment of the arrival phase (i.e.~EIDW, 1.-8. Aug 2017).}\label{fig:distance-flown-segment-eidw}
+\end{figure}
+
+Fig.
+\ref{fig:distance-flown-segment-eidw} shows different footprint.
+It follows from the distribution of the initial approach segment that this portion is already used for sequencing and synchronisation by Dublin approach.
+The middle segment shows a wider variation and less uniform distribution than the approach at London Heathrow.
+
+Fig. \ref{fig:eidw-saturation-example} depicts the variation of the arrival segment throughput at Dublin (EIDW).
+The median for every set of flights in the ASMA area versus flights landing is highlighted.
+The behaviour is a classical saturation curve, i.e.~up to an arrival bunching of 10-11 flights per 20 min (30-33 flights per hour), the point merge procedure seems to be able to support the tactical absorption and sequencing.
+It is noteworthy that the saturation curve falls for higher arrival throughputs.
+As no measurements for 12 through 14 landings had been made in the sample, this behaviour could be caused by the chosen time horizon.
+
+\begin{figure}[hb]
 
@@
 
-\caption{EIDW - arrival throughput saturation
-(1.-8. July 2017).}\label{fig:eidw-saturation-example}
-\end{figure}
-
-Fig. \ref{fig:eidw-saturation-example} depicts the variation of the
-arrival segment throughput at Dublin (EIDW). The median for every set of
-flights in the ASMA area versus flights landing is highlighted. The
-behavour is a classical saturation curve, i.e.~up to an arrival bunching
-of 10-11 flights per 20 min (30-33 flights per hour), the point merge
-procedure seems to be able to support the tactical absorption and
-sequencing. It is noteworthy that the saturation curve falls for higher
-arrival throughputs. As no mearurements for 12 through 14 landings had
-been made in the sample, this behaviour could be caused by the chosen
-time horizon.
-
-\section{Conclusions}\label{conclusions}
-
-The work reported establishes a fully reproducible paper for ANS
-performance in Europe. It is the one of the first attempts to address
-the reproducibility crisis within ANS performance by providing the
-artifacts that allow an independent researcher or professional to
-reproduce the results and verify upstream processing steps.
-
-This paper follows up on previous work
-{[}\protect\hyperlink{ref-spinielli_2018}{5}{]},
-{[}\protect\hyperlink{ref-koelle_open_2017}{8}{]},
-{[}\protect\hyperlink{ref-spinielli_2017}{20}{]} and forms part of the
-roll-out of the performance reference trajectory by the PRU. As such the
-paper demonstrates the feasibility to the approach, the data analytical
-capabilities and preprocessing, and the reporting via the R/RStudio
-ecosystem. This represents an essential milestone moving towards the
-envisaged target environment. However, further work is required to
-establish the reproducible environment for ANS Performance.
-
-To overcome these limitations, PRU will be looking into deploying the
-trajectory processing on a cloud-based platform including the regular
-provision of the identified analytical data levels.
-
-The move to open data for ANS Performance is a \emph{giant leap} in
-terms of transparency. This will foster a new level of cooperation and
-discussion amongst researchers and professionals interested in this
-topic. It removes the burden to acquire and preprocess large-scale
-surveillance data and associated aircraft data or aeronautical
-information. PRU envisages a strong collaboration through a network of
-research and practitioners. The departure point for this development was
-to address the lack and hurdle of verifying results. A further advantage
-is that novel applications, methods and algorithms (e.g.~machine
-learning), can be applied to a harmonised dataset. It is hoped that
-development efforts can accelerate while offering a platform for
-improvements to existing processing methods and algorithms.
-
-The major application of the paper results will inform the performance
-reference trajectory roll-out project. This goes in hand with the
-iterative move of the operational performance measures to trajectory
-derived data. This can take on the form of quality assurance measures
-(e.g.~verifying the order of magnitude, complementing missing reported
-data), but also making the reference trajectory the primary means of
-data collection. The latter will support the further development of the
-operational performance measures responding to the ICAO GANP.
-
-This paper documents a key milestone in the project. As such the future
-work will address:
-
-\begin{itemize}
-\tightlist
-\item
-  roll-out and establishment of the performance reference trajectory:
-  This includes the organisational and technical work to deploy the
-  processing modules and arrange for a regular processing of the source
-  data and its provision as Open Data.
-\item
-  establishing the current performance reporting based on the
-  trajectory: Since more than 2 years PRU addresses the need for Open
-  Data by providing monthly updates for the key performance indicators
-  used under the EUROCONTROL Performance Review System and the SES
-  Performance Scheme. Underlying analytic data is made available via
-  \url{http://ansperformance.eu}. As part of the roll-out, it is
-  foreseen to move associated indicators to a trajectory derived or at
-  least verified process and expand on the analytical data level.
-\item
-  establishment of a network: The underlying idea is that various levels
-  of future work can be orchestrated in close collaboration with
-  interested researchers and professionals. This can span work
-  addressing the improvement of the actual data processing (e.g.~less
-  resource intensive computing) to the further development and
-  validation of novel techniques for measuring operational performance
-  (e.g.~application of machine learning, time series analysis).
-\end{itemize}
-
-\section*{References}\label{references}
+\caption{EIDW - arrival throughput saturation (1.-8. Aug 2017).}\label{fig:eidw-saturation-example}
+\end{figure}
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
+
+The work reported establishes a fully reproducible paper for ANS performance in Europe.
+It is the one of the first attempt to provide the artefacts that allow an independent researcher or professional to reproduce the results and verify upstream processing steps.
+This paper follows up on previous work {[}\protect\hyperlink{ref-spinielli_2018}{5}{]}, {[}\protect\hyperlink{ref-koelle_open_2017}{8}{]}, {[}\protect\hyperlink{ref-spinielli_2017}{19}{]} and forms part of the roll-out of the performance reference trajectory by the PRU.
+As such the paper demonstrates the feasibility to the approach, the data analytical capabilities and preprocessing, and the reporting via the R/RStudio ecosystem.
+This represents an essential milestone moving towards the envisaged target environment.
+However, further work is required to establish the reproducible environment for ANS Performance.
+
+The move to open data for ANS Performance is a \emph{giant leap} in terms of transparency.
+This will foster a new level of cooperation and discussion amongst researchers and professionals interested in this topic.
+It removes the burden to acquire and preprocess large-scale surveillance data and associated aircraft data or aeronautical information.
+The departure point for this development was to address the lack and hurdle of verifying results.
+A further advantage is that novel applications, methods and algorithms (e.g.~machine learning), can be applied to a harmonised dataset.
+
+The major application of the paper results will inform the performance reference trajectory project.
+This goes in hand with the iterative move of the operational performance measures to trajectory derived data.
+This can take on the form of quality assurance measures (e.g.~verifying the order of magnitude, complementing missing reported data), but also making the reference trajectory the primary means of data collection.
+The latter will support the further development of the operational performance measures responding to the ICAO GANP.
+
+This paper documents a key milestone in the project.
+As such the future work will address the roll-out of the performance reference trajectory,
+the establishment of the current performance reporting based on the trajectory (via \url{http://ansperformance.eu}, and setup of a network with interested researchers and professionals.
+
+\hypertarget{references}{%
+\section*{References}\label{references}}
 \addcontentsline{toc}{section}{References}
 
-\vspace{6pt} \setlength{\parskip}{0pt} \footnotesize
+\vspace{6pt}
+\setlength{\parskip}{0pt}
+\footnotesize
 \setlength{\parindent}{0cm}
@@
 \hypertarget{refs}{}
-\hypertarget{ref-kolata_2011}{}
-{[}1{]} G. Kolata, ``How a New Hope in Cancer Fell Apart,'' \emph{The
-New York Times: Health}, 07-Jul-2011.
-
-\hypertarget{ref-ritchie_2012}{}
-{[}2{]} S. J. Ritchie, R. Wiseman, and C. C. French, ``Failing the
-future: Three unsuccessful attempts to replicate Bem's `Retroactive
-Facilitation of Recall'Effect,'' \emph{PloS one}, vol. 7, no. 3, p.
+\leavevmode\hypertarget{ref-kolata_2011}{}%
+{[}1{]} G. Kolata, ``How a New Hope in Cancer Fell Apart,'' \emph{The New York Times: Health}, 07-Jul-2011.
+
+\leavevmode\hypertarget{ref-ritchie_2012}{}%
+{[}2{]} S.
+J.
+Ritchie, R.
+Wiseman, and C.
+C.
+French, ``Failing the future: Three unsuccessful attempts to replicate Bem's `Retroactive Facilitation of Recall'Effect,'' \emph{PloS one}, vol.
+7, no.
+3, p.
 e33423, 2012.
 
-\hypertarget{ref-fraser_etal_2018}{}
-{[}3{]} H. Fraser, T. Parker, S. Nakagawa, A. Barnett, and F. Fidler,
-``Questionable research practices in ecology and evolution,'' \emph{PloS
-one}, vol. 13, no. 7, p. e0200303, 2018.
-
-\hypertarget{ref-fidler_wilcox_2018}{}
-{[}4{]} F. Fidler and J. Wilcox, ``Reproducibility of Scientific
-Results,'' in \emph{The Stanford Encyclopedia of Philosophy}, Winter
-2018., E. N. Zalta, Ed. Metaphysics Research Lab, Stanford University,
-2018.
-
-\hypertarget{ref-spinielli_2018}{}
-{[}5{]} E. Spinielli, R. Koelle, K. Barker, and N. Korbey, ``Open Flight
-Trajectories for Reproducible ANS Performance Review,'' presented at the
-SESAR Innovation Days 2018, 2018, p. 8.
-
-\hypertarget{ref-gandrud_2015}{}
-{[}6{]} C. Gandrud, \emph{Reproducible Research with R and R Studio,
-Second Edition (Chapman \& Hall/CRC The R Series)}. Routledge, 2015.
-
-\hypertarget{ref-peteraldhous_2016}{}
-{[}7{]} C. S. Peter Aldhous, ``Spies in the Skies,'' \emph{See Maps
-Showing Where FBI Planes Are Watching From Above}, 06-Apr-2016.
-{[}Online{]}. Available:
-\url{https://www.buzzfeed.com/peteraldhous/spies-in-the-skies}.
-
-\hypertarget{ref-koelle_open_2017}{}
-{[}8{]} R. Koelle, ``Open Source Software and Crowd Sourced Data for
-Operational Performance Analysis,'' presented at the USA/Europe ATM R\&D
-Seminar, 2017.
-
-\hypertarget{ref-daniel_nust_2018_2361304}{}
-{[}9{]} D. Nüst, M. Schutzeichel, and M. Konkol, ``Opening Reproducible
-Research: A research project website and blog,'' Nov. 2018.
-
-\hypertarget{ref-marwick_2017}{}
-{[}10{]} B. Marwick, ``Open Science in Archaeology,'' \emph{Open Science
-Framework}, Jan. 2017.
-
-\hypertarget{ref-vandewalle_2009}{}
-{[}11{]} P. Vandewalle, J. Kovacevic, and M. Vetterli, ``Reproducible
-research in signal processing,'' \emph{IEEE Signal Processing Magazine},
-vol. 26, no. 3, pp. 37--47, May 2009.
-
-\hypertarget{ref-Peng1226}{}
-{[}12{]} R. D. Peng, ``Reproducible Research in Computational Science,''
-\emph{Science}, vol. 334, no. 6060, pp. 1226--1227, 2011.
-
-\hypertarget{ref-leek_2015}{}
-{[}13{]} J. T. Leek and R. D. Peng, ``Opinion: Reproducible research can
-still be wrong: Adopting a prevention approach,'' \emph{Proc Natl Acad
-Sci USA}, vol. 112, no. 6, p. 1645, Feb. 2015.
-
-\hypertarget{ref-rcoreteam_2018}{}
-{[}14{]} R Core Team, \emph{R: A Language and Environment for
-Statistical Computing}. Vienna, Austria: R Foundation for Statistical
-Computing, 2018.
-
-\hypertarget{ref-R-knitr}{}
-{[}15{]} Y. Xie, \emph{Knitr: A General-Purpose Package for Dynamic
-Report Generation in R}. 2018.
-
-\hypertarget{ref-xie2018}{}
-{[}16{]} Y. Xie, J. Allaire, and G. Grolemund, \emph{R Markdown: The
-Definitive Guide}. Boca Raton, Florida: Chapman and Hall/CRC, 2018.
-
-\hypertarget{ref-wickham_2016}{}
-{[}17{]} H. Wickham, \emph{Ggplot2: Elegant Graphics for Data Analysis}.
-Springer-Verlag New York, 2016.
-
-\hypertarget{ref-rstudioteam_2015}{}
-{[}18{]} RStudio Team, \emph{RStudio: Integrated Development Environment
-for R}. Boston, MA: RStudio, Inc., 2015.
-
-\hypertarget{ref-performancereviewunit_2019}{}
-{[}19{]} Performance Review Unit, ``Reference trajectories for
-Performance Review of European ANS,'' 2019. {[}Online{]}. Available:
-\url{https://doi.org/10.5281/zenodo.2566844}.
-
-\hypertarget{ref-spinielli_2017}{}
-{[}20{]} E. Spinielli, R. Koelle, M. Zanin, and S. Belkoura, ``Initial
-Implementation of Reference Trajectories for Performance Review,''
-\emph{SESAR Innovation Days}, no. 7, p. 8, 2017.
-
-\setlength{\parskip}{6pt}
-
-\vspace*{1\baselineskip}
-\hypertarget{biographies}{%
-\section*{Authors' Biographies}\label{biographies}}
-\addcontentsline{toc}{section}{Authors' Biographies}
-
-\vspace*{-3\baselineskip}
-
-\begin{IEEEbiographynophoto}{\normalsize Enrico Spinielli}\normalsize
-is the Data Manager of the Performance Review Unit at EUROCONTROL, Brussels.
-He holds a Master Degree in Electronic Engineering from the University of Pisa, Italy.
-His intellectual interests include data analytics, data visualization and
-map projections.
-\end{IEEEbiographynophoto}
-
-\vspace*{-2.5\baselineskip}
-
-\begin{IEEEbiographynophoto}{\normalsize Rainer Koelle}\normalsize
-heads the Safety and Quality of Service section
-of the Performance Review Unit at EUROCONTROL, Brussels.
-He holds a PhD from the University of Lancaster, 2013, in Communication Systems.
-His research interests include data analytics for operational performance analysis.
-\end{IEEEbiographynophoto}
-
-\vfill
+\leavevmode\hypertarget{ref-fraser_etal_2018}{}%
+{[}3{]} H. Fraser, T. Parker, S. Nakagawa, A. Barnett, and F. Fidler, ``Questionable research practices in ecology and evolution,'' \emph{PloS one}, vol. 13, no. 7, p. e0200303, 2018.
+
+\leavevmode\hypertarget{ref-fidler_wilcox_2018}{}%
+{[}4{]} F.
+Fidler and J.
+Wilcox, ``Reproducibility of Scientific Results,'' in \emph{The Stanford Encyclopedia of Philosophy}, Winter 2018., E.
+N.
+Zalta, Ed.
+Metaphysics Research Lab, Stanford University, 2018.
+
+\leavevmode\hypertarget{ref-spinielli_2018}{}%
+{[}5{]} E. Spinielli, R. Koelle, K. Barker, and N. Korbey, ``Open Flight Trajectories for Reproducible ANS Performance Review,'' presented at the SESAR Innovation Days 2018, 2018, p. 8.
+
+\leavevmode\hypertarget{ref-gandrud_2015}{}%
+{[}6{]} C. Gandrud, \emph{Reproducible Research with R and R Studio, Second Edition (Chapman \& Hall/CRC The R Series)}. Routledge, 2015.
+
+\leavevmode\hypertarget{ref-peteraldhous_2016}{}%
+{[}7{]} C.
+S.
+Peter Aldhous, ``Spies in the Skies,'' \emph{See Maps Showing Where FBI Planes Are Watching From Above}, 06-Apr-2016.
+{[}Online{]}.
+Available: \url{https://www.buzzfeed.com/peteraldhous/spies-in-the-skies}.
+
+\leavevmode\hypertarget{ref-koelle_open_2017}{}%
+{[}8{]} R. Koelle, ``Open Source Software and Crowd Sourced Data for Operational Performance Analysis,'' presented at the USA/Europe ATM R\&D Seminar, 2017.
+
+\leavevmode\hypertarget{ref-marwick_2017}{}%
+{[}9{]} B. Marwick, ``Open Science in Archaeology,'' \emph{Open Science Framework}, Jan. 2017.
+
+\leavevmode\hypertarget{ref-vandewalle_2009}{}%
+{[}10{]} P. Vandewalle, J. Kovacevic, and M. Vetterli, ``Reproducible research in signal processing,'' \emph{IEEE Signal Processing Magazine}, vol. 26, no. 3, pp. 37--47, May 2009.
+
+\leavevmode\hypertarget{ref-Peng1226}{}%
+{[}11{]} R. D. Peng, ``Reproducible Research in Computational Science,'' \emph{Science}, vol. 334, no. 6060, pp. 1226--1227, 2011.
+
+\leavevmode\hypertarget{ref-leek_2015}{}%
+{[}12{]} J. T. Leek and R. D. Peng, ``Opinion: Reproducible research can still be wrong: Adopting a prevention approach,'' \emph{Proc Natl Acad Sci USA}, vol. 112, no. 6, p. 1645, Feb. 2015.
+
+\leavevmode\hypertarget{ref-rcoreteam_2018}{}%
+{[}13{]} R Core Team, \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing, 2018.
+
+\leavevmode\hypertarget{ref-R-knitr}{}%
+{[}14{]} Y. Xie, \emph{Knitr: A General-Purpose Package for Dynamic Report Generation in R}. 2018.
+
+\leavevmode\hypertarget{ref-xie2018}{}%
+{[}15{]} Y. Xie, J. J. Allaire, and G. Grolemund, \emph{R Markdown: The Definitive Guide}. Boca Raton, Florida: Chapman and Hall/CRC, 2018.
+
+\leavevmode\hypertarget{ref-wickham_2016}{}%
+{[}16{]} H. Wickham, \emph{Ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York, 2016.
+
+\leavevmode\hypertarget{ref-rstudioteam_2015}{}%
+{[}17{]} RStudio Team, \emph{RStudio: Integrated Development Environment for R}. Boston, MA: RStudio, Inc., 2015.
+
+\leavevmode\hypertarget{ref-performancereviewunit_2019}{}%
+{[}18{]} Performance Review Unit, ``Reference trajectories for Performance Review of European ANS,'' 2019. {[}Online{]}. Available: \url{https://doi.org/10.5281/zenodo.2566844}.
+
+\leavevmode\hypertarget{ref-spinielli_2017}{}%
+{[}19{]} E. Spinielli, R. Koelle, M. Zanin, and S. Belkoura, ``Initial Implementation of Reference Trajectories for Performance Review,'' \emph{SESAR Innovation Days}, no. 7, p. 8, 2017.
 

### 2019-10-04 R/extract-arrivals.R

@@ -101,8 +101,8 @@ purrr::pwalk(
 
 #################### SMOOTH #####################
 purrr::walk(
-  fs::dir_ls("data", regexp = ".*_pos_rt_.*-0.*_raw.csv"),
-  ~ smooth_arrival_trajectories(.x))
+  fs::dir_ls("data", regexp = ".*_pos_rt_.*-0._raw.csv"),
+  ~ smooth_arrival_trajectories(.x, size = 17))
 
 
 ######################## EXTRACT at 40NM #####################
@@ -125,6 +125,7 @@ poss_for_apt_on_date_within <- function(apt, date, nm = 40) {
   ) %>%
     extract_airport_distance(arp, nm)
 }
+
 
 purrr::pwalk(
   tidyr::crossing(apts, days_of_data),

### 2019-10-04 R/extract-transition-points.R

@@ -122,7 +122,7 @@ for (apt in apts) {
     #'
     #' @return A data frame with points transitionsing from the various zones
     extract_transition_points <- function(df, holdings_sf, min_ft = 6700) {
-      p <- df %>% group_by(flight_id)
+      p <- df %>% group_by(flight_id) %>% arrange(timestamp)
 
       # indexes within 40 NM
       p_40 <- p %>%

### 2019-10-04 R/plot-eidw-tma.R

@@ -62,6 +62,6 @@ plot_eidw_tma <- function() {
     data = rw,
     ggplot2::aes(x = longitude, y = latitude, xend = longitude2, yend = latitude2),
     size = 1.2)
-  g2d <- g + grw_sf + gbox + gvor + gvorl
+  g2d <- g + gbox + gvor + gvorl + grw_sf
   g2d
 }

### 2019-10-04 R/utils.R

@@ -74,8 +74,7 @@ extract_airport_distance <- function(rawdatafile, arp, distance) {
     vert_speed = col_double(),
     on_ground = col_logical(),
     distance = col_double(),
-    distance_arp = col_double(),
-    sequence_number = col_double()
+    distance_arp = col_double()
   )
   d <- distance * 1852
   rawdatafile %>%
@@ -184,27 +183,41 @@ smooth_arrival_trajectories <- function(rawdatafile, size = 5) {
     timestamp = col_datetime(format = ""),
     latitude = col_double(),
     longitude = col_double(),
-    altitude = col_integer(),
+    altitude = col_double(),
     speed_gnd = col_double(),
     track_gnd = col_double(),
     vert_speed = col_double(),
     on_ground = col_logical(),
     distance = col_double(),
-    distance_arp = col_double(),
-    sequence_number = col_integer()
+    distance_arp = col_double()
   )
 
   # RT trajectories
   vroom(rawdatafile, col_types = c) %>%
-    dplyr::add_count(flight_id) %>%
+    dplyr::group_by(flight_id) %>%
     dplyr::arrange(timestamp) %>%
+    dplyr::mutate(n = n()) %>%
+    # filter out trajectories with too few points
     dplyr::filter(n >= 3) %>%
+    # filter out trajectories with too few non NA altitudes
+    dplyr::mutate(n_not_na = sum(!is.na(altitude))) %>%
+    dplyr::filter(n_not_na >= 3) %>%
+    # fill NA with neaby ones
+    # tidyr::fill(longitude, latitude, altitude, .direction = "downup") %>%
     dplyr::mutate_at(.vars = c("longitude", "latitude", "altitude"),
-                     .funs = ~ zoo::rollmean(.x,
+                     .funs = ~ zoo::rollmedian(.x,
                                              k = size,
                                              fill = c("extend", "extend", "extend"))) %>%
+    # dplyr::mutate_at(.vars = c("longitude", "latitude", "altitude"),
+    #                  .funs = ~ zoo::rollmean(.x,
+    #                                          k = size,
+    #                                          fill = c("extend", "extend", "extend"))) %>%
+    # dplyr::mutate_at(.vars = c("longitude", "latitude"),
+    #                 .funs = ~ pracma::movavg(.x,
+    #                                         n = size,
+    #                                         type = "w")) %>%
     dplyr::ungroup() %>%
[…]

### 2019-10-04 analysis/data.Rmd
wflow_publish(c("analysis/data.Rmd", "analysis/setup.Rmd", "analysis/paper.Rmd", "analysis/website.Rmd"))

@@
     gs://pru-trajectories/products/synth_positions/cpr_fr24/mas_05_cpr_fr24_synth_positions_2017-08-03.csv.bz2 .
+$ gsutil cp \
+    gs://pru-trajectories/sources/merged/overnight_cpr_fr24/cpr_fr24_flights_2017-08-03.csv.bz2 .
 ```
@@
 As explained in the paper, you can use `R/extract-arrivals.R` to produce CSV files
-containing only the flights and positions for your airport of initerest.
+containing only the flights and positions for the airports of initerest.
 You should edit the relevant configuration values at the beginning of the script before launching it.
@@
 * Position reports for arrival at `<airport>` on `<date>`
-  are in a CSV file `<airport>_pos_rt_<date>.csv`.
-  For example `eidw_pos_rt_2017-08-01.csv` is for arrivals to
+  are in a CSV file `<airport>_pos_rt_<date>_raw.csv` (and `<airport>_pos_rt_<date>.csv`
+  for smoothed trajectories, i.e. median filter for longitude,latitude amd altitude.)
+  For example `eidw_pos_rt_2017-08-01.csv` is for (smoothed trajectories) arrivals to
   Dublin Airport (ICAO's id EIDW) on Aug, 1 2017.
